---
title: "Site use - hours (Fullerton case)"
format: html
---

::: {.callout style="border-left: 6px solid #D99B8F; --bs-callout-border: #D99B8F; padding: 1.5rem; border-radius: 8px; margin-bottom: 1rem;"}
**Why this station is treated differently**

Fullerton Entrance carries a single directional Yagi antenna (6-elements). Unlike the small omni-directional antennas used across the rest of the Motus array in the Hunter estuary (see [Motus Array Maintenance](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_4.html)), and capable of tracking site use directly, Fullerton Entrance station is located strategically: it sits on the corridor between the main area (estuary covered by several omni antennas) and the biggest and main mudflat of the area ([Fullerton Cove](https://www.google.com/maps/place/32%C2%B050'03.3%22S+151%C2%B047'40.2%22E/@-32.8437716,151.7726019,11867m/data=!3m1!1e3!4m4!3m3!8m2!3d-32.834252!4d151.794489!5m1!1e4?entry=ttu&g_ep=EgoyMDI2MDgyMy4wIKXMDSoASAFQAw%3D%3D)), which has no antenna coverage at all because of its terrain configuration. So, in order to track movement and site use over this big mudflat (Fullerton Cove, or also called Fullerton mudflat), we placed a station right at the entry ([Fullerton Entrance](https://motus.org/dashboard/#e=profile&d=stations&s=13119&f=%7B%22dtStart%22%3A%222025-08-26%22%7D)), to detect in-and-out passages.

Therefore, a tag detection at Fullerton Entrance station is either:

-   **fly-by**, the bird briefly fly by (in front of) the antenna, probably traveling between the estuary (roost?) and Fullerton mudflat (foraging?)

-   **site use** - the bird is actually sitting by the antenna itself, either roosting or foraging, using the site.

Since we cannot detect birds directly on Fullerton mudflat, we use fly-by detections at Fullerton Entrance station and their direction (another station → Fullerton Entrance or another station → Fullerton Entrance) as an **indirect proxy for mudflat use**.
:::

## Set-up data

As **we will be re-using some of the data already loaded from previous workflow (see [Station usage - hours (OBP method)](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_12.html)**), we will fetch here some of the objects we will re-use including: *visit_duration, tide_data, data_all, recv, shorebird_class, and available_bird_recv_time*.

::: blockquote-blue
*visit_duration* is then used onward so, the same rule applies in the data-set: **\< 1 hour gap between two runs are considered as a unique use of the area covered by a Motus station - the bird remained in the area within this hour gap**.
:::

```{r set-up 1, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, include = FALSE, results = 'hide'}
# Source Station usage - hours (OBP method) script into this environment
source(knitr::purl(here::here("qmd", "chapter_1", "ch1_12.qmd"),
                   output = tempfile(fileext = ".R"),
                   quiet = FALSE))
```

```{r local packages, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
library(mixtools)
library(lubridate)
library(gt)
library(lubridate)
library(ggpattern)
```

Now, we pre-fill some **key-values** we need to set-up the workflow.

```{r fc visits, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
focal_station    <- "Fullerton Entrance"   # the directional antenna

flyby_window_min <- 30                     # Period (min) before/after detection at Fullerton to look for a detection at another station

tide_lookahead_h <- 2                      # Period (hours) before/after of an "unknown" flyby detection used to infer direction from tide

# Keep only Fullerton Entrance data with visit duration (min)
fc_visits <- visit_duration %>%
  filter(recvDeployName == focal_station) %>%
  mutate(duration_min = duration_h * 60)

```


```{r fc visits check, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}

# VISIT COUNT FULLERTON
fc_visits %>%
  
  mutate(hour = lubridate::hour(visitStart)) %>%
  count(hour) %>%
  
  ggplot(aes(x = hour, y = n)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Visit Starts COUNT by Hour of Day - FULLERTON",
       x = "Hour of Day",
       y = "Number of Visits") +
  theme_minimal()


# VISIT COUNT ALL
visit_duration %>%
  
  mutate(hour = lubridate::hour(visitStart)) %>%
  count(hour) %>%
  
  ggplot(aes(x = hour, y = n)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Visit Starts COUNT by Hour of Day - ALL",
       x = "Hour of Day",
       y = "Number of Visits") +
  theme_minimal()


# DURATION ALL
visit_duration %>%
  
  mutate(hour = lubridate::hour(visitStart)) %>%
  group_by(hour) %>%
  summarise(total_duration_h = sum(duration_h, na.rm = TRUE)) %>%
  
  ggplot(aes(x = hour, y = total_duration_h)) +
  
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Sum of HOURS, by Hour of Day - ALL visits",
       x = "Hour of Day",
       y = "Total Duration (hours)") +
  theme_minimal()

# DURATION FULLERTON
fc_visits %>%
  
  mutate(hour = lubridate::hour(visitStart)) %>%
  group_by(hour) %>%
  summarise(total_duration_h = sum(duration_h, na.rm = TRUE)) %>%
  
  ggplot(aes(x = hour, y = total_duration_h)) +
  
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Sum of HOURS, by Hour of Day - FULLERTON visits",
       x = "Hour of Day",
       y = "Total Duration (hours)") +
  theme_minimal()

# DATA RAW COUNT ALL
data_all %>%
  
  mutate(hour = lubridate::hour(timeAus)) %>%
  count(hour) %>%
  
  ggplot(aes(x = hour, y = n)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Raw detection COUNT by Hour of Day - ALL",
       x = "Hour of Day",
       y = "Number of Visits") +
  theme_minimal()

# DATA RAW COUNT FULLERTON
data_all %>%
  
  filter(recvDeployName=="Fullerton Entrance") %>%
  mutate(hour = lubridate::hour(timeAus)) %>%
  count(hour) %>%
  
  ggplot(aes(x = hour, y = n)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Raw detection COUNT by Hour of Day - FULLERTON",
       x = "Hour of Day",
       y = "Number of Visits") +
  theme_minimal()


# OLD DURATION TO CHECK - FULLERTON
data_all %>%
  
  filter(recvDeployName == "Fullerton Entrance") %>%
  mutate(
    burst_inter = ifelse(tagModel == "NTQB2-6-2", dseconds(7.1), dseconds(13.1)),
    hour = lubridate::hour(timeAus)) %>%
  
  group_by(hour) %>%
  summarise(total_duration_h = sum(as.numeric(burst_inter), na.rm = TRUE) / 3600) %>%
  ggplot(aes(x = hour, y = total_duration_h)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Raw detection duration (OLD method) by Hour of Day - Fullerton",
       x = "Hour of Day",
       y = "Total Duration (hours)") +
  theme_minimal()

# OLD DURATION TO CHECK - ALL
data_all %>%
  
  mutate(
    burst_inter = ifelse(tagModel == "NTQB2-6-2", dseconds(7.1), dseconds(13.1)),
    hour = lubridate::hour(timeAus)) %>%
  
  group_by(hour) %>%
  summarise(total_duration_h = sum(as.numeric(burst_inter), na.rm = TRUE) / 3600) %>%
  ggplot(aes(x = hour, y = total_duration_h)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Raw detection duration (OLD method) by Hour of Day - ALL",
       x = "Hour of Day",
       y = "Total Duration (hours)") +
  theme_minimal()


## TEST SAMPLE COUNT ALL
set.seed(123)
data_all %>% 
  
  slice_sample(n = 1000) %>%
  
  mutate(hour = lubridate::hour(timeAus)) %>%
  count(hour) %>%
  
  ggplot(aes(x = hour, y = n)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Raw detection COUNT by Hour of Day - ALL - SAMPLE TEST (n = 1000)",
       x = "Hour of Day",
       y = "Number of Visits") +
  theme_minimal()

## TEST SAMPLE DURATION ALL
set.seed(123)
data_all %>% 
  
  slice_sample(n = 1000) %>%
  
  mutate(
    burst_inter = ifelse(tagModel == "NTQB2-6-2", dseconds(7.1), dseconds(13.1)),
    hour = lubridate::hour(timeAus)) %>%
  
  group_by(hour) %>%
  summarise(total_duration_h = sum(as.numeric(burst_inter), na.rm = TRUE) / 3600) %>%
  
  ggplot(aes(x = hour, y = total_duration_h)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Raw detection DURATION (old) by Hour of Day - ALL - SAMPLE TEST (n = 1000)",
       x = "Hour of Day",
       y = "Number of Visits") +
  theme_minimal()


## TEST SAMPLE DURATION ALL
set.seed(123)
visit_duration %>% 
  
  slice_sample(n = 1000) %>%
  
  mutate(hour = lubridate::hour(visitStart)) %>%
  group_by(hour) %>%
  summarise(total_duration_h = sum(duration_h, na.rm = TRUE)) %>%

  ggplot(aes(x = hour, y = total_duration_h)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "DURATION visitStart by Hour of Day - ALL - SAMPLE TEST (n = 1000)",
       x = "Hour of Day",
       y = "Number of Visits") +
  theme_minimal()

## TEST SAMPLE DURATION ALL
set.seed(123)
visit_duration %>% 
  
  slice_sample(n = 1000) %>%
  
  count(visitStart) %>%
  mutate(hour = lubridate::hour(visitStart)) %>%

  ggplot(aes(x = hour, y = n)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "DURATION visitStart by Hour of Day - ALL - SAMPLE TEST (n = 1000)",
       x = "Hour of Day",
       y = "Number of Visits") +
  theme_minimal()

```

## When is it a fly-by?

A very short visit (seconds to a couple of minutes) most likely reflects a bird flying by the antenna; while a longer visit likely reflects a bird actually using the site covered by antenna's range coverage. We use the same mixture-model approach as for the detection-gap threshold (see [Station usage - hours (OBP method)](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_12.html)), but here fitted to **visit durations** at Fullerton Entrance only, to find a data-driven cut-off between "flyby" and "actual use".

**Fit a 2-component mixture model** on log-transformed visit duration (component 1 = short/flyby, component 2 = long/actual use), and find the cut-point where the posterior probability of belonging to the "flyby" component drops below 0.5.

```{r fc mixture model, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
set.seed(123)

mix_fc <- normalmixEM(               # we assume normal distributions only
  log(fc_visits$duration_min + 1),   # +1 to avoid log(0), and log() used to compress/scale duration distributions and tails (i.e., lot of short vs few long)
  k = 2,                             # we assume two potential distribution (flyby and site use)
  maxit = 5000,                      # number of iteration to estimate what component each observation belongs to
  epsilon = 1e-08)                   # standard threshold value for estimation (convergence tolerance)

comp_order_fc <- order(mix_fc$mu)    # sort components -> comp_1 with smaller mean first (fly-by)

post_short    <- mix_fc$posterior[, comp_order_fc[1]] # extract first col of mix_fc$posterior, a matrix with one row per visit and one column per component, containing each visit's posterior probability of belonging to that component (rows sum to 1). So, post_short is a vector giving one value per visit of "probability this visit is a fly-by."

cut_df <- data.frame(duration_h = fc_visits$duration_h, p_short = post_short) %>%
  arrange(duration_h)

flyby_cutpoint_h <- cut_df$duration_h[which(cut_df$p_short < 0.5)[1]] # /!\ the value chosen here (0.5) finds the first duration in hour (smallest) where it has less than 50 % probability to be a fly-by /!\
```

Under `r flyby_cutpoint_h` hours (\~ `r round(flyby_cutpoint_h * 60)` min), a detection at Fullerton Entrance is considered as a *fly-by* and above as a *site use*. So, let's classify and visualise.

```{r fc classify visits, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
fc_visits <- fc_visits %>%
  mutate(visit_type = if_else(duration_h < flyby_cutpoint_h, "flyby", "actual_use"))
```

```{r fc mixture plot, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
ggplot(fc_visits, aes(x = duration_h, fill = visit_type)) +
  geom_histogram(binwidth = 0.1, color = "white", boundary = 0) +
  geom_vline(xintercept = flyby_cutpoint_h, color = "red", linetype = "dashed", linewidth = 0.7) +
  scale_fill_manual(values = c(flyby = "#E7298A", actual_use = "#66A61E")) +
  coord_cartesian(xlim = c(0, 10)) +
  labs(x = "Visit duration (hours)", y = "Count", fill = "Visit type",
       title = paste("Flyby vs actual-use classification -", focal_station),
       subtitle = paste0("Cut-point = ", round(flyby_cutpoint_h * 60, 1), " min (mixture model)")) +
  theme_minimal(base_size = 13)

ggplot(fc_visits %>%
         filter(duration_h > flyby_cutpoint_h), aes(x = duration_h, fill = visit_type)) +
  geom_histogram(binwidth = 0.1, color = "white", boundary = 0) +
  coord_cartesian(xlim = c(0, 10)) +
  labs(x = "Visit duration (hours)", y = "Count", fill = "Visit type",
       title = paste("Actual-use distribution -", focal_station),
       subtitle = paste0("Visit during more than ", round(flyby_cutpoint_h * 60, 1), " min (cut-point from mixture model)")) +
  theme_minimal(base_size = 13)

```

## Fly-by direction

For flyby events specifically, we want to know whether the bird was **heading to the mudflat** (last seen leaving the roost area, i.e. detected at another station shortly *before* the flyby) or **coming from the mudflat** (detected at another station shortly *after* the flyby). To do so, we need to understand what is the right temporal window to consider for defining the "shortly after" or "shortly before", and to rely on inference only when the detection into another Motus station is beyond this value.

Let's start by splitting the data-set accordingly before to estimate the direction of a fly-by depending on the tide: toward the mudflat of Fullerton cove if the fly-by is detected during a dropping tide, and inversely if detected during a rising tide.

```{r flyby set, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
flyby_visits <- fc_visits %>% filter(visit_type == "flyby")

all_visits_dt   <- as.data.table(visit_duration)
other_visits_dt <- all_visits_dt[recvDeployName != focal_station]
flyby_dt        <- as.data.table(flyby_visits)
```

```{r classification: infer, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
flyby_dt <- flyby_dt %>%
  
  rowwise() %>%
  
  mutate(
    closest_tide = list(
      tide_data %>%
        mutate(diff_time = abs(as.numeric(difftime(tideDateTimeAus, visitStart, units = "mins")))) %>%
        slice_min(diff_time, n = 1, with_ties = FALSE)),  # single closest tide, either side of visitStart
    
    nearestTideDiel   = closest_tide$day_night[1],
    nearestTideLevel  = closest_tide$high_low[1],
    nearestTideTime   = closest_tide$tideDateTimeAus[1]) %>%
  
  ungroup() %>%
  select(-closest_tide) %>%
  
  mutate(
    direction = case_when(
      nearestTideLevel == "High" ~ "From mudflat (rising tide)",
      nearestTideLevel == "Low"  ~ "To mudflat (dropping tide)",
      TRUE ~ NA_character_ ))


barplot(table(flyby_dt$direction, useNA = "ifany"),
        main = "Inferred flyby direction based on next tide",
        ylab = "Count of flybys")

```

For each fly-by detected at Fullerton Entrance, we then look for a **confirming detection at another Motus station**: if the tide was rising, we check for the *closest following* detection after Fullerton Entrance; if the tide was dropping, we check for the *closest preceding* detection before Fullerton Entrance.

```{r classification: verified, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
flyby_dt <- flyby_dt %>%
  
  rowwise() %>%
  
  mutate(
    .band    = as.character(Band.ID),
    .start   = visitStart,
    .station = recvDeployName,
    
    other_visit = list(
      if (direction == "From mudflat (rising tide)") {
        all_visits_dt %>%
          filter(as.character(Band.ID) == .band,
                 recvDeployName != .station,
                 visitStart >= .start) %>%
          slice_min(visitStart, n = 1, with_ties = FALSE)
      } else if (direction == "To mudflat (dropping tide)") {
        all_visits_dt %>%
          filter(as.character(Band.ID) == .band,
                 recvDeployName != .station,
                 visitStart <= .start) %>%
          slice_max(visitStart, n = 1, with_ties = FALSE)
      } else {
        all_visits_dt %>% filter(FALSE)
      }
    ),
    
    side_detection = other_visit$recvDeployName[1],
    side_time      = other_visit$visitStart[1],
    delay_min      = as.numeric(abs(difftime(.start, other_visit$visitStart[1], units = "mins")))
  ) %>%
  ungroup() %>%
  select(-other_visit, -.band, -.start, -.station)

```

We then look at the **distribution of the delay between a Fly-By detection at Fullerton entrance and another Motus station**, in order to identify a temporal window we can use as a cut-off for concluding that *"the bird flew continuously between the fly-by detection at Fullerton Entrance and the next detection at another Motus station"* (or, in the reverse case, between a detection at another station and the fly-by at Fullerton entrance during a dropping tide).

We first filter out any extreme tail in the data, then visualise the distribution to estimate how many underlying components it contains.

```{r delay tail, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
flyby_dt %>%
  arrange(desc(delay_min)) %>%
  select(Band.ID, visitStart, direction, side_detection, delay_min) %>%
  head(20) %>%
  DT::datatable()

hist(flyby_dt$delay_min,
     breaks = 50,
     main  = "Delay to (or period before) nearest next site detection",
     xlab  = "Delay (minutes)")

rect(xleft = 0, xright = 20000,
     ybottom = 0, ytop = par("usr")[4],   # par("usr")[4] = top of plot region
     border = "red", lwd = 2)

flyby_dt %>%
  filter(delay_min < 4000) %>%
  pull(delay_min) %>%
  hist(breaks = 50,
       main  = "Delay to (or period before) nearest next site detection",
       xlab  = "Delay (minutes)") 
```

So we **visually** decided to filter out delays of more than 4000 min (\~ 2.8 days).

```{r filter and plot, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
flyby_dt_trimmed  <- flyby_dt %>% 
  filter(delay_min < 4000)
```

```{r filter and plot 2, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}

flyby_dt_trimmed %>%
  pull(delay_min) %>%
  hist(breaks = 100,
       main  = "Delay to (or period before) nearest next site detection",
       xlab  = "Delay (minutes)") 

flyby_dt_trimmed %>%
  filter(delay_min < 100) %>%
  pull(delay_min) %>%
  hist(breaks = 100,
       main  = "Delay to (or period before) nearest next site detection",
       xlab  = "Delay (minutes)") 

```

**Visually,** we can estimate that the overall temporal window for *"continuous flights"* between the Fullerton entrance and another Motus station is about **10 minutes**. However, we can fit an N-mixture model to confirm this.

```{r extract cutpoint delay, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}

delay_vals <- na.omit(flyby_dt_trimmed$delay_min)
set.seed(1)  # EM is stochastic in its starting values
mix <- normalmixEM(delay_vals, 
                   k = 2)
plot(mix, which = 2)

# order components so mu1 < mu2 (short-delay vs long-delay cluster)
ord    <- order(mix$mu)
mu     <- mix$mu[ord]
sigma  <- mix$sigma[ord]
lambda <- mix$lambda[ord]

# find where the two weighted densities cross, between the two means
f <- function(x) lambda[1]*dnorm(x, mu[1], sigma[1]) - lambda[2]*dnorm(x, mu[2], sigma[2])
flyby_window_min <- uniroot(f, lower = mu[1], upper = mu[2])$root
flyby_window_min

```

With a statistically determined **threshold**, any delay between a detection at the Fullerton entrance and a another Motus station that is shorter than `r flyby_window_min` minutes will be treated as a *verified* flight direction for the bird. Beyond this value, a continuous flight inferred from a double-detection data based might not be considered realistic, suggesting detection at another Motus station cannot be trusted to infer the bird’s flight direction.

**See below for the distribution of these results.**

```{r fc flyby set, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
flyby_dt <- flyby_dt %>%
  mutate(retained_cut = case_when(
    # is.na(delay)                    ~ "inferred_only",
    delay_min <= flyby_window_min   ~ "Verified by another Motus station detection",
    TRUE                            ~ "Inferred only from tide data"))
```

```{r plot directions, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
bp_data <- t(table(flyby_dt$direction, flyby_dt$retained_cut))

bp <- barplot(bp_data,
              beside = TRUE,
              legend.text = FALSE,
              main = paste0("Direction by confirmation \nwith another detection before ",
                             round(flyby_window_min, 1), " min (cutpoint)"),
              ylab = "Count of flybys")

legend("bottom",
       inset = c(0, -0.37),
       legend = rownames(bp_data),
       fill = gray.colors(nrow(bp_data)),
       xpd = TRUE,
       bty = "n")

```

```{r plot directions species/Station, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
# Filter to only verified detections
flyby_verified <- flyby_dt %>%
  filter(retained_cut == "Verified by another Motus station detection") %>%
  mutate(species_type = shorebird_class[speciesEN])

# Get unique species, paired with their species_type
species_df <- flyby_verified %>%
  distinct(speciesEN, species_type)

# Generate one plot per species, each faceted by recvDeployName, filled by direction
plot_list <- purrr::pmap(species_df, function(speciesEN, species_type) {
  
  flyby_verified %>%
    filter(speciesEN == !!speciesEN) %>%
    ggplot(aes(x = delay_min, fill = direction)) +
    geom_histogram(bins = 30, position = "identity", alpha = 0.6, color = "white") +
    facet_wrap(~ side_detection) +
    scale_fill_manual(values = c("From mudflat (rising tide)" = "#E69F00",
                                  "To mudflat (dropping tide)" = "#2C3E50")) +
    labs(title = speciesEN,
         subtitle = species_type,
         x = "Delay (minutes)",
         y = "Count",
         fill = "Direction") +
    theme_minimal(base_size = 12) +
    theme(strip.text = element_text(face = "bold"))
})

names(plot_list) <- species_df$speciesEN

# To view all plots one after another:
plot_list
```

::: panel-tabset
##### Resident

```{r used-rate-resident2, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis'}
resident_species <- species_df$speciesEN[species_df$species_type == "resident"]

for (sp in resident_species) {
  cat("\n\n###### ", sp, "\n\n")
  print(plot_list[[sp]])
  cat("\n\n")
}
```

##### Migratory

```{r used-rate-migratory2, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis'}
migratory_species <- species_df$speciesEN[species_df$species_type == "migratory"]

for (sp in migratory_species) {
  cat("\n\n###### ", sp, "\n\n")
  print(plot_list[[sp]])
  cat("\n\n")
}
```
:::

Looking through the time of the detection for each *fly-bys*, we have been curious to look at the temporal distribution of the fly-by depending the directions to see whether it was tide dependent only or also relying on the time of the day. See below.

```{r hours distribution, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results = 'hide'}
flyby_dt %>%
  
  mutate(hour_bin = lubridate::hour(visitStart)) %>%
  count(hour_bin, direction, retained_cut) %>%
  
  ggplot(aes(x = hour_bin, y = n, fill = direction, pattern = retained_cut)) +
  
  geom_col_pattern(
    width = 0.9, color = "black", linewidth = 0.3,
    pattern_fill = "white", pattern_colour = "black",
    pattern_density = 0.05, pattern_spacing = 0.02, pattern_angle = 45) +
  
  facet_wrap(~ direction, nrow = 2) +
  
  scale_fill_manual(
    values = c("From mudflat (rising tide)" = "#E69F00",
               "To mudflat (dropping tide)" = "#2C3E50"),
    name = "Direction of the flyby",
    guide = guide_legend(override.aes = list(pattern = "none"), order = 1)) +
  
  scale_pattern_manual(
    values = c("Verified by another Motus station detection" = "stripe",
               "Inferred only from tide data" = "none"),
    name = "Classification",
    guide = guide_legend(override.aes = list(fill = "white", pattern_fill = "white"), order = 2)) +
  
  scale_x_continuous(breaks = seq(0, 24, 2), expand = expansion(mult = c(0.01, 0.01))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  
  labs(
    x = "Hour of day",
    y = "Count of flybys",
    title = "Flyby detections at Fullerton Cove",
    subtitle = "Direction inferred from rising/dropping tide, sometimes verified by a\npreceding or following detection at another Motus station") +
  
  theme_minimal(base_size = 15) +
  
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(color = "grey30", size = 11, margin = margin(b = 12)),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(color = "grey20"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_text(face = "bold", size = 11),
    legend.text = element_text(size = 10),
    legend.margin = margin(t = 8),
    plot.margin = margin(10, 15, 10, 10))
```

check confounding effect with **average** tide between Oct-March:

::: {.callout-tip collapse="true" title="See tide table 2"}
```{r 1 Tide table, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
tide_data %>%
  DT::datatable()
```
:::

```{r 1 Tide table curve 2, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}

# Filter to Oct-Mar, extract month + hour of day, average tide height per hour per month
tide_monthly <- tide_data %>%
  filter(monthNum %in% c(10, 11, 12, 1, 2, 3)) %>%
  mutate(
    month = lubridate::month(tideDateTimeAus, label = TRUE, abbr = TRUE),
    hour  = lubridate::hour(tideDateTimeAus) + lubridate::minute(tideDateTimeAus) / 60
  ) %>%
  group_by(month, hour_bin = round(hour * 2) / 2) %>%   # round to nearest half-hour to smooth out noise
  summarise(mean_height = mean(tideHeight, na.rm = TRUE), .groups = "drop")

# Keep month order Oct -> Mar rather than alphabetical/calendar-year order
tide_monthly <- tide_monthly %>%
  mutate(month = factor(month, levels = c("Oct", "Nov", "Dec", "Jan", "Feb", "Mar")))

ggplot(tide_monthly, aes(x = hour_bin, y = mean_height, color = month)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = c(5, 20), color = "black", linewidth = 0.7) +

  scale_x_continuous(breaks = 0:23) +
  scale_color_viridis_d(option = "D", end = 0.9) +
  labs(x = "Hour of Day", y = "Mean Tide Height (m)",
       title = "Tide Height by Hour of Day",
       subtitle = "October - March, one curve per month (averaged across years)",
       color = "Month") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())


# Filter to Oct-Mar, extract hour of day, average tide height per hour
tide_hourly <- tide_data %>%
  
  filter(monthNum %in% c(10, 11, 12, 1, 2, 3)) %>%
  
  mutate(hour = lubridate::hour(tideDateTimeAus)) %>% 
  group_by(hour_bin = floor(hour)) %>%                                     
  summarise(
    mean_height = mean(tideHeight, na.rm = TRUE),
    se_height   = sd(tideHeight, na.rm = TRUE) / sqrt(n()),
    n           = n(),
    .groups = "drop")

ggplot(tide_hourly, aes(x = hour_bin, y = mean_height)) +
  
  geom_vline(xintercept = c(5, 20), color = "black", linewidth = 0.7) +
  
  geom_col(fill = "grey70", width = 0.8, alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.4, color = "#2C3E50", 
              linewidth = 1.2, se = TRUE, fill = "#2C3E50", alpha = 0.15) +
  geom_errorbar(aes(ymin = mean_height - se_height, ymax = mean_height + se_height),
                width = 0.2, color = "grey30") +
  scale_x_continuous(breaks = 0:23) +
  labs(x = "Hour of Day", y = "Mean Tide Height (m)",
       title = "Average Tide Height by Hour of Day",
       subtitle = "October - March, averaged across all years | error bars = SE") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())

```

```{r hours distribution per species, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
# Add species_type and build the species lookup table
flyby_dt <- flyby_dt %>%
  mutate(species_type = shorebird_class[speciesEN])

species_df <- flyby_dt %>%
  distinct(speciesEN, species_type)

# Generate one plot per species
plot_list_hourly <- purrr::pmap(species_df, function(speciesEN, species_type) {
  
  flyby_dt %>%
    filter(speciesEN == !!speciesEN) %>%
    mutate(hour_bin = lubridate::hour(visitStart)) %>%
    count(hour_bin, direction, retained_cut) %>%
    
    ggplot(aes(x = hour_bin, y = n, fill = direction, pattern = retained_cut)) +
    
    geom_col_pattern(
      width = 0.9, color = "black", linewidth = 0.3,
      pattern_fill = "white", pattern_colour = "black",
      pattern_density = 0.05, pattern_spacing = 0.02, pattern_angle = 45) +
    
    facet_wrap(~ direction, nrow = 2) +
    
    geom_vline(xintercept = c(5, 20), color = "black", linewidth = 0.7) +
    
    scale_fill_manual(
      values = c("From mudflat (rising tide)" = "#E69F00",
                 "To mudflat (dropping tide)" = "#2C3E50"),
      name = "Direction of the flyby",
      guide = guide_legend(override.aes = list(pattern = "none"), order = 1)) +
    
    scale_pattern_manual(
      values = c("Verified by another Motus station detection" = "stripe",
                 "Inferred only from tide data" = "none"),
      name = "Classification",
      guide = guide_legend(override.aes = list(fill = "white", pattern_fill = "white"), order = 2)) +
    
    scale_x_continuous(breaks = seq(0, 24, 2), expand = expansion(mult = c(0.01, 0.01))) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    
    labs(
      x = "Hour of day",
      y = "Count of flybys",
      title = speciesEN,
      subtitle = paste0(species_type, 
                         " | Direction inferred from rising/dropping tide, sometimes verified\nby a preceding or following detection at another Motus station")) +
    
    theme_minimal(base_size = 15) +
    
    theme(
      plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
      plot.subtitle = element_text(color = "grey30", size = 11, margin = margin(b = 12)),
      axis.title = element_text(face = "bold", size = 12),
      axis.text = element_text(color = "grey20"),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      strip.text = element_text(face = "bold", size = 12),
      legend.position = "bottom",
      legend.box = "vertical",
      legend.title = element_text(face = "bold", size = 11),
      legend.text = element_text(size = 10),
      legend.margin = margin(t = 8),
      plot.margin = margin(10, 15, 10, 10))
})

names(plot_list_hourly) <- species_df$speciesEN

```

::: panel-tabset
##### Resident

```{r used-rate-resident, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis'}
resident_species <- species_df$speciesEN[species_df$species_type == "resident"]

for (sp in resident_species) {
  cat("\n\n###### ", sp, "\n\n")
  print(plot_list_hourly[[sp]])
  cat("\n\n")
}
```

##### Migratory

```{r used-rate-migratory, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis'}
migratory_species <- species_df$speciesEN[species_df$species_type == "migratory"]

for (sp in migratory_species) {
  cat("\n\n###### ", sp, "\n\n")
  print(plot_list_hourly[[sp]])
  cat("\n\n")
}
```
:::

::: {.callout-tip collapse="true" title="See for summary data table"}
```{r hours distribution per species summary data table, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}

library(gt)

# ---- Helper: decimal hour (e.g. 13.75) -> "HH:MM" string (e.g. "13:45") ----
# Works for both whole hours (peak hours) and fractional hours (mean/median).
decimal_hour_to_hhmm <- function(x) {
  if (is.na(x)) return(NA_character_)
  total_min <- round(x * 60)                 # convert to minutes, rounded to nearest minute
  total_min <- total_min %% (24 * 60)        # wrap around if rounding pushes past 24:00 (e.g. 23.99h -> 00:00)
  h <- total_min %/% 60
  m <- total_min %% 60
  sprintf("%02d:%02d", h, m)
}

make_summary_table <- function() {

  summary_table <- flyby_dt %>%
    filter(!is.na(direction)) %>%             # direction NA = tide rule couldn't be evaluated, exclude from per-direction breakdown
    mutate(
      hour_of_day = lubridate::hour(visitStart) + lubridate::minute(visitStart) / 60   # decimal hour, e.g. 14:30 -> 14.5, so mean/median aren't rounded to whole hours before formatting
    ) %>%

    group_by(speciesEN, direction) %>%
    summarise(
      n_individual  = n_distinct(Band.ID),                                   # how many distinct birds contribute to this species x direction cell
      n_events      = n(),                                                   # total flyby events in this cell

      n_verified    = sum(retained_cut == "Verified by another Motus station detection", na.rm = TRUE),
      n_inferred    = sum(retained_cut == "Inferred only from tide data", na.rm = TRUE),
      pct_verified  = round(100 * n_verified / n_events, 1),                 # share of events with independent confirmation

      median_hour   = decimal_hour_to_hhmm(median(hour_of_day)),
      mean_hour     = decimal_hour_to_hhmm(mean(hour_of_day)),

      # --- peak hour BEFORE midday (00:00-11:59) ---
      # bin to whole hours for the "which hour has the most events" count,
      # then format that whole hour as HH:MM (always :00 since it's a bin label)
      peak_hour_AM  = {
        am <- floor(hour_of_day[hour_of_day < 12])
        if (length(am) == 0) NA_character_
        else {
          tab <- sort(table(am), decreasing = TRUE)
          paste0(decimal_hour_to_hhmm(as.numeric(names(tab)[1])), " (n=", tab[1], ")")
        }
      },

      # --- peak hour AFTER midday (12:00-23:59) ---
      peak_hour_PM  = {
        pm <- floor(hour_of_day[hour_of_day >= 12])
        if (length(pm) == 0) NA_character_
        else {
          tab <- sort(table(pm), decreasing = TRUE)
          paste0(decimal_hour_to_hhmm(as.numeric(names(tab)[1])), " (n=", tab[1], ")")
        }
      },

      .groups = "drop"
    ) %>%
    arrange(speciesEN, direction) %>%
    rename(Species = speciesEN, Direction = direction)

  return(summary_table)
}

# Build the table
flyby_summary <- make_summary_table()

# ---- Render with gt(), same visual style as your existing tables ----------
flyby_summary %>%
  gt() %>%
  tab_header(title = "Fullerton Entrance - Flyby direction, verification & timing by species") %>%
  opt_align_table_header(align = "left") %>%
  cols_label(
    n_individual = "n birds",
    n_events     = "n events",
    n_verified   = "n verified",
    n_inferred   = "n inferred",
    pct_verified = "% verified",
    median_hour  = "Median time",
    mean_hour    = "Mean time",
    peak_hour_AM = "Peak hour (AM)",
    peak_hour_PM = "Peak hour (PM)"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(Species))) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "grey40",
      weight = px(5)),
    locations = cells_body(
      rows = which(!duplicated(flyby_summary$Species))[-1])) %>%
  tab_footnote(
    footnote = "Times shown as HH:MM (24h clock). AM = 00:00-11:59, PM = 12:00-23:59. Peak hour = the single clock hour with the most flyby events in that half of the day.",
    locations = cells_column_labels(columns = c(median_hour, mean_hour, peak_hour_AM, peak_hour_PM))
  ) %>%
  tab_footnote(
    footnote = "Verified = confirmed by a detection at another Motus station within the flyby_window_min cut-point; Inferred = direction assumed from tide state only.",
    locations = cells_column_labels(columns = pct_verified)
  )

```
:::

::: {.callout-tip collapse="true" title="See for raw data table"}
```{r hours distribution per species raw data table, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}

flyby_dt %>%
  DT::datatable()

```
:::

## Fly-by & Site-use

We now bring `visit_type` (fly-by / actual site use) and the direction back together into a single table covering **all** Fullerton entrance visits, so that "Site use" sits alongside the flyby directions as one classification.

```{r fc classified, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
fc_classified <- fc_visits %>%
  left_join(
    flyby_dt %>%
      select(Band.ID, visitID, direction, side_detection, side_time, delay_min,
             retained_cut, nearestTideDiel, nearestTideLevel),  
    by = c("Band.ID", "visitID")) 
```

**Summary of every classified event at Fullerton Entrance** - Actual site use plus each fly-by direction categories, with the number, proportion of events, number of individual birds, mean and median duration for each.Note the proportion of verified fly-bys must be compared with the inferred fly-bys only (and not against site use). However, inferred fly-bys might be compared with site use.

```{r fc summary table, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
fc_classified %>%
  mutate(
    direction    = replace_na(direction, "-"),
    retained_cut = replace_na(retained_cut, "-")) %>%
  group_by(visit_type, direction, retained_cut) %>%
  summarise(
    n_events             = n(),
    n_birds              = n_distinct(Band.ID),
    median_duration_min  = round(median(duration_h*60), 0),
    mean_duration_min    = round(mean(duration_h*60), 0),
    .groups = "drop") %>%
  mutate(perc_event      = round(100 * n_events / sum(n_events), 1)) %>%
  arrange(visit_type, direction, retained_cut) %>%
  gt() %>%
  tab_header(title = paste("Summary -", focal_station)) %>%
  opt_align_table_header(align = "left") %>%
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels())
```

```{r fc events by species plot, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
ggplot(fc_classified
       %>% mutate(direction = replace_na(direction, "Actual site use")), ## Be careful and know your data here
       aes(x = direction)) +
  geom_bar() +
  facet_wrap(~ speciesEN) +
  labs(x = "Direction", y = "Number of events",
       title = paste("Direction of detections at", focal_station)) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text = element_text(face = "bold"))
```

## Rate of use

For actual site use events, the duration a bird spent at Fullerton Entrance is simply duration_h of the visit itself.

For fly-bys there is no direct duration: instead, we estimate how long the bird likely remained on the mudflat using either a corroborating detection at another Motus station, or, when none is available, the surrounding tide turning point as a proxy boundary:

To mudflat (dropping tide) — if the bird was subsequently detected at another station near the next tide event, the duration runs from the moment it was detected here (visitEnd) to that corroborating detection (double_detection). Otherwise, the duration runs from visitEnd to the next tide turning point.

From mudflat (rising tide) — if the bird was previously detected as a flyby near the last tide event, the duration runs from that corroborating detection to the moment it was detected here (visitStart) (double_detection). Otherwise, the duration runs from the previous tide turning point to visitStart.

**Let's first estimate/compute the time spent into Fullerton mudflat for the fly-bys.**

```{r fc tide helpers, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
# Turning points: midpoint between each consecutive pair of tide extremes
tide_times     <- sort(unique(tide_data$tideDateTimeAus))
turning_points <- tide_times[-length(tide_times)] + diff(tide_times) / 2

next_turning_point <- function(t) {
  cand <- turning_points[turning_points > t]
  if (length(cand) == 0) return(as.POSIXct(NA))
  min(cand)
}
prev_turning_point <- function(t) {
  cand <- turning_points[turning_points < t]
  if (length(cand) == 0) return(as.POSIXct(NA))
  max(cand)
}

flyby_dt <- flyby_dt %>%
  rowwise() %>%
  mutate(
    .band      = as.character(Band.ID),
    .start     = visitStart,
    .station   = recvDeployName,
    .tide_time = nearestTideTime,
    
    # Find the closest detection at another station, searched in the
    # correct direction relative to nearestTideTime
    match_visit = list(
      if (direction == "To mudflat (dropping tide)") {
        # nearestTideTime is AFTER visitStart -> search forward
        all_visits_dt %>%
          filter(as.character(Band.ID) == .band,
                 recvDeployName != .station,
                 visitStart > .start) %>%
          mutate(diff_to_tide = abs(as.numeric(difftime(visitStart, .tide_time, units = "mins")))) %>%
          slice_min(diff_to_tide, n = 1, with_ties = FALSE)
      } else if (direction == "From mudflat (rising tide)") {
        # nearestTideTime is BEFORE visitStart -> search backward
        all_visits_dt %>%
          filter(as.character(Band.ID) == .band,
                 recvDeployName != .station,
                 visitStart < .start) %>%
          mutate(diff_to_tide = abs(as.numeric(difftime(visitStart, .tide_time, units = "mins")))) %>%
          slice_min(diff_to_tide, n = 1, with_ties = FALSE)
      } else {
        all_visits_dt %>% filter(FALSE)
      } ),
    
    side_detection = match_visit$recvDeployName[1],
    side_time      = match_visit$visitStart[1],
    
    time_used_type = case_when(
      visit_type != "flyby" ~ NA_character_,
      !is.na(side_time)     ~ "double_detection",
      TRUE                  ~ "single_detection"),
    
    turning_point = case_when(
      direction == "To mudflat (dropping tide)" ~ next_turning_point(.start),
      direction == "From mudflat (rising tide)" ~ prev_turning_point(.start),
      TRUE ~ as.POSIXct(NA) ),
    
    time_used = case_when(
      visit_type != "flyby"                ~ as.numeric(NA),
      time_used_type == "double_detection" ~ as.numeric(abs(difftime(side_time, .start, units = "hours"))),
      time_used_type == "single_detection" ~ as.numeric(abs(difftime(turning_point, .start, units = "hours"))),
      TRUE ~ as.numeric(NA)) ) %>%
  ungroup() %>%
  select(-match_visit, -.band, -.start, -.station, -.tide_time)

```

**Let's now add the duration from actual site use of at the Motus station area to the inferred time spent into the mudflat from fly-bys.**

```{r fc plot use, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
# Dominant tideCategory across a time window (hourly expansion, nearest-tide join)
tide_dt <- as.data.table(tide_data)[order(tideDateTimeAus), .(tideDateTimeAus, tideCategory)]

dominant_tide <- function(w_start, w_end) {
  if (is.na(w_start) || is.na(w_end) || w_end <= w_start) return(NA_character_)
  hrs    <- seq(floor_date(w_start, "hour"), floor_date(w_end, "hour"), by = "hour")
  hrs_dt <- data.table(hour_dt = hrs)
  matched <- tide_dt[hrs_dt, on = .(tideDateTimeAus = hour_dt), roll = "nearest"]
  names(sort(table(matched$tideCategory), decreasing = TRUE))[1]}

# get actual site-use rows from fc_classified
fc_classified_clean <- fc_classified %>%
  filter(visit_type == "actual_use") %>%
  mutate(time_used = duration_h) ## in  hour!!!

# combine with flyby_dt
plot_data <- bind_rows(flyby_dt,  fc_classified_clean)

plot_data <- plot_data %>%    
  mutate(direction = replace_na(direction, "Actual site use")) %>%
  
  filter(direction %in% c("Actual site use", 
                           "To mudflat (dropping tide)",
                           "From mudflat (rising tide)")) %>%
  rowwise() %>%
  mutate(
    category = if_else(direction == "Actual site use",
                        "Fullerton entrance (Motus station)", "Fullerton cove (mudflat)"),
    
    win_start = case_when(
      direction == "Actual site use" ~ visitStart,
      direction == "To mudflat (dropping tide)" ~ visitEnd,
      direction == "From mudflat (rising tide)" & time_used_type == "double_detection" ~ side_time,
      direction == "From mudflat (rising tide)" ~ turning_point),
    win_end = case_when(
      direction == "Actual site use" ~ visitEnd,
      direction == "To mudflat (dropping tide)" & time_used_type == "double_detection" ~ side_time,
      direction == "To mudflat (dropping tide)" ~ turning_point,
      direction == "From mudflat (rising tide)" ~ visitStart),
    
    tideCategory = dominant_tide(win_start, win_end),
    tideDiel     = if_else(grepl("Diurnal", tideCategory), "Diurnal", "Nocturnal")) %>%
    ungroup() %>%
  filter(!is.na(time_used), time_used > 0)
```

Because "Fullerton mudflat" is an *inferred* location rather than an actual listening Motus station, both categories are expressed as a rate against the same denominator: the hours Fullerton Entrance Motus station was available (station ON + bird tagged and alive), per bird and per diel period. This comes from `available_bird_recv_time`, computed in the previous chapter.

```{r fc rate use, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
# Available hours at Fullerton Entrance, per bird / species / diel period
available_fc <- available_bird_recv_time %>%
  filter(recvDeployName == "Fullerton Entrance") %>%
  group_by(Band.ID, speciesEN, tideDiel) %>%
  summarise(available_h = sum(duration_h), .groups = "drop")

# Used hours per bird / category / diel period, then compute rate of use (%)
plot_summary <- plot_data %>%
  group_by(Band.ID, speciesEN, category, tideDiel) %>%
  summarise(used_h = sum(time_used), .groups = "drop") %>%
  left_join(available_fc, by = c("Band.ID", "speciesEN", "tideDiel")) %>%
  mutate(rate_use = (used_h / available_h) * 100,
         rate_use = pmin(rate_use, 100))  # cap: a few cases exceed 100% (see note below)
```

::: blockquote-yellow
**Note:** a small number of birds show `rate_use` above 100%, since mudflat duration is *inferred* from tide turning points rather than directly observed, and can occasionally exceed the bird's directly-available Fullerton Entrance time. These values are capped at 100% rather than excluded.
:::

```{r fc rate use plot, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
ggplot(plot_summary, aes(x = category, y = rate_use, fill = tideDiel)) +
  geom_boxplot(outlier.shape = NA,
               position = position_dodge(width = 0.8, preserve = "single")) +
  geom_point(aes(shape = tideDiel),
             position = position_jitterdodge(jitter.width = 0.6,
                                              jitter.height = 0,
                                              dodge.width = 0.8),
             alpha = 1, size = 1.5,
             show.legend = FALSE) +
  scale_shape_manual(values = c("Diurnal" = 21, "Nocturnal" = 16)) +
  facet_wrap(~ speciesEN) +
  labs(x = "Location", y = "Rate of Use (%)", fill = "Tide period",
       title = "Rate of use at Fullerton Entrance vs Mudflat") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal() +
  scale_fill_manual(values = c("Diurnal" = "white", "Nocturnal" = "darkgrey")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

::: {.callout-tip collapse="true" title="See Rate of Use statistics into this table"}
```{r 4 flyby rate use table, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
# General function to provide one table with stats values, optionally filtered by species_type
make_summary_table <- function(type = NULL) {
  data <- plot_summary %>%
    mutate(species_type = shorebird_class[speciesEN])
  
  if (!is.null(type)) {
    data <- data %>% filter(species_type == type)
  }
  
  summary_table <- data %>%
    group_by(speciesEN, category, tideDiel) %>%
    summarise(
      n_individual = n_distinct(Band.ID),
      min    = round(min(rate_use, na.rm = TRUE), 1),
      q1     = round(quantile(rate_use, 0.25, na.rm = TRUE), 1),
      median = round(median(rate_use, na.rm = TRUE), 1),
      q3     = round(quantile(rate_use, 0.75, na.rm = TRUE), 1),
      max    = round(max(rate_use, na.rm = TRUE), 1),
      mean   = round(mean(rate_use, na.rm = TRUE), 1),
      sd     = round(sd(rate_use, na.rm = TRUE), 1),
      .groups = 'drop'
    ) %>%
    arrange(speciesEN, category, tideDiel) %>%
    rename(Species = "speciesEN", Location = "category", Condition = "tideDiel")
  
  return(summary_table)
}

# Helper to render a summary table as a styled gt table
render_summary_table <- function(summary_table, subtitle = NULL) {
  summary_table %>%
    gt() %>%
    tab_header(title = "Statistics - Rate of Use", subtitle = subtitle) %>%
    opt_align_table_header(align = "left") %>%
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_column_labels()) %>%
    tab_style(
      style = cell_text(style = "italic"),
      locations = cells_body(columns = c(Species))) %>%
    tab_style(
      style = cell_borders(
        sides = "top",
        color = "grey40",
        weight = px(5)),
      locations = cells_body(
        rows = which(!duplicated(summary_table$Species))[-1]))
}

# One table per species_type
rate_use_summary_resident  <- make_summary_table("resident")
rate_use_summary_migratory <- make_summary_table("migratory")

# Render each
render_summary_table(rate_use_summary_resident, subtitle = "Resident species")
render_summary_table(rate_use_summary_migratory, subtitle = "Migratory species")
```
:::

::: {.callout-tip collapse="true" title="See raw data into this table"}
```{r table of plot summary, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}

plot_summary %>%
  DT::datatable()

```
:::

## *Beta \|* Statistical analysis

::: {.callout-tip collapse="true" title="See exploratory analysis"}
Instead of just assuming a tide-based guess about which direction birds were flying, we statistically provide a score for how often it's actually right.

```{r GLM direction, message = FALSE,  warning = FALSE, echo = TRUE, eval = TRUE, results = 'hide'}
# ============================================================================
# VALIDATING THE TIDE-BASED FLYBY DIRECTION RULE
# ============================================================================

# --- WHY THE CURRENT CHECK IS CIRCULAR ---------------------------------------
# In the original code, `other_visit` is only searched:
#   - forward in time IF the tide rule said "From mudflat (rising tide)"
#   - backward in time IF the tide rule said "To mudflat (dropping tide)"
# So "verified" only means "a detection existed on the side the tide rule
# told us to look at" - it never checks the OTHER side, so it can't catch
# cases where the tide rule is wrong. To really validate the rule, we need
# an independent estimate of direction that does NOT use the tide state,
# then compare the two.

other_visits_dt <- as.data.table(visit_duration)[recvDeployName != focal_station]

# --- 1. Independently find the closest OTHER-station detection on EACH side --
# For every flyby at Fullerton, we look both backward and forward in time,
# regardless of what the tide predicted.
# NOTE: pulling from flyby_dt (not flyby_visits) here, because `direction`
# only exists on flyby_dt - it's added by the "classification: infer" chunk
# and was never present on flyby_visits.
flyby_check <- flyby_dt %>%
  dplyr::select(Band.ID, speciesEN, recvDeployName, visitID, visitStart, visitEnd, direction) %>%
  rowwise() %>%
  mutate(
    .band  = Band.ID,
    .start = visitStart,

    # Closest detection at ANOTHER station BEFORE this flyby (any direction)
    prev_visit = list(
      other_visits_dt %>%
        filter(Band.ID == .band, visitStart <= .start) %>%
        slice_max(visitStart, n = 1)
    ),
    prev_delay = as.numeric(difftime(.start, prev_visit$visitStart[1], units = "mins")),

    # Closest detection at ANOTHER station AFTER this flyby (any direction)
    next_visit = list(
      other_visits_dt %>%
        filter(Band.ID == .band, visitStart >= .start) %>%
        slice_min(visitStart, n = 1)
    ),
    next_delay = as.numeric(difftime(next_visit$visitStart[1], .start, units = "mins"))
  ) %>%
  ungroup() %>%
  dplyr::select(-prev_visit, -next_visit, -.band, -.start)

# --- 2. Turn "closest side" into an EMPIRICAL (tide-independent) direction ---
# Logic:
#   - a close detection BEFORE Fullerton => bird was elsewhere, then flew
#     THROUGH Fullerton => heading TO the mudflat
#   - a close detection AFTER Fullerton  => bird left Fullerton and was next
#     seen elsewhere => it had just come FROM the mudflat
#   - close detections on BOTH sides => ambiguous, can't use as ground truth
#   - no close detection on either side => can't validate this event at all
flyby_check <- flyby_check %>%
  mutate(
    empirical_direction = case_when(
      !is.na(prev_delay) & prev_delay <= flyby_window_min &
      !is.na(next_delay) & next_delay <= flyby_window_min   ~ "Ambiguous",
      !is.na(prev_delay) & prev_delay <= flyby_window_min   ~ "To mudflat (dropping tide)",
      !is.na(next_delay) & next_delay <= flyby_window_min   ~ "From mudflat (rising tide)",
      TRUE ~ NA_character_
    )
  )

# --- 3. Build the validation set and compute accuracy -------------------------
# Only events with an unambiguous, independently-determined direction can be
# used to check the tide rule.
validation_set <- flyby_check %>%
  filter(empirical_direction %in% c("To mudflat (dropping tide)", "From mudflat (rising tide)"))

n_validated  <- nrow(validation_set)
n_total_flyby <- nrow(flyby_check)

cat("Flybys with an independently-checkable direction:", n_validated,
    "out of", n_total_flyby, sprintf("(%.1f%%)", 100 * n_validated / n_total_flyby), "\n")

conf_mat <- table(
  tide_predicted = validation_set$direction,
  empirical       = validation_set$empirical_direction
)
print(conf_mat)

accuracy <- sum(diag(conf_mat)) / sum(conf_mat)
cat("Tide-rule accuracy against independently-confirmed direction:",
    sprintf("%.1f%%", 100 * accuracy), "\n")

# --- 4. (Optional but recommended) report accuracy AS A CAVEAT ON THE RESULTS -
# Whatever this number comes out to, put it directly in your ch1_13 write-up,
# e.g.: "Among the n_validated flybys where direction could be independently
# confirmed, the tide-based rule agreed with the confirmed direction in
# accuracy%% of cases." This turns an unstated assumption into a quantified,
# defensible limitation - which is exactly what makes the mudflat rate-of-use
# results in the last section trustworthy (or flags that they need
# a bigger grain of salt if accuracy is low, e.g. < 75%).

# --- 5. Optional: does anything beyond tide predict direction better? --------
# If accuracy is mediocre, check quickly whether hour-of-day adds information
# beyond tide state alone (birds may have diel roosting patterns on top of
# tide-driven movement). This stays simple: one small logistic regression,
# fit only on the validated (ground-truth) subset.
validation_glm <- validation_set %>%
  mutate(
    y         = if_else(empirical_direction == "From mudflat (rising tide)", 1, 0),
    hour      = lubridate::hour(visitStart),
    tide_pred = if_else(direction == "From mudflat (rising tide)", 1, 0)
  )

m_dir_tide_only <- glm(y ~ tide_pred,          data = validation_glm, family = binomial)
m_dir_plus_hour <- glm(y ~ tide_pred + hour,   data = validation_glm, family = binomial)

AIC(m_dir_tide_only, m_dir_plus_hour)   # lower AIC = better; tells you if hour-of-day is worth adding
summary(m_dir_plus_hour)

# If m_dir_plus_hour has meaningfully lower AIC and a significant `hour`
# term, it means tide alone is leaving predictable signal on the table -
# worth reporting, but NOT necessary to rebuild the whole direction/rate
# pipeline on; a one-line caveat in the text is enough to keep things simple.




# ============================================================================
# PLOT: HOUR-OF-DAY EFFECT ON DIRECTION (LOGISTIC MODEL)
# ============================================================================

# Requires: validation_glm, m_dir_plus_hour (from your GLM code above)
# Packages: ggplot2, dplyr, lubridate (already in your workflow)

library(ggplot2)
library(dplyr)
library(lubridate)

# --- 1. Create a data frame for predicted probabilities across hours ---------
# Build a grid of hour values (5–20) for each tide prediction state
hour_seq <- seq(5, 20, by = 0.5)

pred_grid <- expand.grid(
  hour = hour_seq,
  tide_pred = c(0, 1)  # 0 = "To mudflat", 1 = "From mudflat"
) %>%
  mutate(
    pred_prob = predict(
      m_dir_plus_hour,
      newdata = .,
      type = "response"
    ),
    tide_label = factor(
      tide_pred,
      levels = c(0, 1),
      labels = c("Predicted: To mudflat (dropping tide)",
                 "Predicted: From mudflat (rising tide)")
    )
  )

# --- 2. Prepare observed data for plotting -----------------------------------
plot_data <- validation_glm %>%
  mutate(
    hour = lubridate::hour(visitStart),
    obs_direction = factor(
      empirical_direction,
      levels = c("From mudflat (rising tide)", "To mudflat (dropping tide)")
    ),
    tide_label = factor(
      tide_pred,
      levels = c(0, 1),
      labels = c("Predicted: To mudflat (dropping tide)",
                 "Predicted: From mudflat (rising tide)")
    )
  )

# --- 3. Main plot: predicted curves + observed points ------------------------
p1 <- ggplot() +
  # Predicted logistic curves
  geom_line(
    data = pred_grid,
    aes(x = hour, y = pred_prob, colour = tide_label),
    linewidth = 1.2
  ) +
  # Observed data points
  geom_point(
    data = plot_data,
    aes(x = hour, y = as.numeric(empirical_direction == "From mudflat (rising tide)"),
        colour = tide_label),
    alpha = 0.5, size = 2, position = position_jitter(height = 0, width = 0.15)
  ) +
  scale_colour_manual(
    name = "Tide prediction",
    values = c("Predicted: To mudflat (dropping tide)" = "steelblue",
               "Predicted: From mudflat (rising tide)" = "firebrick")
  ) +
  labs(
    x = "Hour of day",
    y = "Probability: From mudflat (rising tide)",
    title = "Logistic Model: \nDirection vs Hour + Tide Prediction",
    subtitle = sprintf("AIC = %.2f | Hour β = %.3f (p = %.3f)",
                       AIC(m_dir_plus_hour),
                       coef(m_dir_plus_hour)["hour"],
                       summary(m_dir_plus_hour)$coefficients["hour", "Pr(>|z|)"])
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# --- 4. Optional: observed proportion by hour bins ---------------------------
# Bin hours into 2-hour intervals for a cleaner summary
plot_data_binned <- plot_data %>%
  mutate(
    hour_bin = cut(
      hour,
      breaks = seq(4, 22, by = 2),
      labels = paste(seq(4, 20, by = 2), seq(6, 22, by = 2), sep = "-"),
      include.lowest = TRUE
    )
  ) %>%
  group_by(hour_bin, .drop = FALSE) %>%
  summarise(
    prop_from_mudflat = mean(empirical_direction == "From mudflat (rising tide)"),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n > 0)  # drop empty bins

p2 <- ggplot(plot_data_binned, aes(x = hour_bin, y = prop_from_mudflat)) +
  geom_col(fill = "steelblue", alpha = 0.7, colour = "black") +
  geom_text(
    aes(label = paste0("n=", n)),
    vjust = -0.3, size = 3.5
  ) +
  labs(
    x = "Hour of day",
    y = "Proportion: From mudflat (rising tide)",
    title = "Observed Direction by Time of Day"
  ) +
  scale_y_continuous(limits = c(0, 1.05), breaks = seq(0, 1, by = 0.25)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "firebrick", linewidth = 1) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# --- 5. Display both plots side-by-side --------------------------------------
library(patchwork)  # or use gridExtra::grid.arrange(p1, p2, ncol = 2)

p_combined <- p1 + p2 +
  plot_layout(ncol = 2, widths = c(1.2, 1)) &
  theme(plot.margin = margin(10, 10, 10, 10))

print(p_combined)

# --- 7. One-line interpretation for your thesis ------------------------------
# You can extract and report:
cat(sprintf(
  "Among %d validated flybys, adding hour to the tide-only model reduced AIC from %.2f to %.2f.\n",
  nrow(validation_glm),
  AIC(m_dir_tide_only),
  AIC(m_dir_plus_hour)
))
cat(sprintf(
  "Hour coefficient: β = %.3f (SE = %.3f, p = %.3f).\n",
  coef(m_dir_plus_hour)["hour"],
  summary(m_dir_plus_hour)$coefficients["hour", "Std. Error"],
  summary(m_dir_plus_hour)$coefficients["hour", "Pr(>|z|)"]
))

```
:::
