---
title: "Site use - hours (OBP method)"
format: html
---

```{r style legend, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, include = FALSE, results = 'hide'}
# Source the script into that environment
source(knitr::purl(here::here("qmd", "chapter_1", "ch1_3.qmd"), 
                   output = tempfile(fileext = ".R"),    
                   quiet = TRUE))    
```

::: {.callout style="border-left: 6px solid #D99B8F; --bs-callout-border: #D99B8F; padding: 1.5rem; border-radius: 8px; margin-bottom: 1rem;"}
TO BE COMPLETED

Most of the analysis below is widely inspired from *Orange-bellied Parrot Migration Tracking Project, 2025 Report, Department of Natural Resources and Environment Tasmania*
:::

::: blockquote-blue
**Load your data** in your R environment *- see [Load & Format \> Reproducibility](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_1.html#reproducibility)*
:::

```{r packages,  message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
library(motus)
library(dplyr)
library(here)
library(ggridges)
library(tidyr)
library(readr)
library(dplyr)
library(ggplot2)
library(vegan)  
library(philentropy) 
library(gridExtra)
library(gt)
library(ggpubr)
library(car)
library(FSA)
library(purrr)
library(patchwork)
library(mixtools)
library(lubridate)
library(data.table)
```

```{r my data, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results = 'hide', include = FALSE}
# Birds
data_all <- readRDS(
  tail(sort(list.files(
    here::here("qmd", "chapter_1", "data", "motus"),
    pattern = "-data\\.rds$", full.names = TRUE
  )), 1)) 

# Receivers info
recv <- readRDS(
  tail(sort(list.files(
    here::here("qmd", "chapter_1", "data", "motus"),
    pattern = "-recv-info\\.rds$", full.names = TRUE
  )), 1)) 

# Receivers activity
sql.motus <- DBI::dbConnect(RSQLite::SQLite(), here::here("qmd", "chapter_1", "data", "project-294.motus"))
recv.act <- tbl(sql.motus, "activity")  %>% 
  collect() %>% 
  as.data.frame() %>%
  rename(deviceID = "motusDeviceID") %>%
  filter(deviceID %in% unique(recv$deviceID)) %>% # keep our deployed antennas only 
  
  # Set the time properly - IMPORTANT
  mutate(date = as_datetime(as.POSIXct(hourBin* 3600, origin = "1970-01-06", tz = "UTC")),
         dateAus = as_datetime(as.POSIXct(hourBin* 3600, origin = "1970-01-06", tz = "UTC"), 
                             tz = "Australia/Sydney")) 

# Tide
tide_data <- readRDS(here::here("qmd", "chapter_1", "data", "tides", "tideData.rds"))

```

::: blockquote-yellow
**Note**: As Fullerton Entrance is particular (single directional antenna), we will treat this case particularly and separately (see [Fullerton Entrance (Fly-by vs Site Use)](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_13.html)).
:::

```{r my data filter, message = FALSE, warning = FALSE, eval = FALSE, echo = TRUE, results = 'hide', include = FALSE}

data_all <- data_all %>%
  filter(recvDeployName != "Fullerton Entrance")

recv <- recv %>%
  filter(recvDeployName != "Fullerton Entrance")

```

# Site use adjusted for detection gap threshold

## Detection threshold

To define independent observations, we needed a time gap that separates a bird leaving a site from just briefly dropping out of antenna range (e.g., by turning away or moving behind vegetation). Short gaps likely mean the bird is still present; long gaps likely mean it has left. We analysed the distribution of time gaps between successive detection runs of the same bird at the same station, and across all individuals, and used this to choose a cut-off separating "same visit" from "new visit". A three-component mixture model (MIXTOOLS 2.0.0.1) was fitted to the gap durations, distinguishing short gaps (same visit) from medium/long gaps (new visit). Gaps were truncated to 5 min – 30 hours to remove noise from very short or very long outliers, and each gap was classified by whichever distribution best matched it.

**1. Build run-level summary** (one row per run) from hits. A "run" is already identified by runID in data_all [(see Motus documentation - Chapter 3)](https://motuswts.github.io/motus/articles/03-accessing-data.html). We need the start/end time of each run, per individual and per station.

```{r run , message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = TRUE}
runs <- data_all %>%
  group_by(speciesEN, Band.ID, recvDeployName, runID) %>%
  summarise(
    runStart = min(timeAus, na.rm = TRUE),
    runEnd   = max(timeAus, na.rm = TRUE),
    n_detect = n(), 
    .groups = "drop")
```

**2. Compute gaps** between successive runs of the same individual at the same antenna/site.

```{r gap inclusion, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = TRUE}
gaps <- runs %>%
  arrange(Band.ID, recvDeployName, runStart) %>%
  group_by(Band.ID, recvDeployName) %>%
  mutate(
    prevRunEnd = dplyr::lag(runEnd),
    gapHours   = round(as.numeric(difftime(runStart, prevRunEnd, units = "hours")), 2)) %>%
  ungroup() %>%
  filter(!is.na(gapHours)) %>%
  filter(gapHours > 0) # we are not dealing with overlaps yet (simultaneous detection of the same tag from two or more antenna)
```

**3. Truncate gaps** (hours) for mixture model fitting. 

::: blockquote-yellow **NOTE:** 
Below are values you should choose based on your own data and on the ecological plausibility of your system.
:::

```{r gap selection, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results = 'hide', include = TRUE}

gap_trunc <- gaps %>%
  filter(gapHours > 0.3,     # ~5 minutes                   
         gapHours < 96)      # 96 hours
   # pull(gapHours)

ggplot(gap_trunc, aes(x = gapHours, y = after_stat(density))) +
  geom_histogram(binwidth = 2, # 2h
                 fill = "grey30", 
                 color = "white") +
  scale_x_continuous(breaks = seq(0, 96, 5)) + # 5h
  labs(x = "Gap Duration (hours)", y = "density") +
  theme_minimal(base_size = 13) +
   annotate("rect", 
           xmin = 0, xmax = 30, 
           ymin = 0, ymax = Inf,   # spans full height of the plot
           fill = NA, color = "red", linewidth = 1) +
  theme(panel.grid.minor = element_blank()) 

ggplot(gap_trunc %>% filter(gapHours < 30), aes(x = gapHours, y = after_stat(density))) +
  geom_histogram(binwidth = 1, # 1h
                 fill = "grey30", 
                 color = "white") +
  scale_x_continuous(breaks = seq(0, 30, 1)) + # 1h
  labs(x = "Gap Duration (hours)", y = "density") +
  theme_minimal(base_size = 13) +
   annotate("rect", 
           xmin = 0, xmax = 15, 
           ymin = 0, ymax = Inf,   # spans full height of the plot
           fill = NA, color = "red", linewidth = 1) +
  theme(panel.grid.minor = element_blank())

ggplot(gap_trunc %>% filter(gapHours < 15), aes(x = gapHours, y = after_stat(density))) +
  geom_histogram(binwidth = 0.12, # ~5min
                 fill = "grey30", 
                 color = "white") +
  scale_x_continuous(breaks = seq(0, 15, 1)) + # 1h
  labs(x = "Gap Duration (hours)", y = "density") +
  theme_minimal(base_size = 13) +
   annotate("rect", 
           xmin = 0, xmax = 2, 
           ymin = 0, ymax = Inf,   # spans full height of the plot
           fill = NA, color = "red", linewidth = 1) +
  theme(panel.grid.minor = element_blank())    # /!\ BASED ON VISUAL GRAPH, I SELECT SUBJECTIVELY 1h AS TRESHOLD - conservative as 2h could be /!\

```

**4. Fit a 3-component normal mixture model**. Now we have visually explored how data is distributed to select a gap threshold value, we might use statistical model to assess this value more systematical and based on the data only, besides visual selection accounting for ecologically plausibility.

```{r gap selection 2, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE,, include = TRUE}

gap_trunc <- gaps %>%
  filter(gapHours > 0.3,     # ~5 minutes                   
         gapHours < 30) %>%  # 30 hours
  pull(gapHours)

set.seed(123)  # for reproducibility of EM starting values

mix_model <- normalmixEM(
  log(gap_trunc +1),         # +1 to avoid log(0), and log() used to compress/scale duration distributions and tails (i.e., lot of short vs few long)
  k = 3,
  maxit = 5000,
  epsilon = 1e-08)

summary(mix_model)

```

**6. Find the threshold value,** the cut-point: the gap value where classification switches from "Non-independent" to an "Independent" class (i.e. where posterior prob of the non-independent component drops below 0.5 as gap duration increases). We will first classify each gap to the component with highest posterior probability, and order components by their mean, so component 1 = shortest-gap (non-independent), 2 = medium, 3 = long (independent).

```{r gap selection 3, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results = 'hide', include = TRUE}
posteriors <- mix_model$posterior
classification <- apply(posteriors, 1, which.max)
comp_order <- order(mix_model$mu)
comp_labels <- c("Non-independent", "Independent (medium)", "Independent (long)")
names(comp_labels) <- as.character(comp_order)

class_df <- data.frame(
  gapHours = gap_trunc,
  component = comp_labels[as.character(classification)])

nonindep_comp <- comp_order[1]

cutpoint_df <- data.frame(
  gapHours = gap_trunc,
  p_nonindep = posteriors[, nonindep_comp]) %>%
  arrange(gapHours)

# first point where posterior probability of non-independence < 0.5
cutpoint <- cutpoint_df$gapHours[which(cutpoint_df$p_nonindep < 0.5)[1]]
cat("Threshold value or Cut-point (hours):", round(cutpoint, 2), "\n")

# Plot
x_grid <- seq(0, 30, length.out = 500)

dens_df <- do.call(rbind, lapply(seq_along(mix_model$mu), function(i) {
  data.frame(
    x = x_grid,
    y = mix_model$lambda[i] * dnorm(x_grid, mix_model$mu[i], mix_model$sigma[i]),
    component = comp_labels[as.character(i)])
}))

comp_colors <- c(
  "Non-independent"      = "darkred",
  "Independent (medium)" = "lightblue",
  "Independent (long)"   = "darkgreen")

ggplot() +
  geom_histogram(
    data = data.frame(gapHours = gap_trunc),
    aes(x = gapHours, y = after_stat(density)),
    binwidth = 1, fill = "grey30", color = "white", boundary = 0) +
  geom_line(
    data = dens_df,
    aes(x = x, y = y, color = component),
    linewidth = 1) +
  geom_vline(xintercept = cutpoint, color = "red", linewidth = 0.6) +
  scale_color_manual(values = comp_colors) +
  scale_x_continuous(breaks = seq(0, 30, 2)) +
  labs(
    x = "Gap Duration (hours)",
    y = "density",
    color = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank())

```

## Define used time (hours)

Now, we decide to merge into one and unique continuous detection, all the tags (birds) detected per a same antenna, including gaps under a duration of 1 (one) hour. Although the model is highlighting a value of `r  round(cutpoint, 2)`, we decided to remain conservative.

**Once detection data summarised as described above, we can see, overall, how long birds stay around one Motus station** - see plot below.

```{r site use, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = TRUE}

visit_duration <- runs %>%
  arrange(Band.ID, recvDeployName, runStart) %>%
  group_by(Band.ID, recvDeployName) %>%
  mutate(
    gapHours = round(as.numeric(runStart - dplyr::lag(runEnd))  / 3600, 2),
    new_visit = is.na(gapHours) | gapHours >= 1,                 # <-- HERE IS THE THRESHOLD VALUE TO SET A "NEW VISIT" (TRUE when gapHours >= 1) 
    visitID = cumsum(new_visit)) %>%
  group_by(Band.ID, speciesEN, recvDeployName, visitID) %>%
  summarise(
    visitStart = min(runStart),
    visitEnd   = max(runEnd),
    n_detect   = sum(n_detect),                                  # total hits/detection across merged runs
    n_runs     = n(),                                            # how many runs got merged into this visit
    .groups = "drop") %>%
  
  mutate(duration_h = round(as.numeric(visitEnd - visitStart) / 3600, 2))    # visit duration in hours


ggplot(visit_duration %>%
         filter(duration_h < 20),
       aes(x = duration_h, y = after_stat(count))) +
  geom_histogram(binwidth = 0.5, 
                 fill = "grey30", 
                 color = "white") +
  scale_x_continuous(breaks = seq(0, 20, 1)) + # 1h
  labs(x = "Visit Duration (hours)", y = "Count (n)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())    

```

\*Note Newcastle NSW tides are turning \~6h cycles between high/low, and our birds seem to mostly visits any site less than 3 hours. An early descriptive sign of site use very tide dependent, as they would stay about 3 hours in the same area, around the highest/lowest point of the tide line, before to leave site, as the tide might be rising or receiving.\*

```{r site use filter, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, include = TRUE}
ggplot(visit_duration %>%
         filter(duration_h > 20),
       aes(x = duration_h, y = after_stat(count))) +
  geom_histogram(binwidth = 25, 
                 boundary = 20,       
                 fill = "grey30", 
                 color = "white") +
  scale_x_continuous(breaks = sort(c(20, seq(500, 3000, 500)))) +
  coord_cartesian(xlim = c(0, 3000)) +
  labs(x = "Visit Duration (hours)", y = "Count (n)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

To avoid inflating site use duration estimates with the very few **individuals showing more than 25 hours of continuous presence in the same area, we excluded these cases**. Such extended duration likely reflect a dropped tag rather than genuine site fidelity. Verifying each case individually would be time-consuming, so we treat this as a working assumption for now; should this pattern become more frequent in future data, it would warrant closer investigation.

```{r site use filter 2, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = TRUE}
visit_duration <- visit_duration %>%
         filter(duration_h < 25)
```

**It remains to attribute at each visit period a tide category.** We expand each visit into hourly bins spanning `visitStart` to `visitEnd` (`visit_duration` itself is untouched). Then, we compute exact overlap (in hours) between each hourly bin and the actual visit window.

```{r site use tide, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = TRUE}
visit_hours <- visit_duration %>%
  rowwise() %>%
  mutate(hour_dt = list(seq(from = floor_date(visitStart, "hour"),
                             to   = floor_date(visitEnd, "hour"),
                             by   = "hour"))) %>%
  unnest(cols = c(hour_dt)) %>%
  ungroup()

visit_hours <- visit_hours %>%
  mutate(
    bin_start     = hour_dt,
    bin_end       = hour_dt + hours(1),
    overlap_start = pmax(visitStart, bin_start),
    overlap_end   = pmin(visitEnd, bin_end),
    overlap_h     = as.numeric(difftime(overlap_end, overlap_start, units = "hours"))
  ) %>%
  filter(overlap_h > 0)   # drops bins with no actual overlap

# Attach nearest tide category to each hourly bin
tide_dt <- as.data.table(tide_data)[, .(tideDateTimeAus, tideCategory)]
setkey(tide_dt, tideDateTimeAus)

vh_dt <- as.data.table(visit_hours)
setkey(vh_dt, hour_dt)

visit_hours_tide <- tide_dt[vh_dt, on = .(tideDateTimeAus = hour_dt), roll = "nearest"]
setnames(visit_hours_tide, "tideDateTimeAus", "hour_dt")

```

**Let's finalise the table giving data for detection relative to the *trio* bird-station-tide.**

```{r site use final table, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = TRUE}

used_bird_recv_time <- visit_hours_tide %>%
  mutate(tideDiel = if_else(grepl("Diurnal", tideCategory), "Diurnal", "Nocturnal"),
         tideHighLow = if_else(grepl("High", tideCategory), "High", "Low")) %>%
  group_by(Band.ID, speciesEN, recvDeployName, tideCategory, tideDiel, tideHighLow) %>%
  summarise(duration_h = sum(overlap_h), .groups = "drop")

```

## Define available time (hours)

Now used time addressed, we need to define for each individual (birds) and across each Motus stations, what was the amount of time available for the four tide categories.

Available time for each Motus stations is accessed with `activity` table: each row is a dated record of noise or a tag detected *- see [Survey effort \> Activity table](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_5.html#activity-table)*.

We need to extract the temporal coverage of each station, starting from deployment date to the last recorded activity (noise or detection).

```{r Available vs Used RECV AVAILABLE, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE, results = 'asis'}

# Sort the terminated serno (if terminated, ie. one box removed from one antenna site, a date comes along)
# but still needed for accessing survey effort as the station is currently running with another serno 
recv.act.term <- recv.act %>%
  left_join(recv %>% 
              filter(!is.na(timeEndAus)) %>%
              select(deviceID, serno, recvDeployName),
            "deviceID") %>%
  filter(!is.na(recvDeployName)) %>%
  mutate(SernoStation = paste0(recvDeployName, "_", serno))

# Sort the currently running serno
recv.act.runn <- recv.act %>%
  left_join(recv %>% 
              filter(is.na(timeEndAus)) %>%
              select(deviceID, serno, recvDeployName),
            "deviceID") %>%
  filter(!is.na(recvDeployName)) %>%
  mutate(SernoStation = paste0(recvDeployName, "_", serno))

# Merging in one data-set to use Station's name further + pick-up the rounded hours
recv.act <- bind_rows(recv.act.runn, recv.act.term) %>%
  mutate(hour_dt = round_date(dateAus, "hour"))

# Providing helpful variables
recv <- recv %>%
  mutate(SernoStation = paste0(recvDeployName, "_", serno),
         lisStart = timeStartAus,
         lisEnd = if_else(
           is.na(timeEndAus), # means the station is still running since the last data downloading
           with_tz(Sys.time(), "Australia/Sydney"),
           with_tz(as_datetime(timeEndAus, tz = "UTC"), "Australia/Sydney")) )

# Generating hourly sequences per SernoStation from start to end dates of the deviceID at particular sites
recv_hours <- recv %>%
  select(recvDeployName, deviceID, SernoStation, lisStart, lisEnd) %>%
  group_by(SernoStation) %>%
  rowwise() %>%
  mutate(hour_dt = list(seq(from = round_date(lisStart, unit = "hour"),
                            to = round_date(lisEnd, unit = "hour"),
                            by = "hour")) ) %>%
  unnest(cols = c(hour_dt)) %>%
  ungroup()

# Simplify station variables (recvDeployName)
recv <- recv %>% 
  select(!recvDeployName) 
recv$recvDeployName <- sub("_SG-.*", "", recv$SernoStation)

recv_hours <- recv_hours %>% 
  select(!recvDeployName) 
recv_hours$recvDeployName <- sub("_SG-.*", "", recv_hours$SernoStation)


```

<br>

We now have the complete temporal coverage, expanded in hours and for each Motus station so it matches with our used time dataset.

However, due to some temporary failures or maintenance *(see [Survey effort \> Operational periods](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_5.html#operational-periods))*, the actual period a station was operational may be different than the complete temporal coverage. So, we must determine when a station was not operational and substrate these off-hours to the complete temporal coverage.

<br>

```{r Available vs Used RECV OFF PERIODS, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE, results = 'asis'}


# Distinguish station from mixed recv + Giving operational variable (= TRUE when existing values from act table)
recv_hours <- recv_hours %>%
  select(-any_of("operational")) %>%   # drop if it already exists, no-op if it doesn't
  left_join(recv.act %>%
              distinct(recvDeployName, hour_dt) %>%
              mutate(operational = TRUE),
            by = c("recvDeployName", "hour_dt")) %>%
  mutate(operational = if_else(is.na(operational), FALSE, TRUE))

# /!\ Due to unknown error? Have to set this manually
recv_hours <- recv_hours %>%
  mutate(operational = case_when(
    recvDeployName == "Fullerton Entrance" & 
      hour_dt > as.POSIXct("2023-04-02") & 
      hour_dt < as.POSIXct("2023-04-05") ~ FALSE,
    TRUE ~ operational))

# Only the hours the station was actually ON — computed once, no regeneration of gaps
valid_station_hours <- recv_hours %>%
  filter(operational == TRUE) %>%
  distinct(recvDeployName, hour_dt)

# Attach nearest tide category directly (this also folds in the Point 7 fix —
tide_dt <- as.data.table(tide_data)[, .(tideDateTimeAus, tideCategory)]
setkey(tide_dt, tideDateTimeAus)

vsh_dt <- as.data.table(valid_station_hours)
setkey(vsh_dt, hour_dt)

valid_station_hours_tide <- tide_dt[vsh_dt, on = .(tideDateTimeAus = hour_dt), roll = "nearest"]
setnames(valid_station_hours_tide, "tideDateTimeAus", "hour_dt")

```

<br>

Temporal coverage for each station is now reduced to its operational periods. **We now assign each bird its own available time period**, spanning from its tagging date to its last Motus detection, and expanded to hourly resolution too.

<br>

```{r Available vs Used BIRD AVAILABLE TIME, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE, results = 'asis'}


# Get the monitored period of each bird
period_sp <- data_all  %>%
  group_by(Band.ID) %>%
  reframe(DateAUS.Trap = dplyr::first(DateAUS.Trap), 
          last_dateAus = max(timeAus),
          speciesEN = dplyr::first(speciesEN)) %>%
  unique()

# Expand one row per hours to each individual across its whole period (this is the available time)
bird_hours <- period_sp %>%
  group_by(Band.ID) %>%
  rowwise() %>%
  mutate(hour_dt = list(seq(from = as.POSIXct(ymd(DateAUS.Trap), tz = "UTC"),
                            to = as.POSIXct(last_dateAus, tz = "UTC"),
                            by = "hour"))) %>%
  unnest(cols = c(hour_dt)) %>%
  ungroup() 

# Duplicate in as many list as many stations
recv_names <- unique(valid_station_hours_tide$recvDeployName)
bird_data_list <- setNames(vector("list", length(recv_names)), recv_names)

for(name in recv_names) {
  valid_hours_recv <- valid_station_hours_tide %>%
    filter(recvDeployName == name) %>%
    pull(hour_dt)
  
  valid_hours_bird_recv <- bird_hours %>%
    filter(hour_dt %in% valid_hours_recv)
  
  valid_hours_bird_recv$name <- name
  bird_data_list[[name]] <- valid_hours_bird_recv
}

# # FINALISE AVAILABLE TIME 
available_bird_recv_time <- bind_rows(bird_data_list) %>%
  mutate(hour_dt = force_tz(hour_dt, "Australia/Sydney")) %>%
  filter(if_all(everything(), ~ !is.na(.))) %>%
  rename(recvDeployName = name) %>%
  left_join(valid_station_hours_tide, by = c("recvDeployName", "hour_dt")) %>%
  filter(!is.na(tideCategory)) %>%    # bird-hours the station wasn't actually ON for get dropped here
  group_by(Band.ID, speciesEN, recvDeployName, tideCategory) %>%
  summarise(duration_h = n(), .groups = "drop") %>%
  mutate(tideDiel = if_else(grepl("Diurnal", tideCategory), "Diurnal", "Nocturnal"),
         tideHighLow = if_else(grepl("High", tideCategory), "High", "Low")) %>%
  select(Band.ID, speciesEN, recvDeployName, tideCategory, tideDiel, tideHighLow, duration_h)
```

::: blockquote-blue
**The *available* time is now defined** for each bird across each Motus station, taking into account the periods where stations where ON only. In the data, the *available* periods are expanded in hours and the tide condition is associated. While most of the couple station listening-bird tagged are potentially detectable between 0 and 1000 hours; note two small peaks around 2700 and 4000 hours of detectability potential as it might be relative to the oldest couple ON.
:::

```{r Available plot, message = FALSE, warning = FALSE, echo = FALSE, include = TRUE,, eval = TRUE }
ggplot() +
  geom_histogram(
    data = available_bird_recv_time,
    aes(x = duration_h, y = after_stat(count)),
    binwidth = 100, 
    fill = "grey30", 
    color = "white", 
    boundary = 0) +
  scale_x_continuous(breaks = seq(0, 4000, 500)) +
  labs(
    x = "Listening Duration for a typical Motus station (hours per birds) [all data]",
    y = "Count (n)",
    color = NULL) +
     annotate("rect", 
           xmin = 0, xmax = 1000, 
           ymin = 0, ymax = Inf,   # spans full height of the plot
           fill = NA, color = "red", linewidth = 1) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank())

ggplot() +
  geom_histogram(
    data = available_bird_recv_time %>%
      filter(duration_h < 1000),
    aes(x = duration_h, y = after_stat(count)),
    binwidth = 100, 
    fill = "grey30", 
    color = "white", 
    boundary = 0) +
  scale_x_continuous(breaks = seq(0, 1000, 100)) +
  labs(
    x = "Listening Duration for a typical Motus station (hours per birds) [0-1000 hours]",
    y = "Count (n)",
    color = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank())

ggplot() +
  geom_histogram(
    data = available_bird_recv_time %>%
      filter(duration_h > 1000),
    aes(x = duration_h, y = after_stat(count)),
    binwidth = 100, 
    fill = "grey30", 
    color = "white", 
    boundary = 0) +
  scale_x_continuous(breaks = seq(0, 4000, 1000)) +
  labs(
    x = "Listening Duration for a typical Motus station (hours per birds) [1000-Inf hours]",
    y = "Count (n)",
    color = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank())

```

## Define rate of Use (%)

Now that the *used* and *available* time have been determined for each individual and for each Motus station, we can relate these two elements together with the **Usage rate** (or rate of use).

::: {style="overflow-x: auto; width: 100%;"}
$$
\text{Usage rate} = \frac{\text{Time used}}{\text{Time available}} \times 100 \quad \textit{across stations and species} 
$$
:::

```{r final table, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE}
use_rate_plot <- used_bird_recv_time %>%
  select(-tideDiel, -tideHighLow) %>%   # drop here, keep the copy from available_bird_recv_time
  rename(used_h = duration_h) %>%
  
  left_join(available_bird_recv_time %>%
      rename(available_h = duration_h),   
    by = c("Band.ID", "speciesEN", "recvDeployName", "tideCategory")) %>%
  
  mutate(rate_use = (used_h / available_h) * 100) %>%
  select(Band.ID, speciesEN, recvDeployName, tideCategory, 
         tideDiel, tideHighLow, used_h, available_h, rate_use)

```

```{r potential error, message = FALSE, warning = FALSE, echo = FALSE, eval = TRUE}


above <- use_rate_plot %>%
  filter(rate_use > 100) %>%
  mutate(used_h = round(used_h, 2),
         available_h = round(available_h, 2),
         rate_use = round(rate_use, 2),
         diff_perc = rate_use - 100,
         diff_h = round(used_h - available_h, 2)) 

```

::: blockquote-yellow
**Note**: A very few cases (n = `r nrow(above)` / `r nrow(use_rate_plot)`) where the *time used* was over the *time* *available* has been re-allocated the value 100%. Reasons might be the tag fell off within range of the antenna. Due to the negligible number of such cases, we have not further investigated. In average, the difference of time between available and used is `r mean(above$diff_h)` hours (`r mean(above$diff_perc)` % of rate of use difference).
:::

```{r potential error 2, message = FALSE, warning = FALSE, echo = FALSE, eval = TRUE}

above %>%
  DT::datatable()

```

```{r potential error filtered, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE}

use_rate_plot <- use_rate_plot %>%
  mutate(rate_use = ifelse(rate_use > 100, 100, rate_use))

```

::: blockquote-yellow
**Note:** We filtered out Masked Lapwing from the analysis, since they have too few data.  
We also filter out **Fullerton Entrance** station as it needs to be worked separately (see [Site use - hours (Fullerton case)](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_13.html))
:::

```{r Time Used Rate, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE, results = 'asis'}

figure_plot <- use_rate_plot %>%
  mutate(speciesType = factor(shorebird_class[speciesEN], 
                               levels = c("migratory", "resident"))) %>%
  
  filter(!speciesEN %in% c("Masked Lapwing"),
         !recvDeployName %in% c("Fullerton Entrance")) 

```

## Results

**The plot below shows you how much each species have been detected over the array, by night and by day.** First displayed overall. Second displayed per condition and showing individual variation.

```{r 4 Used rate tot, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
# Per-bird mean rate_use by species and tide/diel category
per_bird <- use_rate_plot %>%
  group_by(Band.ID, speciesEN, tideDiel) %>%
  summarise(rate_use = mean(rate_use, na.rm = TRUE), .groups = "drop")

# Per-bird "Undetected" category
other_per_bird <- per_bird %>%
  group_by(Band.ID, speciesEN) %>%
  summarise(rate_use = 100 - sum(rate_use, na.rm = TRUE), .groups = "drop") %>%
  mutate(tideDiel = "Undetected")

# Combine per-bird data (this was missing)
per_bird_full <- bind_rows(per_bird, other_per_bird) %>%
  mutate(tideDiel = factor(tideDiel, levels = c("Undetected", "Diurnal", "Nocturnal")))

# --- Stacked bar (species-level means, from per-bird data) ---
plot_data_full <- per_bird_full %>%
  group_by(speciesEN, tideDiel) %>%
  summarise(mean_rate_use = mean(rate_use, na.rm = TRUE), .groups = "drop")

ggplot(plot_data_full, aes(x = speciesEN, y = mean_rate_use, fill = tideDiel)) +
  geom_bar(stat = "identity", position = "stack", width = 0.7) +
  scale_fill_manual(values = c("Nocturnal" = "#2C3E50",
                                "Diurnal" = "#F4A300",
                                "Undetected" = "#B0B0B0"),
                    name = "Period",
                    breaks = c("Nocturnal", "Diurnal", "Undetected")) +
  labs(x = "Species", y = "Use rate (%)",
       title = "Use rate by species") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")
```

::: {.callout-tip collapse="true" title="See data table"}
```{r 4 Used rate tot table, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}


library(dplyr)
library(gt)

# Build summary stats per species x tideDiel
summary_tbl <- per_bird_full %>%
  group_by(speciesEN, tideDiel) %>%
  summarise(
    mean   = mean(rate_use, na.rm = TRUE),
    sd     = sd(rate_use, na.rm = TRUE),
    median = median(rate_use, na.rm = TRUE),
    min    = min(rate_use, na.rm = TRUE),
    max    = max(rate_use, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(speciesEN, tideDiel)

# Build the gt table
summary_tbl %>%
  gt() %>%
  tab_header(title = "Statistics") %>%
  opt_align_table_header(align = "left") %>%
  fmt_number(
    columns = c(mean, sd, median, min, max),
    decimals = 2
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(speciesEN))
  ) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "darkgrey",
      weight = px(2)
    ),
    locations = cells_body(
      rows = which(!duplicated(summary_tbl$speciesEN))[-1]
    )
  )

```
:::

::: {.callout-tip collapse="true" title="See raw data table"}
```{r 4 Used rate tot raw table, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}

per_bird_full %>%
  DT::datatable()
```
:::

```{r 4 Used rate tot 2, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
# --- Dodged bars + individual points (per-bird variation) ---
summary_data <- per_bird_full %>%
  group_by(speciesEN, tideDiel) %>%
  summarise(mean_rate = mean(rate_use, na.rm = TRUE),
            se = sd(rate_use, na.rm = TRUE) / sqrt(sum(!is.na(rate_use))),
            .groups = "drop")

ggplot() +
  geom_bar(data = summary_data,
           aes(x = speciesEN, y = mean_rate, fill = tideDiel),
           stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(data = summary_data,
                aes(x = speciesEN, ymin = pmax(mean_rate - se, 0), ymax = mean_rate + se,
                    group = tideDiel),
                position = position_dodge(width = 0.8), width = 0.2) +
  geom_point(data = per_bird_full,
             aes(x = speciesEN, y = rate_use, group = tideDiel),
             position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.15),
             size = 1.2, alpha = 0.4, color = "black") +
  scale_fill_manual(values = c("Nocturnal" = "#2C3E50",
                                "Diurnal" = "#F4A300",
                                "Undetected" = "#B0B0B0"),
                    name = "Period") +
  labs(x = "Species", y = "Use rate (%)",
       title = "Use rate by species",
       subtitle = "Bars = mean ± SE, points = individual birds") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")
```

::: {.callout-tip collapse="true" title="See data table"}
```{r 4 Used rate tot table 2, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
library(gt)

summary_data %>%
  gt() %>%
  tab_header(title = "Statistics") %>%
  opt_align_table_header(align = "left") %>%
  fmt_number(
    columns = c(mean_rate, se),
    decimals = 2
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(speciesEN))
  ) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "darkgrey",
      weight = px(2)),
    locations = cells_body(
      rows = which(!duplicated(summary_data$speciesEN))[-1]
    )
  )
```
:::

::: {.callout-tip collapse="true" title="See raw data table"}
```{r 4 Used rate tot raw table 2, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
summary_data %>% 
  DT::datatable()
```
:::

```{r tool, message = FALSE, warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
# Track size sample
counts <- data_all %>%
      group_by(speciesEN) %>%
      summarise(n = n_distinct(Band.ID)) %>%
      mutate(label = paste0(speciesEN, " (n = ", n, ")"))
label_vec <- setNames(counts$label, counts$speciesEN)

# Tide and species groupings
tide_levels <- c("Low", "High")
species_types <- unique(shorebird_class)  

# Simplified function - no min_n_nonzero check
make_plot <- function(tide_level, species_type) {
  data_sub <- figure_plot %>%
    filter(tideHighLow == tide_level) %>%
    mutate(species_class = shorebird_class[speciesEN]) %>%
    filter(species_class == species_type)
  
  p <- ggplot(data_sub,
              aes(x    = factor(recvDeployName, levels = sort(unique(recvDeployName))),
                  y    = rate_use,
                  fill = tideDiel))
  
  # Always add boxplot for all available data
  p <- p + 
    geom_boxplot(outlier.shape = NA,
                 varwidth = FALSE,
                 position = position_dodge(width = 0.8, preserve = "single")) +
    
    # Always show individual points
    geom_point(aes(shape = tideDiel), 
               position = position_dodge(width = 0.8),
               alpha = 1, size = 1.5,
               show.legend = FALSE) +
    scale_shape_manual(values = c("Diurnal" = 21, "Nocturnal" = 16)) +
    
    facet_wrap(~ speciesEN,
               labeller = labeller(speciesEN = label_vec)) +
    labs(x     = "Receiver Deployment",
         y     = "Rate of Use (%)",
         title = paste(ifelse(species_type == "migratory",
                              "Migratory species", "Resident species"),
                       "during", tide_level, "tide")) +
    coord_cartesian(ylim = c(0, 100)) +
    theme_minimal() +
    scale_fill_manual(values = c("Diurnal" = "white", "Nocturnal" = "darkgrey")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  return(p)
}

# Generate all plots
plots_used_rate <- purrr::cross2(tide_levels, species_types) %>%
  purrr::map(~ make_plot(.x[[1]], .x[[2]]))








make_plot_v <- function(tide_level, species_type) {
  data_sub <- figure_plot %>%
    filter(tideHighLow == tide_level) %>%
    mutate(species_class = shorebird_class[speciesEN]) %>%
    filter(species_class == species_type)
  
  p <- ggplot(data_sub,
              aes(x    = factor(recvDeployName, levels = sort(unique(recvDeployName))),
                  y    = rate_use,
                  fill = tideDiel))
  
  # Violin plot instead
  p <- p + 
    geom_violin(outlier.shape = NA,
                trim = TRUE,
                varwidth = FALSE,
                scale = "width",
                width = 0.8,
                position = position_dodge(width = 1.2, preserve = "single"),
                alpha = 0.7) +
    
    # Always show individual points
    geom_point(aes(shape = tideDiel), 
               position = position_dodge(width = 0.8),
               alpha = 1, size = 1.5,
               show.legend = FALSE) +
    scale_shape_manual(values = c("Diurnal" = 21, "Nocturnal" = 16)) +
    
    facet_wrap(~ speciesEN,
               labeller = labeller(speciesEN = label_vec)) +
    labs(x     = "Receiver Deployment",
         y     = "Rate of Use (%)",
         title = paste(ifelse(species_type == "migratory",
                              "Migratory species", "Resident species"),
                       "during", tide_level, "tide")) +
    coord_cartesian(ylim = c(0, 100)) +
    theme_minimal() +
    scale_fill_manual(values = c("Diurnal" = "white", "Nocturnal" = "darkgrey")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  return(p)
}

# Generate all plots
plots_used_rate_v <- purrr::cross2(tide_levels, species_types) %>%
  purrr::map(~ make_plot_v(.x[[1]], .x[[2]]))

```

```{r stats for plots, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}

# General function to provide one table with stats values
make_summary_table <- function(tide_level, species_type) {
  data_sub <- figure_plot %>%
    filter(tideHighLow == tide_level) %>%
    mutate(species_class = shorebird_class[speciesEN]) %>%
    filter(species_class == species_type)
  
  summary_table <- data_sub %>%
    group_by(speciesEN, recvDeployName, tideDiel) %>%
    summarise(
      n_individual = n_distinct(Band.ID),
      min = round(min(rate_use, na.rm = TRUE), 1),
      q1 = round(quantile(rate_use, 0.25, na.rm = TRUE),1),
      median = round(median(rate_use, na.rm = TRUE),1),
      q3 = round(quantile(rate_use, 0.75, na.rm = TRUE),1),
      max = round(max(rate_use, na.rm = TRUE),1),
      mean = round(mean(rate_use, na.rm = TRUE),1),
      sd = round(sd(rate_use, na.rm = TRUE),1),
      .groups = 'drop'
    ) %>%
    arrange(speciesEN, recvDeployName, tideDiel) %>%
    rename(Species = "speciesEN", Station = "recvDeployName", condition = "tideDiel")
  
  return(summary_table)
}

# One table per conditions
high_migratory <- make_summary_table("High", "migratory")
high_resident <- make_summary_table("High", "resident")
low_migratory <- make_summary_table("Low", "migratory")
low_resident <- make_summary_table("Low", "resident")
```

### Resident species

::::: panel-tabset
##### High Tide

```{r 4 Used rate 1, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
plots_used_rate[[4]]
plots_used_rate_v[[4]]
```

::: {.callout-tip collapse="true" title="High Tide Resident birds"}
```{r 4 Used rate table, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
high_resident  %>%
  gt() %>%
  tab_header(title = "Statistics - High Tide Resident Species") %>%
  opt_align_table_header(align = "left") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(Species))) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "darkgrey",
      weight = px(2)),
    locations = cells_body(
      rows = which(!duplicated(high_resident$Species))[-1]))
```
:::

##### Low Tide

```{r 4 Used rate 2, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}

plots_used_rate[[3]]
plots_used_rate_v[[3]]

```

::: {.callout-tip collapse="true" title="Low Tide Resident birds"}
```{r 4 Used rate table 2, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
low_resident  %>%
  gt() %>%
  tab_header(title = "Statistics - Low Tide Resident Species") %>%
  opt_align_table_header(align = "left") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(Species))) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "darkgrey",
      weight = px(2)),
    locations = cells_body(
      rows = which(!duplicated(low_resident$Species))[-1]))
```
:::
:::::

### Migrant species

::::: panel-tabset
##### High Tide

```{r 4 Used rate 3, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
plots_used_rate[[2]]
plots_used_rate_v[[2]]

```

::: {.callout-tip collapse="true" title="High Tide Migratory birds"}
```{r 4 Used rate table 3, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
high_migratory  %>%
  gt() %>%
  tab_header(title = "Statistics - High Tide Migratory Species") %>%
  opt_align_table_header(align = "left") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(Species))) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "darkgrey",
      weight = px(2)),
    locations = cells_body(
      rows = which(!duplicated(high_migratory$Species))[-1]))
```
:::

##### Low Tide

```{r 4 Used rate 4, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}

plots_used_rate[[1]]
plots_used_rate_v[[1]]

```

::: {.callout-tip collapse="true" title="Low Tide Migratory birds"}
```{r 4 Used rate table 5, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
low_migratory  %>%
  gt() %>%
  tab_header(title = "Statistics - Low Tide Migratory Species") %>%
  opt_align_table_header(align = "left") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(Species))) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "darkgrey",
      weight = px(2)),
    locations = cells_body(
      rows = which(!duplicated(low_migratory$Species))[-1]))
```
:::
:::::

## *Beta \|* Statistical analysis (GLM)

**EXPLORATORY**

Instead of simple "*the boxplots look different*" we statistically analyse with GLM the data to estimate tidal and circadian cycles effects per species, correctly weighted and accounting for the fact that a same bird may appears many times.

### Additive model (direction \~ tide \* species + diel \* species)

```{r GLM 2 ways, message = FALSE,  warning = FALSE, echo = TRUE, eval = TRUE, results = 'hide'}
# ============================================================================
# STATISTICAL MODEL OF SITE-USE RATE
# Question: does rate of use (used_h / available_h) differ by species,
# depending on tide state (High/Low) and diel period (Diurnal/Nocturnal)?
# ============================================================================

# --- WHY THIS MODEL AND NOT A BOXPLOT/T-TEST -------------------------------
# rate_use is a PROPORTION (used_h / available_h) with very unequal
# denominators: some bird x station x condition combos have available_h = 3h,
# others have available_h = 2000h. A boxplot / Wilcoxon test on rate_use
# treats a 100% rate based on 3h exactly the same as a 100% rate based on
# 500h - which is statistically wrong. A binomial-family model uses
# available_h as the number of "trials" per row, so noisy/low-effort
# estimates are automatically down-weighted relative to well-sampled ones.
# We also add Band.ID as a random effect because the same bird contributes
# many rows (non-independence across stations/conditions for one individual).

library(dplyr)
library(ggplot2)
library(glmmTMB)   # mixed models with beta-binomial family (handles proportion data + overdispersion)
library(DHARMa)    # residual diagnostics for GLMM(s)
library(emmeans)   # marginal means & pairwise post-hoc comparisons, back-transformed to % scale
library(car)       # Anova() omnibus tests
library(patchwork) # combine plots

# ==== 1. Prepare data ========================================================
# betabinomial() wants integer-like "successes" and "trials", so we convert
# hours to minutes and round. This just gives the model finer resolution
# than whole hours - it does NOT change what used_h/available_h represent.
model_data <- figure_plot %>%
  mutate(
    used_min      = round(used_h * 60),
    available_min = round(available_h * 60)
  ) %>%
  filter(available_min > 0,                  # can't divide by zero exposure
         used_min <= available_min) %>%       # safety check, should already be true
  droplevels()                                # drop unused factor levels (e.g. Masked Lapwing already filtered)

# ==== 2. Fit the model =======================================================
# Fixed effects: speciesEN interacted separately with tideDiel and with
# tideHighLow, so we can ask "does the day/night effect differ by species?"
# and "does the tide-height effect differ by species?" without blowing up
# the model with a full 3-way interaction (keeps it simple, as requested).
# Random effects: (1|Band.ID) accounts for repeated measures per bird;
# (1|recvDeployName) accounts for station-to-station baseline differences
# without spending a fixed-effect parameter per station (partial pooling -
# more robust than a fixed effect when some stations have very few birds).
m_site_use <- glmmTMB(
  cbind(used_min, available_min - used_min) ~
    speciesEN * tideDiel +
    speciesEN * tideHighLow +
    (1 | Band.ID) +
    (1 | recvDeployName),
  family = betabinomial(),    # like binomial, but allows extra-binomial variance (very common in ecological count/proportion data)
  data   = model_data
)

summary(m_site_use)

# ==== 3. Check the model actually fits reasonably ===========================
# DHARMa simulates data from the fitted model and compares it to your real
# data: if the model is adequate, the simulated residuals should look
# uniform. This is a quick, honest sanity check before trusting any p-value.
sim_res <- simulateResiduals(m_site_use, n = 500)
plot(sim_res)                     # QQ-plot + residual-vs-predicted plot
testDispersion(sim_res)           # formal test: is there still over/under-dispersion left?

# If the dispersion test or QQ-plot looks bad, the most common simple fixes are:
#   - drop the recvDeployName random effect if it's estimated as ~0 (singular fit warning)
#   - simplify to speciesEN + tideDiel + tideHighLow (no interactions) if you have
#     convergence warnings, then only add interactions back one at a time.

# ==== 4. Omnibus test: does each term matter overall? =======================
# Answers "does species matter at all", "does tideDiel matter at all", etc.,
# collapsing across factor levels - useful before diving into pairwise
# comparisons below.
Anova(m_site_use, type = "II")

# ==== 5. Pairwise comparisons: the actual biological question ===============
# (a) For each species, is Diurnal use different from Nocturnal use?
emm_diel <- emmeans(m_site_use, ~ tideDiel | speciesEN, type = "response")
diel_contrasts <- as.data.frame(pairs(emm_diel))   # keep as object: reused for stars below
print(diel_contrasts)

# (b) For each species, is High-tide use different from Low-tide use?
emm_tide <- emmeans(m_site_use, ~ tideHighLow | speciesEN, type = "response")
tide_contrasts <- as.data.frame(pairs(emm_tide))   # keep as object: reused for stars below
print(tide_contrasts)

# (c) Optional: are species different from each other overall (within a given
#     diel/tide condition)? Useful if you want a compact "who uses the array
#     most" ranking.
emm_species <- emmeans(m_site_use, ~ speciesEN, type = "response")
pairs(emm_species)

# ==== 6. Plot MODEL-ESTIMATED rate of use, with significance stars ==========

# --- helper: convert a p-value to conventional significance stars ----------
p_to_stars <- function(p) {
  dplyr::case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    TRUE      ~ "ns"      # "not significant"
  )
}

# --- helper: emmeans calls the back-transformed estimate column "prob" for
# binomial/beta-binomial models, and "response" for Gaussian-family models.
# This makes downstream code robust to either, so you don't need to remember
# which one applies for a given model family.
standardise_response_col <- function(df) {
  if ("prob" %in% names(df))          df <- df %>% rename(response = prob)
  else if (!"response" %in% names(df)) stop("No back-transformed estimate column found - check names(df).")
  df
}

# ---- 6a. Diel plot ----------------------------------------------------------
plot_df_diel <- standardise_response_col(as.data.frame(emm_diel))

sig_diel <- diel_contrasts %>%
  select(speciesEN, p.value) %>%
  mutate(stars = p_to_stars(p.value))

# one star label per species, placed just above that species' highest CI
label_y_diel <- plot_df_diel %>%
  group_by(speciesEN) %>%
  summarise(y_label = max(asymp.UCL) * 100 + 5, .groups = "drop") %>%
  left_join(sig_diel, by = "speciesEN")

p_diel <- ggplot(plot_df_diel, aes(x = speciesEN, y = response * 100,
                                    ymin = asymp.LCL * 100, ymax = asymp.UCL * 100,
                                    color = tideDiel)) +
  geom_pointrange(position = position_dodge(width = 0.5), size = 0.6) +
  geom_text(data = label_y_diel, aes(x = speciesEN, y = y_label, label = stars),
            inherit.aes = FALSE, size = 4, fontface = "bold") +
  labs(x = "Species", y = "Model-estimated rate of use (%) [95% CI]",
       color = "Period",
       title = "Modelled site-use rate by species and diel period",
       subtitle = "Beta-binomial GLMM | stars = significance of Diurnal vs Nocturnal per species\n*** p<0.001, ** p<0.01, * p<0.05, ns = not significant") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_diel

# ---- 6b. Tide-height plot, built the same way -------------------------------
plot_df_tide <- standardise_response_col(as.data.frame(emm_tide))

sig_tide <- tide_contrasts %>%
  select(speciesEN, p.value) %>%
  mutate(stars = p_to_stars(p.value))

label_y_tide <- plot_df_tide %>%
  group_by(speciesEN) %>%
  summarise(y_label = max(asymp.UCL) * 100 + 5, .groups = "drop") %>%
  left_join(sig_tide, by = "speciesEN")

p_tide <- ggplot(plot_df_tide, aes(x = speciesEN, y = response * 100,
                                    ymin = asymp.LCL * 100, ymax = asymp.UCL * 100,
                                    color = tideHighLow)) +
  geom_pointrange(position = position_dodge(width = 0.5), size = 0.6) +
  geom_text(data = label_y_tide, aes(x = speciesEN, y = y_label, label = stars),
            inherit.aes = FALSE, size = 4, fontface = "bold") +
  labs(x = "Species", y = "Model-estimated rate of use (%) [95% CI]",
       color = "Tide",
       title = "Modelled site-use rate by species and tide height",
       subtitle = "Beta-binomial GLMM | stars = significance of High vs Low tide per species\n*** p<0.001, ** p<0.01, * p<0.05, ns = not significant") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_tide

# ---- 6c. Combined view -------------------------------------------------------
p_diel / p_tide     # stacked vertically; use p_diel | p_tide for side-by-side

```

### Mixte model (direction \~ tide \* diel \* species)

```{r GLM 3 ways, message = FALSE,  warning = FALSE, echo = TRUE, eval = TRUE, results = 'hide'}

# ============================================================================
# THREE-WAY INTERACTION MODEL
# Question: does the tide-height effect on rate of use depend on BOTH
# species AND diel period simultaneously? i.e. for each species, do we get
# four genuinely different estimates - Low Night, Low Day, High Night,
# High Day - or do some of these collapse into "not significantly different"?
# ============================================================================

# --- WHY THIS IS A BIGGER ASK THAN THE TWO SEPARATE TWO-WAY MODELS ----------
# speciesEN * tideDiel * tideHighLow estimates a separate mean for every
# species x diel x tide combination (4 conditions per species). This needs
# enough data in EVERY one of those cells to be trustworthy; some species or
# stations you've already seen are thin (n=1-2 birds), so expect wider CIs,
# some "ns" results, and possibly convergence warnings for the rarer species.
# This is the necessary trade-off for asking a more specific question.

library(conflicted)   # makes function-masking conflicts (e.g. select()) fail loudly instead of silently
library(dplyr)
library(ggplot2)
library(glmmTMB)
library(DHARMa)
library(emmeans)
library(car)
library(multcomp)     # for cld() - compact letter display of pairwise significance; also loads MASS,
                       # which is what silently masked dplyr::select() before - now handled below

# --- Lock in which version of commonly-masked functions to use. Run this
# once, right after all libraries are loaded, so it doesn't matter what
# order packages were loaded in, or what gets loaded later in the session.
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::filter)

# ==== 1. Prepare data =========================================================
model_data <- figure_plot %>%
  mutate(
    used_min      = round(used_h * 60),
    available_min = round(available_h * 60)
  ) %>%
  filter(available_min > 0,                  # can't divide by zero exposure
         used_min <= available_min) %>%       # safety check, should already be true
  droplevels()                                # drop unused factor levels (e.g. Masked Lapwing already filtered)

# ==== 2. Fit the 3-way interaction model ======================================
# Fixed effect: full three-way interaction speciesEN * tideDiel * tideHighLow.
# Random effects unchanged: Band.ID for repeated measures per bird,
# recvDeployName for station-level baseline differences.
m_site_use_3way <- glmmTMB(
  cbind(used_min, available_min - used_min) ~
    speciesEN * tideDiel * tideHighLow +
    (1 | Band.ID) +
    (1 | recvDeployName),
  family = betabinomial(),
  data   = model_data
)

summary(m_site_use_3way)

# ==== 3. Model fit check =======================================================
sim_res_3way <- simulateResiduals(m_site_use_3way, n = 500)
plot(sim_res_3way)
testDispersion(sim_res_3way)

# If this throws convergence warnings (likely, given thin cells for some
# species), the simplest fallback is the two separate two-way models -
# that's the data telling you it can't support this many parameters yet,
# not a coding error.

# ==== 4. Omnibus test: is the 3-way interaction itself significant? ==========
# The key new test the two separate two-way models couldn't ask: does the
# tide-height effect change shape between day and night, and does THAT
# pattern differ by species? Look at the speciesEN:tideDiel:tideHighLow row.
Anova(m_site_use_3way, type = "II")
# If that row is NOT significant, the simpler two-way models remain the
# better-supported choice - the full 3-way term isn't needed.

# ==== 5. Pairwise comparisons across all 4 conditions per species ============
# For each species: 4 diel x tide combinations, compared pairwise
# (6 comparisons per species).
emm_3way <- emmeans(m_site_use_3way, ~ tideDiel * tideHighLow | speciesEN, type = "response")
contrasts_3way <- as.data.frame(pairs(emm_3way))
print(contrasts_3way)

# --- Compact Letter Display (cld): with 4 groups per species, individual
# stars per pair get cluttered fast. cld assigns a letter to each of the 4
# conditions per species; conditions sharing NO letter are significantly
# different, conditions sharing a letter are not. Standard way to summarise
# >2-group comparisons in one readable tag.
cld_3way <- cld(emm_3way, Letters = letters, adjust = "tukey") %>%
  as.data.frame()

# ==== 6. Plot: one panel per species, FOUR estimates per panel ===============
# (Low Night, Low Day, High Night, High Day) on a single x-axis

# --- helper: emmeans calls the back-transformed estimate column "prob" for
# binomial/beta-binomial models. Standardise so downstream code doesn't
# need to remember which name applies.
standardise_response_col <- function(df) {
  if ("prob" %in% names(df))          df <- df %>% rename(response = prob)
  else if (!"response" %in% names(df)) stop("No back-transformed estimate column found - check names(df).")
  df
}

# --- helper: collapse tideDiel + tideHighLow into one readable 4-level factor
combine_condition <- function(df) {
  df %>%
    mutate(
      condition = case_when(
        tideHighLow == "Low"  & tideDiel == "Nocturnal" ~ "Low Night",
        tideHighLow == "Low"  & tideDiel == "Diurnal"   ~ "Low Day",
        tideHighLow == "High" & tideDiel == "Nocturnal" ~ "High Night",
        tideHighLow == "High" & tideDiel == "Diurnal"   ~ "High Day"
      ),
      # fix the order left-to-right so it reads logically rather than alphabetically
      condition = factor(condition, levels = c("Low Night", "Low Day", "High Night", "High Day"))
    )
}

# --- 6a. Build the plotting data: estimates + CIs, with the 4-level condition
plot_df_3way <- standardise_response_col(as.data.frame(emm_3way)) %>%
  combine_condition()

# --- 6b. Build the matching letter-label data, same condition encoding
cld_3way <- cld_3way %>% combine_condition()

# --- 6c. Join significance letters onto the plotting data
# (select() explicitly namespaced - see note below the script for why)
plot_df_3way <- plot_df_3way %>%
  left_join(cld_3way %>% dplyr::select(speciesEN, condition, .group),
            by = c("speciesEN", "condition"))

# --- 6d. Plot: ONE x-axis with all 4 conditions, one panel per species
ggplot(plot_df_3way, aes(x = condition, y = response * 100,
                          ymin = asymp.LCL * 100, ymax = asymp.UCL * 100)) +
  geom_pointrange(size = 0.6, color = "#6CA0DC") +
  geom_text(aes(label = trimws(.group), y = asymp.UCL * 100 + 4),
            size = 3.2, fontface = "bold", color = "#2C3E50") +
  facet_wrap(~ speciesEN, scales = "free_y") +
  labs(x = NULL, y = "Model-estimated rate of use (%) [95% CI]",
       title = "Modelled site-use rate: Low/High tide x Night/Day, per species",
       subtitle = "Beta-binomial GLMM | letters = Tukey-adjusted compact letter display\nconditions sharing a letter are NOT significantly different",
       caption = "Letters shared between conditions = no significant difference; letters not shared = significant difference\n (p < 0.05, Tukey-adjusted).") +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.caption = element_text(hjust = 0, size = 9, color = "grey30", face = "italic"))

```

### Complete model (direction \~ tide \* diel \* species \* station)

```{r GLM complete, message = FALSE,  warning = FALSE, echo = TRUE, eval = TRUE, results = 'hide'}

# ============================================================================
# FOUR-WAY MODEL: species x station x tide x diel
# Question: for each species, does site-use rate differ by station AND by
# tide/diel condition simultaneously?
# ============================================================================

# --- IMPORTANT CAVEAT --------------------------------------------------------
# This is a much larger model than the 3-way one. Many species only visit a
# handful of stations, and several species-station cells have very few birds
# (n=1-2, as seen in the earlier summary tables). Expect:
#   - wide confidence intervals for thin cells
#   - possible convergence warnings
#   - some panels essentially uninformative (huge CIs, all "a")
# This is the model correctly reflecting real data limits, not a bug.
# If it doesn't converge, the documented fallback is at the bottom.

library(conflicted)
library(dplyr)
library(ggplot2)
library(glmmTMB)
library(DHARMa)
library(emmeans)
library(car)
library(multcomp)

conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::filter)

# ==== 1. Prepare data =========================================================
model_data <- figure_plot %>%
  mutate(
    used_min      = round(used_h * 60),
    available_min = round(available_h * 60)
  ) %>%
  filter(available_min > 0,
         used_min <= available_min) %>%
  droplevels()

# ==== 2. Fit the model =========================================================
# Fixed effect: full 4-way interaction speciesEN * recvDeployName * tideDiel * tideHighLow.
# Random effect: (1|Band.ID) only - recvDeployName is now a FIXED term (it's
# part of the interaction we want estimates for), so it's no longer also a
# random effect (that would double-count it).
m_site_use_4way <- glmmTMB(
  cbind(used_min, available_min - used_min) ~
    speciesEN * recvDeployName * tideDiel * tideHighLow +
    (1 | Band.ID),
  family = betabinomial(),
  data   = model_data
)

summary(m_site_use_4way)

# --- FALLBACK if the above fails to converge or throws a rank-deficiency
# warning (very possible given the sparsity noted above): drop the full
# 4-way interaction down to pairwise interactions only, which is much more
# parsimonious and usually converges fine, while still letting station
# effects vary by species and by tide/diel:
#
# m_site_use_4way <- glmmTMB(
#   cbind(used_min, available_min - used_min) ~
#     speciesEN * recvDeployName +
#     speciesEN * tideDiel + speciesEN * tideHighLow +
#     recvDeployName * tideDiel + recvDeployName * tideHighLow +
#     (1 | Band.ID),
#   family = betabinomial(),
#   data   = model_data
# )

# ==== 3. Model fit check ========================================================
sim_res_4way <- simulateResiduals(m_site_use_4way, n = 500)
plot(sim_res_4way)
testDispersion(sim_res_4way)

# ==== 4. Omnibus test ============================================================
Anova(m_site_use_4way, type = "II")
# Look at the speciesEN:recvDeployName:tideDiel:tideHighLow row specifically.
# If it (or lower-order terms) aren't significant, the simpler 3-way model
# from before remains the better-supported, more honest choice.

# ==== 5. Get estimates - RESTRICTED TO COMBINATIONS THAT ACTUALLY EXIST ========
# This step is critical: emmeans by default will produce an estimate for
# EVERY theoretical species x station x tide x diel combination, including
# ones with zero real birds (e.g. a species that never visited that station).
# Those would be pure model extrapolation - not real information. We compute
# the full grid, then keep only rows matching combinations actually present
# in model_data.
emm_4way <- emmeans(m_site_use_4way,
                     ~ tideDiel * tideHighLow | speciesEN * recvDeployName,
                     type = "response")

observed_combos <- model_data %>%
  distinct(speciesEN, recvDeployName, tideDiel, tideHighLow)

emm_4way_df <- as.data.frame(emm_4way) %>%
  inner_join(observed_combos, by = c("speciesEN", "recvDeployName", "tideDiel", "tideHighLow"))

nrow(as.data.frame(emm_4way))   # theoretical grid size
nrow(emm_4way_df)               # actual, data-backed grid size - expect this much smaller

# ==== 6. Pairwise comparisons + compact letter display, per species x station ===
# Same logic as before: for each species x station panel, compare the 4
# tide/diel conditions pairwise, letters show which pairs are/aren't
# significantly different. Restricted to observed combos only, same as above.
contrasts_4way <- as.data.frame(pairs(emm_4way)) %>%
  semi_join(observed_combos, by = c("speciesEN", "recvDeployName"))
print(contrasts_4way)

cld_4way <- cld(emm_4way, Letters = letters, adjust = "tukey") %>%
  as.data.frame() %>%
  inner_join(observed_combos, by = c("speciesEN", "recvDeployName", "tideDiel", "tideHighLow"))

# ==== 7. Plot: one row per species, one column per station, 4 bars per panel ===

standardise_response_col <- function(df) {
  if ("prob" %in% names(df))          df <- df %>% rename(response = prob)
  else if (!"response" %in% names(df)) stop("No back-transformed estimate column found - check names(df).")
  df
}

combine_condition <- function(df) {
  df %>%
    mutate(
      condition = case_when(
        tideHighLow == "Low"  & tideDiel == "Nocturnal" ~ "Low Night",
        tideHighLow == "Low"  & tideDiel == "Diurnal"   ~ "Low Day",
        tideHighLow == "High" & tideDiel == "Nocturnal" ~ "High Night",
        tideHighLow == "High" & tideDiel == "Diurnal"   ~ "High Day"
      ),
      condition = factor(condition, levels = c("Low Night", "Low Day", "High Night", "High Day"))
    )
}

plot_df_4way <- standardise_response_col(emm_4way_df) %>% combine_condition()
cld_4way     <- cld_4way %>% combine_condition()

plot_df_4way <- plot_df_4way %>%
  left_join(cld_4way %>% dplyr::select(speciesEN, recvDeployName, condition, .group),
            by = c("speciesEN", "recvDeployName", "condition"))

# # facet_grid gives species as rows and station as columns - matches "one
# # panel per species, split by station" more literally than facet_wrap would,
# # and keeps species groupings visually clear as row labels down the left.
# ggplot(plot_df_4way, aes(x = condition, y = response * 100,
#                           ymin = asymp.LCL * 100, ymax = asymp.UCL * 100)) +
#   geom_pointrange(size = 0.5, color = "#6CA0DC") +
#   geom_text(aes(label = trimws(.group), y = asymp.UCL * 100 + 4),
#             size = 2.8, fontface = "bold", color = "#2C3E50") +
#   facet_grid(speciesEN ~ recvDeployName, scales = "free_y") +
#   labs(x = NULL, y = "Model-estimated rate of use (%) [95% CI]",
#        title = "Modelled site-use rate: Low/High tide x Night/Day, per species and station",
#        subtitle = "Beta-binomial GLMM | letters = Tukey-adjusted compact letter display",
#        caption = "Letters shared between conditions = no significant difference; letters not shared = significant difference (p < 0.05, Tukey-adjusted).\nOnly species-station combinations present in the data are shown.") +
#   theme_minimal(base_size = 10) +
#   theme(strip.text = element_text(face = "bold", size = 8),
#         axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
#         plot.caption = element_text(hjust = 0, size = 8, color = "grey30", face = "italic"))


# ============================================================================
# ONE PLOT PER SPECIES, each faceted by recvDeployName
# Produces a NAMED LIST of ggplot objects, so you can call them individually
# in a panel-tabset the same way you already use plots_used_rate[[1]] etc.
# ============================================================================

# --- (assumes plot_df_4way already built exactly as in the previous script:
#      species x station x tideDiel x tideHighLow estimates, restricted to
#      observed combinations, with the .group significance letters attached)

standardise_response_col <- function(df) {
  if ("prob" %in% names(df))          df <- df %>% rename(response = prob)
  else if (!"response" %in% names(df)) stop("No back-transformed estimate column found - check names(df).")
  df
}

combine_condition <- function(df) {
  df %>%
    mutate(
      condition = case_when(
        tideHighLow == "Low"  & tideDiel == "Nocturnal" ~ "Low Night",
        tideHighLow == "Low"  & tideDiel == "Diurnal"   ~ "Low Day",
        tideHighLow == "High" & tideDiel == "Nocturnal" ~ "High Night",
        tideHighLow == "High" & tideDiel == "Diurnal"   ~ "High Day"
      ),
      condition = factor(condition, levels = c("Low Night", "Low Day", "High Night", "High Day"))
    )
}

# --- helper: build ONE species' plot, faceted by station -------------------
make_species_plot <- function(data, sp) {
  d <- data %>% filter(speciesEN == sp) %>% droplevels()

  ggplot(d, aes(x = condition, y = response * 100,
                ymin = asymp.LCL * 100, ymax = asymp.UCL * 100)) +
    geom_pointrange(size = 0.6, color = "#6CA0DC") +
    geom_text(aes(label = trimws(.group), y = asymp.UCL * 100 + 4),
              size = 3.2, fontface = "bold", color = "#2C3E50") +
    facet_wrap(~ recvDeployName) +
    coord_cartesian(ylim = c(0, 100)) +                   # cap the y-axis at 0-100%
    labs(x = NULL, y = "Model-estimated rate of use (%) [95% CI]",
         title = paste("Modelled site-use rate -", sp),
         subtitle = "Beta-binomial GLMM | letters = Tukey-adjusted compact letter display",
         caption = "Letters shared between conditions = no significant difference; letters not shared = significant difference (p < 0.05, Tukey-adjusted).\nOnly station x condition combinations present in the data are shown.") +
    theme_minimal(base_size = 12) +
    theme(strip.text = element_text(face = "bold"),
          axis.text.x = element_text(angle = 45, hjust = 1),
          plot.caption = element_text(hjust = 0, size = 8, color = "grey30", face = "italic"))
}

# --- build the plotting data once (same logic as before) --------------------
plot_df_4way <- standardise_response_col(emm_4way_df) %>% combine_condition()
cld_4way_j   <- cld_4way %>% combine_condition()

plot_df_4way <- plot_df_4way %>%
  left_join(cld_4way_j %>% dplyr::select(speciesEN, recvDeployName, condition, .group),
            by = c("speciesEN", "recvDeployName", "condition"))

# --- generate the named list: one entry per species --------------------------
species_list <- sort(unique(plot_df_4way$speciesEN))

species_plots <- setNames(
  lapply(species_list, function(sp) make_species_plot(plot_df_4way, sp)),
  species_list
)

# --- usage in your panel-tabset, e.g.: ---------------------------------------
# species_plots[["Bar-tailed Godwit"]]
# species_plots[[1]]   # same thing, by position instead of name

```

::: panel-tabset
```{r 4 glm per species, message = FALSE,  warning = FALSE, echo = TRUE, eval = TRUE, results = 'asis'}

for (sp in names(species_plots)) {
  cat("##### ", sp, "\n\n")   # creates one tab, titled with the species name
  print(species_plots[[sp]])  # renders that species' plot inside the tab
  cat("\n\n")                 # blank line so the next tab marker is parsed correctly
}
```
:::
