---
title: "Site use - hours (Fullerton case)"
format: html
---

::: {.callout style="border-left: 6px solid #D99B8F; --bs-callout-border: #D99B8F; padding: 1.5rem; border-radius: 8px; margin-bottom: 1rem;"}
**Why this station is treated differently**

Fullerton Entrance carries a single directional Yagi antenna (6-elements). Unlike the small omni-directional antennas used across the rest of the Motus array in the Hunter estuary (see [Motus Array Maintenance](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_4.html)), and capable of tracking site use directly, Fullerton Entrance station is located strategically: it sits on the corridor between the main area (estuary covered by several omni antennas) and the biggest and main mudflat of the area ([Fullerton Cove](https://www.google.com/maps/place/32%C2%B050'03.3%22S+151%C2%B047'40.2%22E/@-32.8437716,151.7726019,11867m/data=!3m1!1e3!4m4!3m3!8m2!3d-32.834252!4d151.794489!5m1!1e4?entry=ttu&g_ep=EgoyMDI2MDgyMy4wIKXMDSoASAFQAw%3D%3D)), which has no antenna coverage at all because of its terrain configuration. So, in order to track movement and site use over this big mudflat (Fullerton Cove, or also called Fullerton mudflat), we placed a station right at the entry ([Fullerton Entrance](https://motus.org/dashboard/#e=profile&d=stations&s=13119&f=%7B%22dtStart%22%3A%222025-08-26%22%7D)), to detect in-and-out passages.

Therefore, a tag detection at Fullerton Entrance station is either:

-   **fly-by**, the bird briefly fly by (in front of) the antenna, probably traveling between the estuary (roost?) and Fullerton mudflat (foraging?)

-   **site use** - the bird is actually sitting by the antenna itself, either roosting or foraging, using the site.

Since we cannot detect birds directly on Fullerton mudflat, we use fly-by detections at Fullerton Entrance station and their direction (another station → Fullerton Entrance or another station → Fullerton Entrance) as an **indirect proxy for mudflat use**.
:::

## Set-up data

As **we will be re-using some of the data already loaded from previous workflow (see [Station usage - hours (OBP method)](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_12.html)**), we will fetch here some of the objects we will re-use including: *visit_duration, tide_data, data_all, recv, shorebird_class, and available_bird_recv_time*.

```{r set-up 1, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, include = FALSE, results = 'hide'}
# Source Station usage - hours (OBP method) script into this environment
source(knitr::purl(here::here("qmd", "chapter_1", "ch1_12.qmd"),
                   output = tempfile(fileext = ".R"),
                   quiet = FALSE))
```

```{r local packages, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
library(mixtools)
library(lubridate)
library(gt)
library(lubridate)
library(ggpattern)
```

Now, we pre-fill some **key-values** we need to set-up the workflow.

```{r fc visits, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
focal_station    <- "Fullerton Entrance"   # the directional antenna

flyby_window_min <- 30                     # Period (min) before/after detection at Fullerton to look for a detection at another station

tide_lookahead_h <- 2                      # Period (hours) before/after of an "unknown" flyby detection used to infer direction from tide

# Keep only Fullerton Entrance data with visit duration (min)
fc_visits <- visit_duration %>%
  filter(recvDeployName == focal_station) %>%
  mutate(duration_min = duration_h * 60)
```

## When is it a fly-by?

A very short visit (seconds to a couple of minutes) most likely reflects a bird flying by the antenna; while a longer visit likely reflects a bird actually using the site covered by antenna's range coverage. We use the same mixture-model approach as for the detection-gap threshold (see [Station usage - hours (OBP method)](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_12.html)), but here fitted to **visit durations** at Fullerton Entrance only, to find a data-driven cut-off between "flyby" and "actual use".

**Fit a 2-component mixture model** on log-transformed visit duration (component 1 = short/flyby, component 2 = long/actual use), and find the cut-point where the posterior probability of belonging to the "flyby" component drops below 0.5.

```{r fc mixture model, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
set.seed(123)

mix_fc <- normalmixEM(               # we assume normal distributions only
  log(fc_visits$duration_min + 1),   # +1 to avoid log(0), and log() used to compress/scale duration distributions and tails (i.e., lot of short vs few long)
  k = 2,                             # we assume two potential distribution (flyby and site use)
  maxit = 5000,                      # number of iteration to estimate what component each observation belogs to
  epsilon = 1e-08)                   # standard threshold value for estimation (convergence tolerance)

comp_order_fc <- order(mix_fc$mu)    # sort components -> comp_1 with smaller mean first (fly-by)

post_short    <- mix_fc$posterior[, comp_order_fc[1]] # extract first col of mix_fc$posterior, a matrix with one row per visit and one column per component, containing each visit's posterior probability of belonging to that component (rows sum to 1). So, post_short is a vector giving one value per visit of "probability this visit is a fly-by."

cut_df <- data.frame(duration_h = fc_visits$duration_h, p_short = post_short) %>%
  arrange(duration_h)

flyby_cutpoint_h <- cut_df$duration_h[which(cut_df$p_short < 0.5)[1]] # /!\ the value chosen here (0.5) finds the first duration in hour (smallest) where it has less than 50 % probability to be a fly-by /!\
```

Under `r flyby_cutpoint_h` hours (\~ `r round(flyby_cutpoint_h * 60)` min), a detection at Fullerton Entrance is considered as a *fly-by* and above as a *site use*. So, let's classify and visualise.

```{r fc classify visits, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
fc_visits <- fc_visits %>%
  mutate(visit_type = if_else(duration_h < flyby_cutpoint_h, "flyby", "actual_use"))
```

```{r fc mixture plot, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
ggplot(fc_visits, aes(x = duration_h, fill = visit_type)) +
  geom_histogram(binwidth = 0.1, color = "white", boundary = 0) +
  geom_vline(xintercept = flyby_cutpoint_h, color = "red", linetype = "dashed", linewidth = 0.7) +
  scale_fill_manual(values = c(flyby = "#E7298A", actual_use = "#66A61E")) +
  coord_cartesian(xlim = c(0, 10)) +
  labs(x = "Visit duration (hours)", y = "Count", fill = "Visit type",
       title = paste("Flyby vs actual-use classification -", focal_station),
       subtitle = paste0("Cut-point = ", round(flyby_cutpoint_h * 60, 1), " min (mixture model)")) +
  theme_minimal(base_size = 13)

ggplot(fc_visits %>%
         filter(duration_h > flyby_cutpoint_h), aes(x = duration_h, fill = visit_type)) +
  geom_histogram(binwidth = 0.1, color = "white", boundary = 0) +
  coord_cartesian(xlim = c(0, 10)) +
  labs(x = "Visit duration (hours)", y = "Count", fill = "Visit type",
       title = paste("Actual-use distribution -", focal_station),
       subtitle = paste0("Visit during more than ", round(flyby_cutpoint_h * 60, 1), " min (cut-point from mixture model)")) +
  theme_minimal(base_size = 13)

```

## Fly-by direction

For flyby events specifically, we want to know whether the bird was **heading to the mudflat** (last seen leaving the roost area, i.e. detected at another station shortly *before* the flyby) or **coming from the mudflat** (detected at another station shortly *after* the flyby). To do so, we need to understand what is the right temporal window to consider for defining the "shortly after" or "shortly before", and to rely on inference only when the detection into anothe Motus station is beyond this value.

Let's start by splitting the data-set accordingly before to estimate the direction of a fly-by depending on the tide: toward the mudflat of Fullerton cove if the fly-by is detected during a falling tide, and inversly if detected during a rising tide.

```{r flyby set, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
flyby_visits <- fc_visits %>% filter(visit_type == "flyby")

all_visits_dt   <- as.data.table(visit_duration)
other_visits_dt <- all_visits_dt[recvDeployName != focal_station]
flyby_dt        <- as.data.table(flyby_visits)
```

```{r classification: infer, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}

flyby_dt <- flyby_dt %>%
  
  rowwise() %>%
  
  mutate(
    next_tide = list(
      tide_data %>%
        filter(tideDateTimeAus >= visitStart,
               tideDateTimeAus <= visitStart + lubridate::hours(10)) %>%
        slice_min(tideDateTimeAus, n = 1)), # takes the next tide peak from NSW forecast
    
    nearestNextDiel = next_tide$day_night[1],
    nearestNextTide = next_tide$high_low[1]) %>%
  
  ungroup() %>%
  select(-next_tide) %>%
  
  mutate(
    direction = case_when(
      nearestNextTide == "High" ~ "From mudflat (rising tide)",
      nearestNextTide == "Low"  ~ "To mudflat (falling tide)",
      TRUE ~ NA_character_ ))


barplot(table(flyby_dt$direction, useNA = "ifany"),
        main = "Inferred flyby direction based on next tide",
        ylab = "Count of flybys")

```

For each fly-by detected at Fullerton Entrance, we then look for a **confirming detection at another Motus station**: if the tide was rising, we check for the *closest following* detection after Fullerton Entrance; if the tide was falling, we check for the *closest preceding* detection before Fullerton Entrance.

```{r classification: verified, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}

flyby_dt <- flyby_dt %>%
  
  rowwise() %>%
  
  mutate(
    .band    = Band.ID,
    .start   = visitStart,
    .station = recvDeployName,
    
    other_visit = list(
      if (!is.na(direction) && direction == "From mudflat (rising tide)") {
        all_visits_dt %>%
          filter(Band.ID == .band,
                 recvDeployName != .station,
                 visitStart >= .start) %>%
          slice_min(visitStart, n = 1)
      } else if (!is.na(direction) && direction == "To mudflat (falling tide)") {
        all_visits_dt %>%
          filter(Band.ID == .band,
                 recvDeployName != .station,
                 visitStart <= .start) %>%
          slice_max(visitStart, n = 1)
      } else {
        all_visits_dt %>% filter(FALSE)
      }
    ),
    side_detection = other_visit$recvDeployName[1],
    delay = as.numeric(abs(difftime(.start, other_visit$visitStart[1], units = "mins")))
  ) %>%
  ungroup() %>%
  select(-other_visit, -.band, -.start, -.station)

```

We then look at the **distribution of the delay** between a detectiona at Fullerton entrance and another Motus station, in order to identify a temporal window we can use as a cut-off for concluding that *"the bird flew continuously between the fly-by detection at Fullerton Entrance and the next detection at another Motus station"* (or, in the reverse case, between a detection at another station and the fly-by at Fullerton entrance during a falling tide).

We first filter out any extreme tail in the data, then visualise the distribution to estimate how many underlying components it contains.

```{r delay tail, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
flyby_dt %>%
  arrange(desc(delay)) %>%
  select(Band.ID, visitStart, direction, side_detection, delay) %>%
  head(20) %>%
  DT::datatable()
```

```{r filter and plot, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
flyby_dt_trimmed  <- flyby_dt %>% 
  filter(flyby_dt$delay < 4000)

hist(flyby_dt_trimmed$delay,
     breaks = 50,
     main  = "Delay to (or period before) nearest next site detection",
     xlab  = "Delay (minutes)")
```

**Visually,** we can estimate that the overall temporal window for continuous flights between the Fullerton entrance and another Motus station is about 10 minutes. However, we can fit an N-mixture model to confirm this.

```{r extract cutpoint delay, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}

delay_vals <- na.omit(flyby_dt_trimmed$delay)
set.seed(1)  # EM is stochastic in its starting values
mix <- normalmixEM(delay_vals, 
                   k = 2)
plot(mix, which = 2)

# order components so mu1 < mu2 (short-delay vs long-delay cluster)
ord    <- order(mix$mu)
mu     <- mix$mu[ord]
sigma  <- mix$sigma[ord]
lambda <- mix$lambda[ord]

# find where the two weighted densities cross, between the two means
f <- function(x) lambda[1]*dnorm(x, mu[1], sigma[1]) - lambda[2]*dnorm(x, mu[2], sigma[2])
flyby_window_min <- uniroot(f, lower = mu[1], upper = mu[2])$root
flyby_window_min

```

With a statistically determined **threshold**, any delay between a detection at the Fullerton entrance and a another Motus station that is shorter than `r flyby_window_min` ​ minutes will be treated as a *verified* flight direction for the bird. Beyond this value, a continuous flight is not considered realistic, and detections at other Motus stations cannot be trusted to infer the bird’s flight direction. See below for the distribution of these results.

```{r fc flyby set, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
flyby_dt <- flyby_dt %>%
  mutate(retained_cut = case_when(
    # is.na(delay)                    ~ "inferred_only",
    delay <= flyby_window_min       ~ "Verified by another Motus station detection",
    TRUE                            ~ "Inferred only from tide data"))
```

```{r plot directions, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
bp_data <- t(table(flyby_dt$direction, flyby_dt$retained_cut))

bp <- barplot(bp_data,
              beside = TRUE,
              legend.text = FALSE,
              main = paste0("Direction by confirmation \nwith another detection before ",
                             round(flyby_window_min, 1), " min (cutpoint)"),
              ylab = "Count of flybys")

legend("bottom",
       inset = c(0, -0.37),
       legend = rownames(bp_data),
       fill = gray.colors(nrow(bp_data)),
       xpd = TRUE,
       bty = "n")

```

Looking through the time of the detection for each fly-by, we have been curious to look at the temporal distribution of the fly-by depending the directions to see whether it was tide dependent only or also relying on the time of the day. See below.

```{r hours distribution, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}

flyby_dt %>%
  
  mutate(hour_bin = hour(visitStart)) %>%
  count(hour_bin, direction, retained_cut) %>%
  
  ggplot(aes(x = hour_bin, y = n, fill = direction, pattern = retained_cut)) +
  
  geom_col_pattern(
    width = 0.9, color = "black", linewidth = 0.3,
    pattern_fill = "white", pattern_colour = "black",
    pattern_density = 0.05, pattern_spacing = 0.02, pattern_angle = 45) +
  
  scale_fill_manual(
    values = c("From mudflat (rising tide)" = "white",
               "To mudflat (falling tide)"  = "grey70"),
    name = "Direction of the flyby",
    guide = guide_legend(override.aes = list(pattern = "none"), order = 1)) +
  
  scale_pattern_manual(
    values = c("Verified by another Motus station detection" = "stripe",
               "Inferred only from tide data" = "none"),
    name = "Classification",
    guide = guide_legend(override.aes = list(fill = "white", pattern_fill = "white"), order = 2)) +
  
  scale_x_continuous(breaks = seq(0, 24, 2), expand = expansion(mult = c(0.01, 0.01))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  
  labs(
    x = "Hour of day",
    y = "Count of flybys",
    title = "Flyby detections at Fullerton Cove",
    subtitle = "Direction inferred from rising/falling tide, sometimes verified by a\npreceding or following detection at another Motus station") +
  
  theme_minimal(base_size = 15) +
  
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(color = "grey30", size = 11, margin = margin(b = 12)),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(color = "grey20"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_text(face = "bold", size = 11),
    legend.text = element_text(size = 10),
    legend.margin = margin(t = 8),
    plot.margin = margin(10, 15, 10, 10))
```

## Fly-by & Site-use

We now bring `visit_type` (fly-by / actual site use) and the direction back together into a single table covering **all** Fullerton entrance visits, so that "Site use" sits alongside the flyby directions as one classification.

```{r fc classified, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
fc_classified <- fc_visits %>%
  left_join(
    flyby_dt %>%
      select(Band.ID, visitID, direction, side_detection, delay, retained_cut),
    by = c("Band.ID", "visitID")) 
```

**Summary of every classified event at Fullerton Entrance** - Actual site use plus each fly-by direction categories, with the number, proportion of events, number of individual birds, mean and median duration for each.

```{r fc summary table, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
fc_classified %>%
  mutate(
    direction    = replace_na(direction, "-"),
    retained_cut = replace_na(retained_cut, "-")) %>%
  group_by(visit_type, direction, retained_cut) %>%
  summarise(
    n_events             = n(),
    n_birds              = n_distinct(Band.ID),
    median_duration_min  = round(median(duration_h*60), 0),
    mean_duration_min    = round(mean(duration_h*60), 0),
    .groups = "drop") %>%
  mutate(perc_event      = round(100 * n_events / sum(n_events), 1)) %>%
  arrange(visit_type, direction, retained_cut) %>%
  gt() %>%
  tab_header(title = paste("Summary -", focal_station)) %>%
  opt_align_table_header(align = "left") %>%
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels())
```

```{r fc events by species plot, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
ggplot(fc_classified
       %>% mutate(direction = replace_na(direction, "Actual site use")), ## Be careful and know your data here
       aes(x = direction, fill = direction)) +
  geom_bar() +
  facet_wrap(~ speciesEN) +
  labs(x = "Direction", y = "Number of events",
       title = paste("Direction of detections at", focal_station)) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text = element_text(face = "bold"))
```

## Rate of use

For actual site use events, the duration a bird spent at Fullerton Entrance is simply `duration_h` of the visit itself.

For fly-bys there is no direct duration - instead we use the surrounding **tide turning point** as a proxy boundary for how long the bird likely remained on the mudflat:

-   **To mudflat (falling tide)** - from the moment the bird was detected (`visitEnd`) to the *next* tide turning point (i.e. the remainder of the current tide phase).
-   **From mudflat (rising tide)** - from the *previous* tide turning point back to the moment the bird was detected (`visitStart`) (i.e. how long the current tide phase had already run before the bird arrived for detection).

```{r fc tide helpers, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
# Nearest tide turning point before / after a given timestamp
tide_times <- sort(tide_data$tideDateTimeAus)

next_tide_turn <- function(t) {
  cand <- tide_times[tide_times > t]
  if (length(cand) == 0) return(as.POSIXct(NA))
  min(cand)
}
prev_tide_turn <- function(t) {
  cand <- tide_times[tide_times < t]
  if (length(cand) == 0) return(as.POSIXct(NA))
  max(cand)
}

# Dominant tideCategory across a time window (hourly expansion, nearest-tide join)
tide_dt <- as.data.table(tide_data)[order(tideDateTimeAus), .(tideDateTimeAus, tideCategory)]

dominant_tide <- function(w_start, w_end) {
  if (is.na(w_start) || is.na(w_end) || w_end <= w_start) return(NA_character_)
  hrs    <- seq(floor_date(w_start, "hour"), floor_date(w_end, "hour"), by = "hour")
  hrs_dt <- data.table(hour_dt = hrs)
  matched <- tide_dt[hrs_dt, on = .(tideDateTimeAus = hour_dt), roll = "nearest"]
  names(sort(table(matched$tideCategory), decreasing = TRUE))[1]
}
```

**Build the per-visit duration/tide table**.

```{r fc plot data, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
plot_data <- fc_classified %>% 
  mutate(direction = replace_na(direction, "Actual site use")) %>% ## Be careful and know your data here
  
  filter(direction %in% c("Actual site use", 
                          "To mudflat (falling tide)",
                          "From mudflat (rising tide)")) %>%
  rowwise() %>%
  mutate(
    category = if_else(direction == "Actual site use",
                       "Fullerton entrance", "Fullerton mudflat"),
    win_start = case_when(
      direction == "Actual site use"                    ~ visitStart,
      direction == "To mudflat (falling tide)"    ~ visitEnd,
      direction == "From mudflat (rising tide)"   ~ prev_tide_turn(visitStart)),
    win_end = case_when(
      direction == "Actual site use"                    ~ visitEnd,
      direction == "To mudflat (falling tide)"    ~ next_tide_turn(visitEnd),
      direction == "From mudflat (rising tide)"   ~ visitStart),
    duration_h_final = as.numeric(difftime(win_end, win_start, units = "hours")),
    tideCategory_dom  = dominant_tide(win_start, win_end),
    tideDiel = if_else(grepl("Diurnal", tideCategory_dom), "Diurnal", "Nocturnal")
  ) %>%
  ungroup() %>%
  filter(!is.na(duration_h_final), duration_h_final > 0)
```

Because "Fullerton mudflat" is an *inferred* location rather than a real receiver, both categories are expressed as a rate against the same denominator: the hours Fullerton Entrance was available (station on + bird tagged and alive), per bird and per diel period. This comes from `available_bird_recv_time`, computed in the previous chapter.

```{r fc rate use, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}
# Available hours at Fullerton Entrance, per bird / species / diel period
available_fc <- available_bird_recv_time %>%
  filter(recvDeployName == "Fullerton Entrance") %>%
  group_by(Band.ID, speciesEN, tideDiel) %>%
  summarise(available_h = sum(duration_h), .groups = "drop")

# Used hours per bird / category / diel period, then compute rate of use (%)
plot_summary <- plot_data %>%
  group_by(Band.ID, speciesEN, category, tideDiel) %>%
  summarise(used_h = sum(duration_h_final), .groups = "drop") %>%
  left_join(available_fc, by = c("Band.ID", "speciesEN", "tideDiel")) %>%
  mutate(rate_use = (used_h / available_h) * 100,
         rate_use = pmin(rate_use, 100))  # cap: a few cases exceed 100% (see note below)
```

::: blockquote-yellow
**Note:** a small number of birds show `rate_use` above 100%, since mudflat duration is *inferred* from tide turning points rather than directly observed, and can occasionally exceed the bird's directly-available Fullerton Entrance time. These values are capped at 100% rather than excluded.
:::

```{r fc rate use plot, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
ggplot(plot_summary, aes(x = category, y = rate_use, fill = tideDiel)) +
  geom_boxplot(outlier.shape = NA,
               position = position_dodge(width = 0.8, preserve = "single")) +
  geom_point(aes(shape = tideDiel),
             position = position_dodge(width = 0.8),
             alpha = 1, size = 1.5,
             show.legend = FALSE) +
  scale_shape_manual(values = c("Diurnal" = 21, "Nocturnal" = 16)) +
  facet_wrap(~ speciesEN) +
  labs(x = "Location", y = "Rate of Use (%)", fill = "Tide period",
       title = "Rate of use at Fullerton Entrance vs Mudflat") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal() +
  scale_fill_manual(values = c("Diurnal" = "white", "Nocturnal" = "darkgrey")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


ggplot(plot_summary, aes(x = category, y = rate_use, fill = tideDiel)) +
  geom_violin(outlier.shape = NA,
              trim = TRUE,
              position = position_dodge(width = 0.8, preserve = "single"),
              alpha = 0.7) +
  geom_point(aes(shape = tideDiel),
             position = position_dodge(width = 0.8),
             alpha = 1, size = 1.5,
             show.legend = FALSE) +
  scale_shape_manual(values = c("Diurnal" = 21, "Nocturnal" = 16)) +
  facet_wrap(~ speciesEN) +
  labs(x = "Location", y = "Rate of Use (%)", fill = "Tide period",
       title = "Rate of use at Fullerton Entrance vs Mudflat") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal() +
  scale_fill_manual(values = c("Diurnal" = "white", "Nocturnal" = "darkgrey")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.callout-tip collapse="true" title="See Rate of Use statistics into this table"}
```{r 4 flyby rate use table, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
# General function to provide one table with stats values
make_summary_table <- function() {
  summary_table <- plot_summary %>%
    group_by(speciesEN, category, tideDiel) %>%
    summarise(
      n_individual = n_distinct(Band.ID),
      min    = round(min(rate_use, na.rm = TRUE), 1),
      q1     = round(quantile(rate_use, 0.25, na.rm = TRUE), 1),
      median = round(median(rate_use, na.rm = TRUE), 1),
      q3     = round(quantile(rate_use, 0.75, na.rm = TRUE), 1),
      max    = round(max(rate_use, na.rm = TRUE), 1),
      mean   = round(mean(rate_use, na.rm = TRUE), 1),
      sd     = round(sd(rate_use, na.rm = TRUE), 1),
      .groups = 'drop'
    ) %>%
    arrange(speciesEN, category, tideDiel) %>%
    rename(Species = "speciesEN", Location = "category", Condition = "tideDiel")
  
  return(summary_table)
}

# Single table, all species
rate_use_summary <- make_summary_table()

rate_use_summary %>%
  gt() %>%
  tab_header(title = "Statistics - Rate of Use") %>%
  opt_align_table_header(align = "left") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(Species))) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "grey40",
      weight = px(5)),
    locations = cells_body(
      rows = which(!duplicated(rate_use_summary$Species))[-1]))
```
:::

::: {.callout-tip collapse="true" title="See raw data into this table"}
```{r table of plot summary, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}

plot_summary %>%
  DT::datatable()

```
:::
