---
title: "Explore"
format: html
---

```{r setup, include=FALSE}
library(here)
source(here::here("qmd", "chapter_1", "R", "ch1_2_explore.R"))
```

::: {.callout style="border-left: 6px solid #D99B8F; --bs-callout-border: #D99B8F; padding: 1.5rem; border-radius: 8px; margin-bottom: 1rem;"}
The following analysis are based on data downloaded on **`r format(file_date, "%d-%b-%Y")`** from Motus server.

Since 2021, **`r moni$value[moni$variable == "nb_released"]` birds** belonging to **`r length(unique(df.alltags$speciesEN))` species** have been captured, tagged and released.

**You can find the complete record of the captures [here](https://uonstaff.sharepoint.com/:x:/r/sites/StudentGroupPhD-LouiseWilliamsandMatteaTaylor/_layouts/15/Doc2.aspx?action=edit&sourcedoc=%7Ba23e66f0-c551-4ba4-acd0-de0d5e83ca42%7D&wdOrigin=TEAMS-MAGLEV.teamsSdk_ns.rwc&wdExp=TEAMS-TREATMENT&wdhostclicktime=1765667294608&web=1).**

Motus allows you to openly [explore the data](https://motus.org/dashboard/#e=profile&d=projects&s=294&f=%7B%7D) on the server.

However, it might be useful and more talkative to observe fine and direct movement events from the detections, and to breakdown the data into different groups depending particular point of views and wishes.

Below are displayed several tables to explore the data with every tagged bird uniquely identified with a \`Band.ID*\`*, also related to a *MOTUS tag ID*.
:::

## Motus table

**Find below our data summarised across species and up to date** (last detection `r max(df.alltags$dateAus)`)**.**

```{r 1 table 1 pop, message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE}
table_1_pub
```

```{r print table, message=FALSE, include=FALSE, warning=FALSE, echo=FALSE, eval=TRUE}
gtsave(
  data = table_1_pub,
  filename =  here::here("material", "table_1.png"),
  path = "outputs",
  expand = 10,
  vwidth = 1600,
  vheight = 0
)
```

Open and download the table [here](material/table_1.png){target="_blank"}.

<br> <br>

And have a hint about the **conditions of detection for each species**, in the figure below *- see further details in [Station usage - hours \> Detection conditions](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_7_station_usage_hours.html#detection-conditions)*. <br>

```{r explore 2, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE}
explore
```

<br>

## Captures

### SUMMARY

Birds are trapped with mist-nets and banded. Blood and Feathers are sampled. Then a Motus tag is glued on the back of individual. Monitored for a while afterwards so we can assure his welfare. It's finally released once we confirmed the bird is behaving correctly.

Explore the below tables to get informed on the the number of birds trapped, tagged, and monitored (detected, undetected, retagged or euthanised) since 2021.

```{r 1 data overview capt, message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE}
dt_moni
```

### DETAILS

With above a summarised table of the captures and the monitoring, find below the list of the birds that has been trapped, tagged and detected, attached with the **details of the capture procedures**.

```{r 1 detect, message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE}
dt_detect
```

## Detected

The table displays **the five last detections** (`timeAus`) for each birds.

```{r 1 table 1 indiv, message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE}
dt_indiv
```

## Undetected

```{r 1 undetect, message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE}
dt_undetect
```

## Retagged

```{r 1 retagged, message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE}
dt_retag
```

## Euthanised

```{r 1 euthanised, message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE}
dt_eutha
```
