---
title: "Movement & Time (Exploratory analysis)"
format: html
toc-location: left
---

::: {.callout style="border-left: 6px solid #D99B8F; --bs-callout-border: #D99B8F; padding: 1.5rem; border-radius: 8px; margin-bottom: 1rem;"}
In this section, we **explore *how bird detections are distributed over time*.**

A preliminary manual examination of the raw detection data indicated that most detections occurred during specific periods, particularly in the early morning and late evening.

**We therefore aim to investigate how birds move within the estuary in relation to time of day; hourly instead of day/night, light conditions, and tidal cycle**. Several environmental variables may be considered, including natural light levels during the day, represented by solar radiation, and at night, represented by moonlight. Tidal cycles obviously also influence bird movements, although it might drive shorebird movements after the circadian cycle.

*An important limitation remains: most birds were detected during the day and around high tide. This pattern must partly reflect the placement of the Motus stations, which were located in areas where birds were known to occur. Consequently, it is difficult to determine whether the observed detection pattern reflects genuine habitat use or is partly a result of station placement.*
:::

## Descriptive analysis

```{r set-up 1, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, include = FALSE, results = 'hide'}
# Source Station usage - hours (OBP method) script into this environment
source(knitr::purl(here::here("qmd", "chapter_1", "ch1_12.qmd"),
                   output = tempfile(fileext = ".R"),
                   quiet = FALSE))
```

```{r local packages, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
library(mixtools)
library(lubridate)
library(gt)
library(lubridate)
library(ggpattern)
```

```{r initialise-htmlwidgets, include=FALSE}
library(DT)
library(htmltools)

# Initialise the DT JavaScript dependencies once
htmltools::tagList(
  DT::datatable(
    data.frame(initialise = "DT"),
    options = list(dom = "t")
  )
)
```

```{r station visits, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}

# Real stations, alphabetical
real_stations <- visit_duration %>%
  distinct(recvDeployName) %>%
  arrange(recvDeployName) %>%
  pull(recvDeployName)

# "All Stations" pinned as the first tab, everything else after
stations <- c("All Stations", real_stations)

# Per-station subsets — "All Stations" just uses the full data, unfiltered
visit_by_station <- setNames(
  lapply(stations, function(s) {
    if (s == "All Stations") visit_duration else filter(visit_duration, recvDeployName == s)
  }),
  stations
)

raw_by_station <- setNames(
  lapply(stations, function(s) {
    if (s == "All Stations") data_all else filter(data_all, recvDeployName == s)
  }),
  stations
)
```

```{r station visits check, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results = 'hide'}

# ============================================================
# Sunrise/sunset lookup — 24 hours
# ============================================================

sun_lookup <- data_all %>%
  mutate(date = as_date(dateAus)) %>%
  distinct(date, sunriseNewc, sunsetNewc) %>%
  mutate(
    sunrise_hour = lubridate::hour(sunriseNewc) + lubridate::minute(sunriseNewc) / 60 + lubridate::second(sunriseNewc) / 3600,
    sunset_hour  = lubridate::hour(sunsetNewc)  + lubridate::minute(sunsetNewc)  / 60 + lubridate::second(sunsetNewc)  / 3600
  )

add_sun_reference <- function(p, dates) {
  sun_sub <- sun_lookup %>% filter(date %in% unique(as_date(dates)))
  if (nrow(sun_sub) == 0) return(p)

  sr <- quantile(sun_sub$sunrise_hour, c(0.1, 0.5, 0.9), na.rm = TRUE)
  ss <- quantile(sun_sub$sunset_hour,  c(0.1, 0.5, 0.9), na.rm = TRUE)

  p +
    annotate("rect", xmin = sr[1], xmax = sr[3], ymin = -Inf, ymax = Inf,
             fill = "orange", alpha = 0.15) +
    geom_vline(xintercept = sr[2], color = "orange", linewidth = 0.8, alpha = 0.8) +
    annotate("rect", xmin = ss[1], xmax = ss[3], ymin = -Inf, ymax = Inf,
             fill = "purple", alpha = 0.15) +
    geom_vline(xintercept = ss[2], color = "purple", linewidth = 0.8, alpha = 0.8)
}

add_sun_reference_h <- function(p, dates) {
  sun_sub <- sun_lookup %>% filter(date %in% unique(as_date(dates)))
  if (nrow(sun_sub) == 0) return(p)

  sr <- quantile(sun_sub$sunrise_hour, c(0.1, 0.5, 0.9), na.rm = TRUE)
  ss <- quantile(sun_sub$sunset_hour,  c(0.1, 0.5, 0.9), na.rm = TRUE)

  to_pos <- function(x) x + 1

  p +
    annotate("rect", ymin = to_pos(sr[1]), ymax = to_pos(sr[3]), xmin = -Inf, xmax = Inf,
             fill = "orange", alpha = 0.15) +
    geom_hline(yintercept = to_pos(sr[2]), color = "orange", linewidth = 0.8, alpha = 0.8) +
    annotate("rect", ymin = to_pos(ss[1]), ymax = to_pos(ss[3]), xmin = -Inf, xmax = Inf,
             fill = "purple", alpha = 0.15) +
    geom_hline(yintercept = to_pos(ss[2]), color = "purple", linewidth = 0.8, alpha = 0.8)
}


# ============================================================
# 48-hour ("double day") sunrise / sunset / night overlay
# ============================================================
add_sun_reference_h_48 <- function(p, dates) {
  sun_sub <- sun_lookup %>% filter(date %in% unique(as_date(dates)))
  if (nrow(sun_sub) == 0) return(p)

  sr <- quantile(sun_sub$sunrise_hour, c(0.1, 0.5, 0.9), na.rm = TRUE)
  ss <- quantile(sun_sub$sunset_hour,  c(0.1, 0.5, 0.9), na.rm = TRUE)

  to_pos <- function(x) x + 1   # hour -> discrete y position (hour 0 = position 1)

  sr_lo <- to_pos(sr[1]); sr_mid <- to_pos(sr[2]); sr_hi <- to_pos(sr[3])
  ss_lo <- to_pos(ss[1]); ss_mid <- to_pos(ss[2]); ss_hi <- to_pos(ss[3])

  p +
    # --- NIGHT (light grey): everything outside the sunrise-sunset window,
    #     across both copies of the cycle — this is what makes the "night"
    #     read as one continuous block in the middle of the 48h view ---
    annotate("rect", ymin = -Inf,        ymax = sr_mid,        xmin = -Inf, xmax = Inf, fill = "grey60", alpha = 0.25) +
    annotate("rect", ymin = ss_mid,      ymax = sr_mid + 24,   xmin = -Inf, xmax = Inf, fill = "grey60", alpha = 0.25) +
    annotate("rect", ymin = ss_mid + 24, ymax = Inf,           xmin = -Inf, xmax = Inf, fill = "grey60", alpha = 0.25) +
    # --- DAWN (orange), duplicated ---
    annotate("rect", ymin = sr_lo,      ymax = sr_hi,      xmin = -Inf, xmax = Inf, fill = "orange", alpha = 0.15) +
    annotate("rect", ymin = sr_lo + 24, ymax = sr_hi + 24, xmin = -Inf, xmax = Inf, fill = "orange", alpha = 0.15) +
    geom_hline(yintercept = sr_mid,      color = "orange", linewidth = 0.8, alpha = 0.8) +
    geom_hline(yintercept = sr_mid + 24, color = "orange", linewidth = 0.8, alpha = 0.8) +
    # --- DUSK (purple), duplicated ---
    annotate("rect", ymin = ss_lo,      ymax = ss_hi,      xmin = -Inf, xmax = Inf, fill = "purple", alpha = 0.15) +
    annotate("rect", ymin = ss_lo + 24, ymax = ss_hi + 24, xmin = -Inf, xmax = Inf, fill = "purple", alpha = 0.15) +
    geom_hline(yintercept = ss_mid,      color = "purple", linewidth = 0.8, alpha = 0.8) +
    geom_hline(yintercept = ss_mid + 24, color = "purple", linewidth = 0.8, alpha = 0.8)
}

# ============================================================
# Reusable plot builders — unchanged
# ============================================================

plot_hour_count <- function(data, time_col, title,
                             x_lab = "Hour of Day", y_lab = "Number of Visits") {
  p <- data %>%
    mutate(hour = lubridate::hour(.data[[time_col]])) %>%
    count(hour) %>%
    ggplot(aes(x = hour, y = n)) +
    geom_col(fill = "steelblue") +
    scale_x_continuous(breaks = 0:23) +
    labs(title = title, x = x_lab, y = y_lab) +
    theme_minimal()

  add_sun_reference(p, data[[time_col]])
}

plot_hour_duration <- function(data, time_col, duration_col, title,
                                x_lab = "Hour of Day", y_lab = "Total Duration (hours)") {
  p <- data %>%
    mutate(hour = lubridate::hour(.data[[time_col]])) %>%
    group_by(hour) %>%
    summarise(total_duration_h = sum(.data[[duration_col]], na.rm = TRUE), .groups = "drop") %>%
    ggplot(aes(x = hour, y = total_duration_h)) +
    geom_col(fill = "steelblue") +
    scale_x_continuous(breaks = 0:23) +
    labs(title = title, x = x_lab, y = y_lab) +
    theme_minimal()

  add_sun_reference(p, data[[time_col]])
}

plot_hour_duration_burst <- function(data, time_col, title,
                                      x_lab = "Hour of Day", y_lab = "Total Duration (hours)") {
  p <- data %>%
    mutate(
      burst_inter_s = ifelse(tagModel == "NTQB2-6-2", 7.1, 13.1),
      hour = lubridate::hour(.data[[time_col]])
    ) %>%
    group_by(hour) %>%
    summarise(total_duration_h = sum(burst_inter_s, na.rm = TRUE) / 3600, .groups = "drop") %>%
    ggplot(aes(x = hour, y = total_duration_h)) +
    geom_col(fill = "steelblue") +
    scale_x_continuous(breaks = 0:23) +
    labs(title = title, x = x_lab, y = y_lab) +
    theme_minimal()

  add_sun_reference(p, data[[time_col]])
}

plot_hour_box_duration <- function(data, time_col, duration_col, title,
                                    x_lab = "Visit Duration (hours)",
                                    y_lab = "Hour of Day",
                                    show_points = TRUE) {
  df <- data %>%
    mutate(hour = lubridate::hour(.data[[time_col]]),
           hour_f = factor(hour, levels = 0:23))

  p <- ggplot(df, aes(x = .data[[duration_col]], y = hour_f)) +
    geom_boxplot(fill = "steelblue", alpha = 0.5, outlier.shape = NA, width = 0.6)

  if (show_points) {
    p <- p +
      geom_jitter(aes(color = Band.ID), height = 0.15, width = 0,
                  size = 0.8, alpha = 0.4, show.legend = FALSE)
  }

  p <- p +
    scale_y_discrete(limits = as.character(0:23), drop = FALSE) +   
    labs(title = title, x = x_lab, y = y_lab) +
    theme_minimal()

  add_sun_reference_h(p, data[[time_col]])
}


# 
# plot_hour_count_duration_pyramid <- function(data, time_col, duration_col, title,
#                                               y_lab       = "Hour of Day",
#                                               label_left  = "Number of Passage (counts)",
#                                               label_right = "Passage Duration (hours)",
#                                               show_points = TRUE) {
# 
#   bar_color <- "steelblue"
#   box_color <- "black"
# 
#   df <- data %>%
#     mutate(hour = lubridate::hour(.data[[time_col]]),
#            hour_f = factor(hour, levels = 0:23))   # keep all 24 hours, even if empty
# 
#   # Counts per hour — explicitly complete 0:23 so empty hours show as 0, not gaps
#   count_df <- df %>%
#     count(hour, name = "n") %>%
#     right_join(tibble(hour = 0:23), by = "hour") %>%
#     mutate(n = tidyr::replace_na(n, 0),
#            hour_f = factor(hour, levels = 0:23)) %>%
#     arrange(hour)
# 
#   # Scale factor maps duration onto the same numeric range as counts, so both
#   # sides of the pyramid are visually comparable; breaks/labels below undo it
#   max_count    <- max(count_df$n, na.rm = TRUE)
#   max_duration <- suppressWarnings(max(df[[duration_col]], na.rm = TRUE))
#   if (!is.finite(max_duration) || max_duration == 0) max_duration <- 1
#   scale_factor <- if (max_count == 0) 1 else max_count / max_duration
# 
#   # Build symmetric breaks: negative side in raw count units, positive side in
#   # raw duration units (scaled back to plot coordinates)
#   count_breaks    <- scales::extended_breaks()(c(0, max_count))
#   count_breaks    <- count_breaks[count_breaks >= 0]
#   duration_breaks <- scales::extended_breaks()(c(0, max_duration))
#   duration_breaks <- duration_breaks[duration_breaks >= 0]
# 
#   x_breaks <- c(-rev(count_breaks), duration_breaks * scale_factor)
#   x_labels <- c(rev(count_breaks), duration_breaks)
# 
#   p <- ggplot() +
#     # LEFT side — bars, counts (negative x = leftward)
#     geom_col(data = count_df, aes(x = -n, y = hour_f),
#              fill = bar_color, alpha = 0.65, width = 0.8) +
#     # RIGHT side — boxplots, duration (positive x = rightward)
#     geom_boxplot(data = df,
#                  aes(x = .data[[duration_col]] * scale_factor, y = hour_f, group = hour_f),
#                  fill = box_color, color = box_color, alpha = 0.2,
#                  outlier.shape = NA, width = 0.6)
# 
#   if (show_points) {
#     p <- p +
#       geom_jitter(data = df,
#                   aes(x = .data[[duration_col]] * scale_factor, y = hour_f),
#                   height = 0.15, width = 0, size = 0.7, alpha = 0.15, color = box_color)
#   }
# 
#   p <- p +
#     geom_vline(xintercept = 0, color = "grey30", linewidth = 0.6) +
#     scale_x_continuous(breaks = x_breaks, labels = x_labels) +
#     scale_y_discrete(limits = as.character(0:23), drop = FALSE) +
#     labs(title = title, x = NULL, y = y_lab) +
#     theme_minimal() +
#     theme(axis.text.x = element_text(color = "grey20")) +
#     # Color-coded headers so it's unambiguous which side reads which unit
#     annotate("text", x = -max(count_breaks) * 0.5, y = -1,
#              label = label_left, color = bar_color,
#              fontface = "bold", hjust = 0.5, size = 3.5) +
#     annotate("text", x = max(duration_breaks) * scale_factor * 0.5, y = -1,
#              label = label_right, color = box_color,
#              fontface = "bold", hjust = 0.5, size = 3.5) +
#     coord_cartesian(ylim = c(1, 24), clip = "off") +
#     theme(plot.margin = margin(t = 25, r = 10, b = 25, l = 10))
# 
#   add_sun_reference_h(p, data[[time_col]])   # y = hour here, so the horizontal-band version applies
# }
# 
# plot_hour_count_duration_pyramid_species <- function(data, time_col, duration_col, title,
#                                                        y_lab       = "Hour of Day",
#                                                        label_left  = " ",
#                                                        label_right = " ",
#                                                        show_points = TRUE) {
# 
#   bar_color <- "steelblue"
#   box_color <- "black"
# 
#   df <- data %>%
#     mutate(hour = lubridate::hour(.data[[time_col]]),
#            hour_f = factor(hour, levels = 0:23))
# 
#   # Counts per hour PER SPECIES — explicitly complete 0:23 x species so empty
#   # hours show as 0, not gaps, for every species/hour combination
#   species_list <- sort(unique(df$speciesEN))
# 
#   count_df <- df %>%
#     count(speciesEN, hour, name = "n") %>%
#     right_join(
#       tidyr::expand_grid(speciesEN = species_list, hour = 0:23),
#       by = c("speciesEN", "hour")
#     ) %>%
#     mutate(n = tidyr::replace_na(n, 0),
#            hour_f = factor(hour, levels = 0:23)) %>%
#     arrange(speciesEN, hour)
# 
#   # Scale factor computed GLOBALLY (across all species) so facets stay
#   # comparable to one another rather than each being independently rescaled
#   max_count    <- max(count_df$n, na.rm = TRUE)
#   max_duration <- suppressWarnings(max(df[[duration_col]], na.rm = TRUE))
#   if (!is.finite(max_duration) || max_duration == 0) max_duration <- 1
#   scale_factor <- if (max_count == 0) 1 else max_count / max_duration
# 
#   count_breaks    <- scales::extended_breaks()(c(0, max_count))
#   count_breaks    <- count_breaks[count_breaks >= 0]
#   duration_breaks <- scales::extended_breaks()(c(0, max_duration))
#   duration_breaks <- duration_breaks[duration_breaks >= 0]
# 
#   x_breaks <- c(-rev(count_breaks), duration_breaks * scale_factor)
#   x_labels <- c(rev(count_breaks), duration_breaks)
# 
#   p <- ggplot() +
#     geom_col(data = count_df, aes(x = -n, y = hour_f),
#              fill = bar_color, alpha = 0.65, width = 0.8) +
#     geom_boxplot(data = df,
#                  aes(x = .data[[duration_col]] * scale_factor, y = hour_f, group = hour_f),
#                  fill = box_color, color = box_color, alpha = 0.2,
#                  outlier.shape = NA, width = 0.6)
# 
#   if (show_points) {
#     p <- p +
#       geom_jitter(data = df,
#                   aes(x = .data[[duration_col]] * scale_factor, y = hour_f),
#                   height = 0.15, width = 0, size = 0.6, alpha = 0.12, color = box_color)
#   }
# 
#   p <- p +
#     geom_vline(xintercept = 0, color = "grey30", linewidth = 0.6) +
#     scale_x_continuous(breaks = x_breaks, labels = x_labels) +
#     scale_y_discrete(limits = as.character(0:23), drop = FALSE) +
#     facet_wrap(~ speciesEN, ncol = 2) +
#     labs(title = title, x = NULL, y = y_lab) +
#     theme_minimal() +
#     theme(
#       axis.text.x = element_text(color = "grey20", size = 8),
#       axis.text.y = element_text(size = 7),
#       strip.text  = element_text(face = "bold", size = 9)
#     ) +
#     annotate("text", x = -max(count_breaks) * 0.5, y = -1,
#              label = label_left, color = bar_color,
#              fontface = "bold", hjust = 0.5, size = 3) +
#     annotate("text", x = max(duration_breaks) * scale_factor * 0.5, y = -1,
#              label = label_right, color = box_color,
#              fontface = "bold", hjust = 0.5, size = 3) +
#     coord_cartesian(ylim = c(1, 24), clip = "off") +
#     theme(plot.margin = margin(t = 25, r = 10, b = 25, l = 10))
# 
#   add_sun_reference_h(p, data[[time_col]])
# }
# 
# plot_hour_pyramid_detailed <- function(data, time_col, duration_col, title,
#                                         y_lab       = "Hour of Day",
#                                         label_left  = "Number of Passages (per Band.ID, counts)",
#                                         label_right = "Passage Duration (per Band.ID, hours)") {
# 
#   bar_color <- "steelblue"
#   box_color <- "black"
# 
#   df <- data %>%
#     mutate(hour = lubridate::hour(.data[[time_col]]),
#            hour_f = factor(hour, levels = 0:23))
# 
#   # LEFT: number of passages per Band.ID, per hour (boxplot distribution across birds)
#   count_band_df <- df %>%
#     count(hour, Band.ID, name = "n") %>%
#     mutate(hour_f = factor(hour, levels = 0:23))
# 
#   max_left  <- max(count_band_df$n, na.rm = TRUE)
#   max_right <- suppressWarnings(max(df[[duration_col]], na.rm = TRUE))
#   if (!is.finite(max_left)  || max_left  == 0) max_left  <- 1
#   if (!is.finite(max_right) || max_right == 0) max_right <- 1
#   scale_factor <- max_left / max_right
# 
#   left_breaks  <- scales::extended_breaks()(c(0, max_left))
#   left_breaks  <- left_breaks[left_breaks >= 0]
#   right_breaks <- scales::extended_breaks()(c(0, max_right))
#   right_breaks <- right_breaks[right_breaks >= 0]
# 
#   x_breaks <- c(-rev(left_breaks), right_breaks * scale_factor)
#   x_labels <- c(rev(left_breaks), right_breaks)
# 
#   p <- ggplot() +
#     # LEFT — boxplot of passage counts per Band.ID
#     geom_boxplot(data = count_band_df,
#                  aes(x = -n, y = hour_f, group = hour_f),
#                  fill = bar_color, color = bar_color, alpha = 0.25,
#                  outlier.shape = NA, width = 0.6) +
#     geom_jitter(data = count_band_df,
#                 aes(x = -n, y = hour_f),
#                 height = 0.15, width = 0, size = 0.6, alpha = 0.15, color = bar_color) +
#     # RIGHT — boxplot of duration per Band.ID (unchanged from before)
#     geom_boxplot(data = df,
#                  aes(x = .data[[duration_col]] * scale_factor, y = hour_f, group = hour_f),
#                  fill = box_color, color = box_color, alpha = 0.2,
#                  outlier.shape = NA, width = 0.6) +
#     geom_jitter(data = df,
#                 aes(x = .data[[duration_col]] * scale_factor, y = hour_f),
#                 height = 0.15, width = 0, size = 0.6, alpha = 0.15, color = box_color) +
#     geom_vline(xintercept = 0, color = "grey30", linewidth = 0.6) +
#     scale_x_continuous(breaks = x_breaks, labels = x_labels) +
#     scale_y_discrete(limits = as.character(0:23), drop = FALSE) +
#     labs(title = title, x = NULL, y = y_lab) +
#     theme_minimal() +
#     theme(axis.text.x = element_text(color = "grey20")) +
#     annotate("text", x = -max(left_breaks) * 0.5, y = -1,
#              label = label_left, color = bar_color,
#              fontface = "bold", hjust = 0.5, size = 3.5) +
#     annotate("text", x = max(right_breaks) * scale_factor * 0.5, y = -1,
#              label = label_right, color = box_color,
#              fontface = "bold", hjust = 0.5, size = 3.5) +
#     coord_cartesian(ylim = c(1, 24), clip = "off") +
#     theme(plot.margin = margin(t = 25, r = 10, b = 25, l = 10))
# 
#   add_sun_reference_h(p, data[[time_col]])
# }

# plot_hour_pyramid_simple <- function(data, time_col, duration_col, title,
#                                       y_lab       = "Hour of Day",
#                                       label_left  = " ",
#                                       label_right = " ") {
# 
#   bar_color_left  <- "steelblue"
#   bar_color_right <- "black"
# 
#   df <- data %>%
#     mutate(hour = lubridate::hour(.data[[time_col]]),
#            hour_f = factor(hour, levels = 0:23))
# 
#   count_df <- df %>%
#     count(hour, name = "n") %>%
#     right_join(tibble(hour = 0:23), by = "hour") %>%
#     mutate(n = tidyr::replace_na(n, 0), hour_f = factor(hour, levels = 0:23)) %>%
#     arrange(hour)
# 
#   duration_df <- df %>%
#     group_by(hour) %>%
#     summarise(total_duration = sum(.data[[duration_col]], na.rm = TRUE), .groups = "drop") %>%
#     right_join(tibble(hour = 0:23), by = "hour") %>%
#     mutate(total_duration = tidyr::replace_na(total_duration, 0),
#            hour_f = factor(hour, levels = 0:23)) %>%
#     arrange(hour)
# 
#   max_count    <- max(count_df$n, na.rm = TRUE)
#   max_duration <- max(duration_df$total_duration, na.rm = TRUE)
#   if (!is.finite(max_count)    || max_count    == 0) max_count    <- 1
#   if (!is.finite(max_duration) || max_duration == 0) max_duration <- 1
#   scale_factor <- max_count / max_duration
# 
#   count_breaks    <- scales::extended_breaks()(c(0, max_count))
#   count_breaks    <- count_breaks[count_breaks >= 0]
#   duration_breaks <- scales::extended_breaks()(c(0, max_duration))
#   duration_breaks <- duration_breaks[duration_breaks >= 0]
# 
#   x_breaks <- c(-rev(count_breaks), duration_breaks * scale_factor)
#   x_labels <- c(rev(count_breaks), duration_breaks)
# 
#   p <- ggplot() +
#     geom_col(data = count_df, aes(x = -n, y = hour_f),
#              fill = bar_color_left, alpha = 0.65, width = 0.8) +
#     geom_col(data = duration_df, aes(x = total_duration * scale_factor, y = hour_f),
#              fill = bar_color_right, alpha = 0.65, width = 0.8) +
#     geom_vline(xintercept = 0, color = "grey30", linewidth = 0.6) +
#     scale_x_continuous(breaks = x_breaks, labels = x_labels) +
#     scale_y_discrete(limits = as.character(0:23), drop = FALSE) +
#     labs(title = title, x = NULL, y = y_lab) +
#     theme_minimal() +
#     theme(axis.text.x = element_text(color = "grey20")) +
#     annotate("text", x = -max(count_breaks) * 0.5, y = -1,
#              label = label_left, color = bar_color_left,
#              fontface = "bold", hjust = 0.5, size = 3.5) +
#     annotate("text", x = max(duration_breaks) * scale_factor * 0.5, y = -1,
#              label = label_right, color = bar_color_right,
#              fontface = "bold", hjust = 0.5, size = 3.5) +
#     coord_cartesian(ylim = c(1, 24), clip = "off") +
#     theme(plot.margin = margin(t = 25, r = 10, b = 25, l = 10))
# 
#   add_sun_reference_h(p, data[[time_col]])
# }

# plot_hour_pyramid_detailed <- function(data, time_col, duration_col, title,
#                                         y_lab       = "Hour of Day",
#                                         label_left  = "Number of Passages (Total)",
#                                         label_right = "Passage Duration (per Band.ID, hours)") {
# 
#   bar_color <- "steelblue"
#   box_color <- "black"
# 
#   df <- data %>%
#     mutate(hour = lubridate::hour(.data[[time_col]]),
#            hour_f = factor(hour, levels = 0:23))
# 
#   # LEFT: total number of passages per hour (simple bar, not per-Band.ID)
#   count_df <- df %>%
#     count(hour, name = "n") %>%
#     right_join(tibble(hour = 0:23), by = "hour") %>%
#     mutate(n = tidyr::replace_na(n, 0),
#            hour_f = factor(hour, levels = 0:23)) %>%
#     arrange(hour)
# 
#   max_left  <- max(count_df$n, na.rm = TRUE)
#   max_right <- suppressWarnings(max(df[[duration_col]], na.rm = TRUE))
#   if (!is.finite(max_left)  || max_left  == 0) max_left  <- 1
#   if (!is.finite(max_right) || max_right == 0) max_right <- 1
#   scale_factor <- max_left / max_right
# 
#   left_breaks  <- scales::extended_breaks()(c(0, max_left))
#   left_breaks  <- left_breaks[left_breaks >= 0]
#   right_breaks <- scales::extended_breaks()(c(0, max_right))
#   right_breaks <- right_breaks[right_breaks >= 0]
# 
#   x_breaks <- c(-rev(left_breaks), right_breaks * scale_factor)
#   x_labels <- c(rev(left_breaks), right_breaks)
# 
#   p <- ggplot() +
#     # LEFT — barplot of total passage counts
#     geom_col(data = count_df,
#              aes(x = -n, y = hour_f),
#              fill = bar_color, alpha = 0.65, width = 0.8) +
#     # RIGHT — boxplot of duration per Band.ID (unchanged)
#     geom_boxplot(data = df,
#                  aes(x = .data[[duration_col]] * scale_factor, y = hour_f, group = hour_f),
#                  fill = box_color, color = box_color, alpha = 0.2,
#                  outlier.shape = NA, width = 0.6) +
#     geom_jitter(data = df,
#                 aes(x = .data[[duration_col]] * scale_factor, y = hour_f),
#                 height = 0.15, width = 0, size = 0.6, alpha = 0.15, color = box_color) +
#     geom_vline(xintercept = 0, color = "grey30", linewidth = 0.6) +
#     scale_x_continuous(breaks = x_breaks, labels = x_labels) +
#     scale_y_discrete(limits = as.character(0:23), drop = FALSE) +
#     labs(title = title, x = NULL, y = y_lab) +
#     theme_minimal() +
#     theme(axis.text.x = element_text(color = "grey20")) +
#     annotate("text", x = -max(left_breaks) * 0.5, y = -1,
#              label = label_left, color = bar_color,
#              fontface = "bold", hjust = 0.5, size = 3.5) +
#     annotate("text", x = max(right_breaks) * scale_factor * 0.5, y = -1,
#              label = label_right, color = box_color,
#              fontface = "bold", hjust = 0.5, size = 3.5) +
#     coord_cartesian(ylim = c(1, 24), clip = "off") +
#     theme(plot.margin = margin(t = 25, r = 10, b = 25, l = 10))
# 
#   add_sun_reference_h(p, data[[time_col]])
# }
# 
# plot_hour_pyramid_detailed_species <- function(data, time_col, duration_col, title,
#                                                 y_lab       = "Hour of Day",
#                                                 label_left  = " ",
#                                                 label_right = " ") {
# 
#   bar_color <- "steelblue"
#   box_color <- "black"
# 
#   df <- data %>%
#     mutate(hour = lubridate::hour(.data[[time_col]]),
#            hour_f = factor(hour, levels = 0:23))
# 
#   species_list <- sort(unique(df$speciesEN))
# 
#   # LEFT: total passage counts per hour, per species (simple bar)
#   count_df <- df %>%
#     count(speciesEN, hour, name = "n") %>%
#     right_join(
#       tidyr::expand_grid(speciesEN = species_list, hour = 0:23),
#       by = c("speciesEN", "hour")
#     ) %>%
#     mutate(n = tidyr::replace_na(n, 0),
#            hour_f = factor(hour, levels = 0:23)) %>%
#     arrange(speciesEN, hour)
# 
#   # Global scaling so facets stay comparable
#   max_left  <- max(count_df$n, na.rm = TRUE)
#   max_right <- suppressWarnings(max(df[[duration_col]], na.rm = TRUE))
#   if (!is.finite(max_left)  || max_left  == 0) max_left  <- 1
#   if (!is.finite(max_right) || max_right == 0) max_right <- 1
#   scale_factor <- max_left / max_right
# 
#   left_breaks  <- scales::extended_breaks()(c(0, max_left))
#   left_breaks  <- left_breaks[left_breaks >= 0]
#   right_breaks <- scales::extended_breaks()(c(0, max_right))
#   right_breaks <- right_breaks[right_breaks >= 0]
# 
#   x_breaks <- c(-rev(left_breaks), right_breaks * scale_factor)
#   x_labels <- c(rev(left_breaks), right_breaks)
# 
#   p <- ggplot() +
#     # LEFT — barplot of total passage counts, per species
#     geom_col(data = count_df,
#              aes(x = -n, y = hour_f),
#              fill = bar_color, alpha = 0.65, width = 0.8) +
#     # RIGHT — boxplot of duration per Band.ID, per species (unchanged)
#     geom_boxplot(data = df,
#                  aes(x = .data[[duration_col]] * scale_factor, y = hour_f, group = hour_f),
#                  fill = box_color, color = box_color, alpha = 0.2,
#                  outlier.shape = NA, width = 0.6) +
#     geom_jitter(data = df,
#                 aes(x = .data[[duration_col]] * scale_factor, y = hour_f),
#                 height = 0.15, width = 0, size = 0.5, alpha = 0.12, color = box_color) +
#     geom_vline(xintercept = 0, color = "grey30", linewidth = 0.6) +
#     scale_x_continuous(breaks = x_breaks, labels = x_labels) +
#     scale_y_discrete(limits = as.character(0:23), drop = FALSE) +
#     facet_wrap(~ speciesEN, ncol = 2) +
#     labs(title = title, x = NULL, y = y_lab) +
#     theme_minimal() +
#     theme(
#       axis.text.x = element_text(color = "grey20", size = 8),
#       axis.text.y = element_text(size = 7),
#       strip.text  = element_text(face = "bold", size = 9)
#     ) +
#     annotate("text", x = -max(left_breaks) * 0.5, y = -1,
#              label = label_left, color = bar_color,
#              fontface = "bold", hjust = 0.5, size = 3) +
#     annotate("text", x = max(right_breaks) * scale_factor * 0.5, y = -1,
#              label = label_right, color = box_color,
#              fontface = "bold", hjust = 0.5, size = 3) +
#     coord_cartesian(ylim = c(1, 24), clip = "off") +
#     theme(plot.margin = margin(t = 25, r = 10, b = 25, l = 10))
# 
#   add_sun_reference_h(p, data[[time_col]])
# }



plot_hour_pyramid_detailed_species <- function(data, time_col, duration_col, title,
                                                y_lab       = "Hour of Day",
                                                label_left  = " ",
                                                label_right = " ") {

  bar_color <- "steelblue"
  box_color <- "black"

  df <- data %>%
    mutate(hour = lubridate::hour(.data[[time_col]]),
           hour_f = factor(hour, levels = 0:23))

  # LEFT: passages per Band.ID per hour, WITHIN each species
  count_band_df <- df %>%
    count(speciesEN, hour, Band.ID, name = "n") %>%
    mutate(hour_f = factor(hour, levels = 0:23))

  max_left  <- max(count_band_df$n, na.rm = TRUE)
  max_right <- suppressWarnings(max(df[[duration_col]], na.rm = TRUE))
  if (!is.finite(max_left)  || max_left  == 0) max_left  <- 1
  if (!is.finite(max_right) || max_right == 0) max_right <- 1
  scale_factor <- max_left / max_right

  left_breaks  <- scales::extended_breaks()(c(0, max_left))
  left_breaks  <- left_breaks[left_breaks >= 0]
  right_breaks <- scales::extended_breaks()(c(0, max_right))
  right_breaks <- right_breaks[right_breaks >= 0]

  x_breaks <- c(-rev(left_breaks), right_breaks * scale_factor)
  x_labels <- c(rev(left_breaks), right_breaks)

  p <- ggplot() +
    geom_boxplot(data = count_band_df,
                 aes(x = -n, y = hour_f, group = hour_f),
                 fill = bar_color, color = bar_color, alpha = 0.25,
                 outlier.shape = NA, width = 0.6) +
    geom_jitter(data = count_band_df,
                aes(x = -n, y = hour_f),
                height = 0.15, width = 0, size = 0.5, alpha = 0.12, color = bar_color) +
    geom_boxplot(data = df,
                 aes(x = .data[[duration_col]] * scale_factor, y = hour_f, group = hour_f),
                 fill = box_color, color = box_color, alpha = 0.2,
                 outlier.shape = NA, width = 0.6) +
    geom_jitter(data = df,
                aes(x = .data[[duration_col]] * scale_factor, y = hour_f),
                height = 0.15, width = 0, size = 0.5, alpha = 0.12, color = box_color) +
    geom_vline(xintercept = 0, color = "grey30", linewidth = 0.6) +
    scale_x_continuous(breaks = x_breaks, labels = x_labels) +
    scale_y_discrete(limits = as.character(0:23), drop = FALSE) +
    facet_wrap(~ speciesEN, ncol = 2) +
    labs(title = title, x = NULL, y = y_lab) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(color = "grey20", size = 8),
      axis.text.y = element_text(size = 7),
      strip.text  = element_text(face = "bold", size = 9)
    ) +
    annotate("text", x = -max(left_breaks) * 0.5, y = -1,
             label = label_left, color = bar_color,
             fontface = "bold", hjust = 0.5, size = 3) +
    annotate("text", x = max(right_breaks) * scale_factor * 0.5, y = -1,
             label = label_right, color = box_color,
             fontface = "bold", hjust = 0.5, size = 3) +
    coord_cartesian(ylim = c(1, 24), clip = "off") +
    theme(plot.margin = margin(t = 25, r = 10, b = 25, l = 10))

  add_sun_reference_h(p, data[[time_col]])
}

# plot_hour_pyramid_simple_species <- function(data, time_col, duration_col, title,
#                                               y_lab       = "Hour of Day",
#                                               label_left  = " ",
#                                               label_right = " ") {
# 
#   bar_color_left  <- "steelblue"
#   bar_color_right <- "black"
# 
#   df <- data %>%
#     mutate(hour = lubridate::hour(.data[[time_col]]),
#            hour_f = factor(hour, levels = 0:23))
# 
#   species_list <- sort(unique(df$speciesEN))
# 
#   count_df <- df %>%
#     count(speciesEN, hour, name = "n") %>%
#     right_join(
#       tidyr::expand_grid(speciesEN = species_list, hour = 0:23),
#       by = c("speciesEN", "hour")
#     ) %>%
#     mutate(n = tidyr::replace_na(n, 0), hour_f = factor(hour, levels = 0:23)) %>%
#     arrange(speciesEN, hour)
# 
#   duration_df <- df %>%
#     group_by(speciesEN, hour) %>%
#     summarise(total_duration = sum(.data[[duration_col]], na.rm = TRUE), .groups = "drop") %>%
#     right_join(
#       tidyr::expand_grid(speciesEN = species_list, hour = 0:23),
#       by = c("speciesEN", "hour")
#     ) %>%
#     mutate(total_duration = tidyr::replace_na(total_duration, 0),
#            hour_f = factor(hour, levels = 0:23)) %>%
#     arrange(speciesEN, hour)
# 
#   max_count    <- max(count_df$n, na.rm = TRUE)
#   max_duration <- max(duration_df$total_duration, na.rm = TRUE)
#   if (!is.finite(max_count)    || max_count    == 0) max_count    <- 1
#   if (!is.finite(max_duration) || max_duration == 0) max_duration <- 1
#   scale_factor <- max_count / max_duration
# 
#   count_breaks    <- scales::extended_breaks()(c(0, max_count))
#   count_breaks    <- count_breaks[count_breaks >= 0]
#   duration_breaks <- scales::extended_breaks()(c(0, max_duration))
#   duration_breaks <- duration_breaks[duration_breaks >= 0]
# 
#   x_breaks <- c(-rev(count_breaks), duration_breaks * scale_factor)
#   x_labels <- c(rev(count_breaks), duration_breaks)
# 
#   p <- ggplot() +
#     geom_col(data = count_df, aes(x = -n, y = hour_f),
#              fill = bar_color_left, alpha = 0.65, width = 0.8) +
#     geom_col(data = duration_df, aes(x = total_duration * scale_factor, y = hour_f),
#              fill = bar_color_right, alpha = 0.65, width = 0.8) +
#     geom_vline(xintercept = 0, color = "grey30", linewidth = 0.6) +
#     scale_x_continuous(breaks = x_breaks, labels = x_labels) +
#     scale_y_discrete(limits = as.character(0:23), drop = FALSE) +
#     facet_wrap(~ speciesEN, ncol = 2) +
#     labs(title = title, x = NULL, y = y_lab) +
#     theme_minimal() +
#     theme(
#       axis.text.x = element_text(color = "grey20", size = 8),
#       axis.text.y = element_text(size = 7),
#       strip.text  = element_text(face = "bold", size = 9)
#     ) +
#     annotate("text", x = -max(count_breaks) * 0.5, y = -1,
#              label = label_left, color = bar_color_left,
#              fontface = "bold", hjust = 0.5, size = 3) +
#     annotate("text", x = max(duration_breaks) * scale_factor * 0.5, y = -1,
#              label = label_right, color = bar_color_right,
#              fontface = "bold", hjust = 0.5, size = 3) +
#     coord_cartesian(ylim = c(1, 24), clip = "off") +
#     theme(plot.margin = margin(t = 25, r = 10, b = 25, l = 10))
# 
#   add_sun_reference_h(p, data[[time_col]])
# }


# ============================================================
# 48-hour version of the "Passages vs Duration" pyramid,
# with an average-tide-height overlay
# ============================================================
plot_hour_pyramid_detailed_48h <- function(data, time_col, duration_col, title,
                                            tide_data   = NULL,
                                            y_lab       = "Hour of Day (24h cycle repeated)",
                                            label_left  = "Number of Passages (total, Counts)",
                                            label_right = "Passage Duration (per Band.ID, Hours)",
                                            tide_color  = "darkblue") {

  bar_color <- "steelblue"
  box_color <- "black"

  df <- data %>%
    mutate(hour = lubridate::hour(.data[[time_col]]))

  df_48 <- bind_rows(
    df %>% mutate(hour48 = hour),
    df %>% mutate(hour48 = hour + 24)
  ) %>%
    mutate(hour_f = factor(hour48, levels = 0:47))

  count_df <- df_48 %>%
    count(hour48, name = "n") %>%
    right_join(tibble(hour48 = 0:47), by = "hour48") %>%
    mutate(n = tidyr::replace_na(n, 0), hour_f = factor(hour48, levels = 0:47)) %>%
    arrange(hour48)

  max_left  <- max(count_df$n, na.rm = TRUE)
  max_right <- suppressWarnings(max(df_48[[duration_col]], na.rm = TRUE))
  if (!is.finite(max_left)  || max_left  == 0) max_left  <- 1
  if (!is.finite(max_right) || max_right == 0) max_right <- 1
  scale_factor <- max_left / max_right

  left_breaks  <- scales::extended_breaks()(c(0, max_left));  left_breaks  <- left_breaks[left_breaks >= 0]
  right_breaks <- scales::extended_breaks()(c(0, max_right)); right_breaks <- right_breaks[right_breaks >= 0]

  x_breaks <- c(-rev(left_breaks), right_breaks * scale_factor)
  x_labels <- c(rev(left_breaks), right_breaks)

  y_break_vals   <- seq(0, 47, by = 3)
  y_break_labels <- as.character(y_break_vals %% 24)

  # ------------------------------------------------------------
  # Pre-compute tide scaling constants BEFORE the main scale, so the
  # secondary (top) x-axis can invert the same transform used for the line.
  # ------------------------------------------------------------
  tide_48 <- NULL
  if (!is.null(tide_data)) {
    tide_hour <- tide_data %>%
      mutate(hour = lubridate::hour(tideDateTimeAus)) %>%
      group_by(hour) %>%
      summarise(avg_height = mean(tideHeight, na.rm = TRUE), .groups = "drop") %>%
      right_join(tibble(hour = 0:23), by = "hour") %>%
      arrange(hour)

    tide_48 <- bind_rows(
      tide_hour %>% mutate(hour48 = hour),
      tide_hour %>% mutate(hour48 = hour + 24)
    ) %>%
      arrange(hour48)

    loess_fit <- loess(avg_height ~ hour48, data = tide_48, span = 0.3)
    tide_48$avg_height_smooth <- predict(loess_fit)

    tide_range     <- range(tide_48$avg_height_smooth, na.rm = TRUE)
    tide_mid       <- mean(tide_range)
    plot_half_span <- max(max_left, max_right * scale_factor) * 0.9

    tide_48 <- tide_48 %>%
      mutate(
        tide_scaled = ((avg_height_smooth - tide_mid) / (diff(tide_range) / 2)) * plot_half_span,
        y_pos       = hour48 + 1
      )
  }

  # Secondary x-axis: inverse of the tide_scaled transform above, so the top
  # axis reads directly in metres regardless of the bottom axis's units.
  sec_x_axis <- if (!is.null(tide_data)) {
    sec_axis(
      transform = ~ (. / plot_half_span) * (diff(tide_range) / 2) + tide_mid,
      name   = "Average Tide Height (m)",
      breaks = scales::pretty_breaks(n = 5)(tide_range)
    )
  } else {
    waiver()
  }

  p <- ggplot() +
    geom_col(data = count_df, aes(x = -n, y = hour_f), fill = bar_color, alpha = 0.65, width = 0.8) +
    geom_boxplot(data = df_48, aes(x = .data[[duration_col]] * scale_factor, y = hour_f, group = hour_f),
                 fill = box_color, color = box_color, alpha = 0.2, outlier.shape = NA, width = 0.6) +
    geom_jitter(data = df_48, aes(x = .data[[duration_col]] * scale_factor, y = hour_f),
                height = 0.15, width = 0, size = 0.6, alpha = 0.15, color = box_color) +
    geom_vline(xintercept = 0, color = "grey30", linewidth = 0.6) +
    scale_x_continuous(breaks = x_breaks, labels = x_labels, sec.axis = sec_x_axis) +
    scale_y_discrete(limits = as.character(0:47), drop = FALSE,
                      breaks = as.character(y_break_vals), labels = y_break_labels) +
    labs(title = title, x = NULL, y = y_lab) +
    theme_minimal() +
    theme(
      axis.text.x.top  = element_text(color = tide_color),
      axis.title.x.top = element_text(color = tide_color, face = "bold"),
      axis.text.x      = element_text(color = "grey20")
    ) +
    annotate("text", x = -max(left_breaks) * 0.5, y = -1.5,
             label = label_left, color = bar_color, fontface = "bold", hjust = 0.5, size = 3.5) +
    annotate("text", x = max(right_breaks) * scale_factor * 0.5, y = -1.5,
             label = label_right, color = box_color, fontface = "bold", hjust = 0.5, size = 3.5) +
    coord_cartesian(ylim = c(1, 48), clip = "off") +
    theme(plot.margin = margin(t = 25, r = 10, b = 30, l = 10))

  p <- add_sun_reference_h_48(p, data[[time_col]])

  if (!is.null(tide_data)) {
    p <- p +
      geom_path(data = tide_48, aes(x = tide_scaled, y = y_pos),
                color = tide_color, linewidth = 1, inherit.aes = FALSE)
  }

  p
}

# ============================================================
# Build one set of plots PER STATION
# ============================================================
plots_hourly <- list(
  count                              = list(),
  raw_count                          = list(),
  duration                           = list(),
  raw_duration_old                   = list(),
  box_duration                       = list(),
  # duration_pyramid_simple_species    = list(),
  duration_pyramid_detailed_species  = list()
  # duration_pyramid_simple            = list(),
  # duration_pyramid_detailed          = list()
)

for (s in stations) {


  plots_hourly$count[[s]] <- plot_hour_count(
    visit_by_station[[s]], "visitStart",
    title = paste0("Count of passages by Hour of Day (1h gap rule applied) \u2014 ", s))

  plots_hourly$raw_count[[s]] <- plot_hour_count(
    raw_by_station[[s]], "timeAus",
    title = paste0("Count of passages by Hour of Day (Raw Detections) \u2014 ", s))

  plots_hourly$duration[[s]] <- plot_hour_duration(
    visit_by_station[[s]], "visitStart", "duration_h",
    title = paste0("Total Visit Duration by Hour of Day \u2014 ", s))

  plots_hourly$raw_duration_old[[s]] <- plot_hour_duration_burst(
    raw_by_station[[s]], "timeAus",
    title = paste0("Raw Detection Duration (Old Burst-Interval Method) by Hour of Day \u2014 ", s))

  plots_hourly$box_duration[[s]] <- plot_hour_box_duration(
    visit_by_station[[s]], "visitStart", "duration_h",
    title = paste0("Visit Duration by Hour of Day \u2014 ", s, " (points = individual Band.ID visits)"))

  # plots_hourly$duration_pyramid_simple[[s]] <- plot_hour_pyramid_simple(
  #   visit_by_station[[s]], "visitStart", "duration_h",
  #   title = paste0("Total Passages (left) vs. Total Duration (right) \u2014 ", s))
  # 
  # plots_hourly$duration_pyramid_detailed[[s]] <- plot_hour_pyramid_detailed(
  #   visit_by_station[[s]], "visitStart", "duration_h",
  #   title = paste0("Passages (left) vs. Duration (right), by Band.ID \u2014 ", s))
  # 
  # plots_hourly$duration_pyramid_simple_species[[s]] <- plot_hour_pyramid_simple_species(
  #   visit_by_station[[s]], "visitStart", "duration_h",
  #   title = paste0("Total Passages (left) vs. Total Duration (right), by Species \u2014 ", s))
  # 
  plots_hourly$duration_pyramid_detailed_species[[s]] <- plot_hour_pyramid_detailed_species(
    visit_by_station[[s]], "visitStart", "duration_h",
    title = paste0("Passages (left) vs. Duration (right), by Band.ID & Species \u2014 ", s))
}
```

### Exploratory analysis

:::::: {.callout collapse="true" title="Click to unfold and see more details!" appearance="simple"}
#### Total count

::: panel-tabset
```{r hourly-count, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis'}
for (s in stations) {
  cat("##### ", s, "\n\n")
  print(plots_hourly$count[[s]])
  cat("\n\n")
  print(plots_hourly$raw_count[[s]])
  cat("\n\n")
}
```
:::

#### Total duration

::: panel-tabset
```{r hourly-duration, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis'}
for (s in stations) {
  cat("##### ", s, "\n\n")
  print(plots_hourly$duration[[s]])
  cat("\n\n")
  print(plots_hourly$raw_duration_old[[s]])
  cat("\n\n")
}
```
:::

#### Duration distribution

::: panel-tabset
```{r hourly-distribution, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis'}
for (s in stations) {
  cat("##### ", s, "\n\n")

  print(plots_hourly$box_duration[[s]])
  cat("\n\n")
}
```
:::
::::::

### Exploratory synthesis

**CIRCADIAN OR TIDAL CYCLE?**

The **raw data table** (one row = one detection; several consecutive detections = one run) is first processed using the **"one-hour gap" rule**, whereby any gap of *less than one hour between two runs are systematically merged into a single and continuous run*.

We refer to this merged *run* as a **Passage**. At this stage, all data is treated the same way, without distinguishing between Fly-Bys and Visits.

A **Passage** reflects *how many times a bird has been detected by the station* (count), versus *how long it was continuously detected during each passage* (duration). In other words **How often** *versus* **How long each detection lasted.**

Now, see below how the data behaves when we overlay **circadian and tidal cycles**. The detection suggests the birds are more **active** at dawn and dusk than during the day or the night (number of detections). In another hand, it seems birds **settle** somewhere during the day or night, while they just fly-by at dawn and dusk (duration of the detection).

However, it looks like most of the sites surveyed by our Motus array is used during the diurnal high tides, leaving nocturnal and low tide less documented or used.

```{r conclusion, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=8}

# Table
library(DT)
build_hour_summary_dt <- function(data, label) {
  hour_summary <- data %>%
    mutate(hour = lubridate::hour(visitStart)) %>%
    group_by(hour) %>%
    summarise(
      n_visits = n(),
      total_h  = sum(duration_h, na.rm = TRUE),
      median_h = median(duration_h, na.rm = TRUE),
      mean_h   = mean(duration_h, na.rm = TRUE),
      max_h    = max(duration_h, na.rm = TRUE),
      min_h    = min(duration_h, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(across(c(total_h, median_h, mean_h, max_h, min_h), ~ round(.x, 2)))

  datatable(
    hour_summary,
    rownames = FALSE,
    colnames = c("Hour of Day", "Number of Visits", "Total Duration (h)",
                 "Median Duration (h)", "Mean Duration (h)", "Max Duration (h)", "Min Duration (h)"),
    options = list(pageLength = 24, order = list(list(0, "asc")), dom = "ftip"),
    caption = paste0("Visit duration summary by hour of day \u2014 ", label)
  ) %>%
    formatStyle("total_h",  background = styleColorBar(hour_summary$total_h, "lightblue")) %>%
    formatStyle("median_h", background = styleColorBar(hour_summary$median_h, "lightblue"))
}

summary_tables <- setNames(
  lapply(stations, function(s) build_hour_summary_dt(visit_by_station[[s]], s)),
  stations
)


# Plot
plot_hour_pyramid_detailed_48h_ccl <- function(data, time_col, duration_col, title,
                                           tide_data     = NULL,
                                           show_shading  = TRUE,
                                           show_tide     = TRUE,
                                           indiv_dot     = TRUE,
                                           y_lab         = "Hour of Day (0\u201347, cycle repeated)",
                                           label_left    = "Number of Passages (Total)",
                                           label_right   = "Passage Duration (per Band.ID, hours)",
                                           tide_color    = "darkblue") {
  
  bar_color <- "steelblue"
  box_color <- "black"
  
  df <- data %>%
    mutate(hour = lubridate::hour(.data[[time_col]]))
  
  df_48 <- bind_rows(
    df %>% mutate(hour48 = hour),
    df %>% mutate(hour48 = hour + 24)
  ) %>%
    mutate(hour_f = factor(hour48, levels = 0:47))
  
  count_df <- df_48 %>%
    count(hour48, name = "n") %>%
    right_join(tibble(hour48 = 0:47), by = "hour48") %>%
    mutate(n = tidyr::replace_na(n, 0), hour_f = factor(hour48, levels = 0:47)) %>%
    arrange(hour48)
  
  max_left  <- max(count_df$n, na.rm = TRUE)
  max_right <- suppressWarnings(max(df_48[[duration_col]], na.rm = TRUE))
  if (!is.finite(max_left)  || max_left  == 0) max_left  <- 1
  if (!is.finite(max_right) || max_right == 0) max_right <- 1
  scale_factor <- max_left / max_right
  
  left_breaks  <- scales::extended_breaks()(c(0, max_left));  left_breaks  <- left_breaks[left_breaks >= 0]
  right_breaks <- scales::extended_breaks()(c(0, max_right)); right_breaks <- right_breaks[right_breaks >= 0]
  
  x_breaks <- c(-rev(left_breaks), right_breaks * scale_factor)
  x_labels <- c(rev(left_breaks), right_breaks)
  
  y_break_vals   <- seq(0, 47, by = 3)
  y_break_labels <- as.character(y_break_vals %% 24)
  
  # ------------------------------------------------------------
  # Tide pre-computation — only needed if the tide curve will be drawn
  # ------------------------------------------------------------
  tide_48 <- NULL
  tide_range <- tide_mid <- plot_half_span <- NULL
  
  if (show_tide && !is.null(tide_data)) {
    tide_hour <- tide_data %>%
      mutate(hour = lubridate::hour(tideDateTimeAus)) %>%
      group_by(hour) %>%
      summarise(avg_height = mean(tideHeight, na.rm = TRUE), .groups = "drop") %>%
      right_join(tibble(hour = 0:23), by = "hour") %>%
      arrange(hour)
    
    tide_48 <- bind_rows(
      tide_hour %>% mutate(hour48 = hour),
      tide_hour %>% mutate(hour48 = hour + 24)
    ) %>%
      arrange(hour48)
    
    loess_fit <- loess(avg_height ~ hour48, data = tide_48, span = 0.3)
    tide_48$avg_height_smooth <- predict(loess_fit)
    
    tide_range     <- range(tide_48$avg_height_smooth, na.rm = TRUE)
    tide_mid       <- mean(tide_range)
    plot_half_span <- max(max_left, max_right * scale_factor) * 0.9
    
    tide_48 <- tide_48 %>%
      mutate(
        tide_scaled = ((avg_height_smooth - tide_mid) / (diff(tide_range) / 2)) * plot_half_span,
        y_pos       = hour48 + 1
      )
  }
  
  sec_x_axis <- if (show_tide && !is.null(tide_data)) {
    sec_axis(
      transform = ~ (. / plot_half_span) * (diff(tide_range) / 2) + tide_mid,
      name   = "Average Tide Height (m)",
      breaks = scales::pretty_breaks(n = 5)(tide_range)
    )
  } else {
    waiver()
  }
  
  p <- ggplot() +
    geom_col(data = count_df, aes(x = -n, y = hour_f), fill = bar_color, alpha = 0.65, width = 0.8) +
    geom_boxplot(data = df_48, aes(x = .data[[duration_col]] * scale_factor, y = hour_f, group = hour_f),
                 fill = box_color, color = box_color, alpha = 0.2, outlier.shape = NA, width = 0.6)
  
  # Individual dots on the duration boxplots — only added if requested
  if (indiv_dot) {
    p <- p +
      geom_jitter(data = df_48, aes(x = .data[[duration_col]] * scale_factor, y = hour_f),
                  height = 0.15, width = 0, size = 0.6, alpha = 0.15, color = box_color)
  }
  
  p <- p +
    geom_vline(xintercept = 0, color = "grey30", linewidth = 0.6) +
    scale_x_continuous(breaks = x_breaks, labels = x_labels, sec.axis = sec_x_axis) +
    scale_y_discrete(limits = as.character(0:47), drop = FALSE,
                     breaks = as.character(y_break_vals), labels = y_break_labels) +
    labs(title = title, x = NULL, y = y_lab) +
    theme_minimal() +
    theme(
      axis.text.x.top  = element_text(color = tide_color),
      axis.title.x.top = element_text(color = tide_color, face = "bold"),
      axis.text.x      = element_text(color = "grey20")
    ) +
    annotate("text", x = -max(left_breaks) * 0.5, y = -1.5,
             label = label_left, color = bar_color, fontface = "bold", hjust = 0.5, size = 3.5) +
    annotate("text", x = max(right_breaks) * scale_factor * 0.5, y = -1.5,
             label = label_right, color = box_color, fontface = "bold", hjust = 0.5, size = 3.5) +
    coord_cartesian(ylim = c(1, 48), clip = "off") +
    theme(plot.margin = margin(t = 25, r = 10, b = 30, l = 10))
  
  # Dusk/dawn/night shading — only added if requested
  if (show_shading) {
    p <- add_sun_reference_h_48(p, data[[time_col]])
  }
  
  # Tide curve — only added if requested
  if (show_tide && !is.null(tide_48)) {
    p <- p +
      geom_path(data = tide_48, aes(x = tide_scaled, y = y_pos),
                color = tide_color, linewidth = 1, inherit.aes = FALSE)
  }
  
  p
}


  
  # Figure 1 vs Figure 2
  print(
    plot_hour_pyramid_detailed_48h_ccl(
      visit_by_station[["All Stations"]], "visitStart", "duration_h",
      title        = paste0("Passages vs. Duration \u2014 All stations", " (48h, tidal cycle perspective)"),
      tide_data    = tide_data,
      show_shading = FALSE,
      show_tide    = TRUE,
      indiv_dot    = FALSE    )  )

  # Figure 2 vs Figure 1
  print(
    plot_hour_pyramid_detailed_48h_ccl(
      visit_by_station[["All Stations"]], "visitStart", "duration_h",
      title        = paste0("Passages vs. Duration \u2014 All stations"," (48h, circadian cycle perspective)"),
      tide_data    = NULL,
      show_shading = TRUE,
      show_tide    = FALSE,
      indiv_dot    = FALSE    )  )



```

::: {.callout-tip collapse="true" title="Summary statistics"}
```{r hourly-circa-tide-table, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=8}
print(htmltools::tagList(summary_tables[["All Stations"]]))
```
:::

:::: {.callout collapse="true" title="Click to See variation per station!" appearance="simple"}
::: panel-tabset
```{r hourly-circa-tide, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=8}
for (s in stations) {
  cat("#### ", s, "\n\n")

  print(
    plot_hour_pyramid_detailed_48h(
      visit_by_station[[s]], "visitStart", "duration_h",
      title     = paste0("Passages (left) vs. Duration (right), by Band.ID \u2014 ", s, " (48h view)"),
      tide_data = tide_data
    )
  )
  cat("\n\n")
  
  cat('::: {.callout-tip collapse="true" title="Summary statistics \u2014 ', s, '"}\n\n', sep = "")
  print(htmltools::tagList(summary_tables[[s]]))
  cat("\n\n:::\n\n")
}
```
:::
::::

:::: {.callout collapse="true" title="Click to See variation per station & species!" appearance="simple"}
```{r build-summary-tables-SP, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='hide'}
library(DT)
build_hour_summary_dt <- function(data, label) {

  hour_summary <- data %>%
    mutate(hour = lubridate::hour(visitStart)) %>%
    group_by(speciesEN, hour) %>%
    summarise(
      n_visits = n(),
      total_h  = sum(duration_h, na.rm = TRUE),
      median_h = median(duration_h, na.rm = TRUE),
      mean_h   = mean(duration_h, na.rm = TRUE),
      max_h    = max(duration_h, na.rm = TRUE),
      min_h    = min(duration_h, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(speciesEN, hour) %>%
    mutate(
      across(
        c(total_h, median_h, mean_h, max_h, min_h),
        ~ round(.x, 2)
      )
    )

  DT::datatable(
    hour_summary,
    rownames = FALSE,
    colnames = c(
      "Species",
      "Hour of Day",
      "Number of Visits",
      "Total Duration (h)",
      "Median Duration (h)",
      "Mean Duration (h)",
      "Max Duration (h)",
      "Min Duration (h)"
    ),
    filter = "top",
    options = list(
      pageLength = 24,
      order = list(
        list(0, "asc"),
        list(1, "asc")
      ),
      dom = "ftip",
      scrollX = TRUE
    ),
    caption = paste0(
      "Visit duration summary by species and hour of day — ",
      label
    )
  ) %>%
    DT::formatStyle(
      "total_h",
      background = DT::styleColorBar(
        hour_summary$total_h,
        "lightblue"
      )
    ) %>%
    DT::formatStyle(
      "median_h",
      background = DT::styleColorBar(
        hour_summary$median_h,
        "lightblue"
      )
    )
}


summary_tables_sp <- setNames(
  lapply(stations, function(s) build_hour_summary_dt(visit_by_station[[s]], s)),
  stations
)
```

::: panel-tabset
```{r species-distribution, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=9.5}
for (s in stations) {
  cat("##### ", s, "\n\n")
  
  # print(plots_hourly$duration_pyramid_simple_species[[s]])
  # cat("\n\n")

  print(plots_hourly$duration_pyramid_detailed_species[[s]])
  cat("\n\n")

  cat('::: {.callout-tip collapse="true" title="Summary statistics \u2014 ', s, '"}\n\n', sep = "")
  print(htmltools::tagList(summary_tables_sp[[s]]))
  cat("\n\n:::\n\n")
}
```
:::
::::

### Method review

Overall, the results obtained in the above are descriptive, yet highly validated by the *Australasian Shorebirds Conference (2026)* community, including academic researchers. Therefore, we account for some feedback and checks, see below.

#### Count vs Duration, no overlap

::: {.callout collapse="true" title="Unfold and check the method in details!" appearance="simple"}
Following several feedback *(Australasian Shorebirds Conference - 2026)*, we ensured the counts were not double-noted when the duration is over one hour. For example, if one visit starts at 04:00 am and last 2 hours 30 min, this visit should be counted only once at 04:00 am with a duration of 2,5 hours; and not as 3 visits with one of a duration of 1 hour at 04:00 am, another of 1 hour at 05:00 am and a last one of 0.5 hours (30 min) at 06:00 am.

To do so, we will manually work through one bird with a lot of detections, so a lot of visits, spanning for some several hours and others less than an hour.

This bird in our case is a Curlew sandpiper with the Band.ID = *4281407.* Then, **we run the exact same code we used throughout the analysis above.**

**First, we determine the duration of each runs, merging two runs when they are spaced less than one hour counting the number of visit and for each their duration.**

```{r method review, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=9.5}

# filter data_all down to one Band.ID
data_dummy <- data_all %>%
  filter(Band.ID == 4281407)

# build run-level summary (exact same code as ch1_12.qmd)
runs_dummy <- data_dummy %>%
  group_by(speciesEN, Band.ID, recvDeployName, runID) %>%
  summarise(
    runStart = min(timeAus, na.rm = TRUE),
    runEnd   = max(timeAus, na.rm = TRUE),
    n_detect = n(),
    .groups = "drop")

# build visit_duration
visit_duration_dummy <- runs_dummy %>%
  arrange(Band.ID, recvDeployName, runStart) %>%
  group_by(Band.ID, recvDeployName) %>%
  mutate(
    gapHours  = round(as.numeric(difftime(runStart, dplyr::lag(runEnd), units = "hours")), 2),
    new_visit = is.na(gapHours) | gapHours >= 1,   # <-- THRESHOLD: new visit when gap >= 1h
    visitID   = cumsum(new_visit)) %>%
  group_by(Band.ID, speciesEN, recvDeployName, visitID) %>%
  summarise(
    visitStart = min(runStart),
    visitEnd   = max(runEnd),
    n_detect   = sum(n_detect),
    n_runs     = n(),
    .groups = "drop") %>%
  mutate(duration_h = round(as.numeric(difftime(visitEnd, visitStart, units = "hours")), 2))
```

**Finally, we select one particular visit to see exactly what happens in our code, and what output it provides for review.**

This visit has the id (`r visitID`) = *51* and occurring at the station (`r recvDeployName`) of *Curlew point.*

```{r method review 2, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=9.5}

# filter with unique Band.ID behaving with many runs, long and short ones, spanning multiple hours, etc.
d <- visit_duration_dummy %>%
  filter(visitID == 51, recvDeployName == "Curlew Point") %>%
  mutate(hour = lubridate::hour(visitStart))

gt(d)
```

**We compute a count column and a duration column, for each hour of the day.**

This way, we can easily check the duration is associated to a single visit, regardless its values, and to a single hour bin which must be allocated to the start time of the relative `r visitID`.

```{r method review 3, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=9.5}
# check manually how duration vs count behave against each other
plot_df_check <- tibble(hour = 0:23) %>%
  left_join(count(d, hour, name = "count"), by = "hour") %>%
  left_join(d %>% select(hour, duration_h), by = "hour") %>%
  mutate(count = tidyr::replace_na(count, 0))

gt(plot_df_check)
```

**Let's now see how it displays the result, visually.**

```{r method review 4, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=9.5}

# plot settings
max_left  <- max(plot_df_check$count, 1)
max_right <- max(d$duration_h, 1)
scale_factor <- max_left / max_right

left_breaks  <- scales::extended_breaks()(c(0, max_left))
right_breaks <- scales::extended_breaks()(c(0, max_right))
x_breaks <- c(-rev(left_breaks), right_breaks * scale_factor)
x_labels <- c(rev(left_breaks), right_breaks)

# plot
ggplot() +
  geom_col(data = plot_df_check, aes(x = -count, y = factor(hour, levels = 0:23)),
           fill = "steelblue", width = 0.8) +
  geom_point(data = d, aes(x = duration_h * scale_factor, y = factor(hour, levels = 0:23)),
             size = 3, color = "black") +
  geom_vline(xintercept = 0, color = "grey30") +
  scale_x_continuous(breaks = x_breaks, labels = x_labels) +
  labs(x = NULL, y = "Hour of Day", title = "Count (left) vs Duration (right)") +
  theme_minimal()

```
:::

#### Auto-correlation, keep in mind

::: {.callout collapse="true" title="Unfold and see how count is correlated to duration!" appearance="simple"}
As another feedback, a correlation is highly suspected between number of visits and the visit duration. For example, many visit must be shorts when few visit must be long.

In order to check this, we will plot below, collapsed into each hour bin (n = 24), the total number of visit (x axis) against the duration of these (y axis). The duration is once displayed as the mean and once as the median. Besides, a correlation test is applied to both the mean and the median scenarios.

```{r auto correlation, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=9.5}
hour_summary <- visit_duration %>%
  mutate(hour = lubridate::hour(visitStart)) %>%
  group_by(hour) %>%
  summarise(n = n(), 
            mean_duration_h = mean(duration_h, na.rm = TRUE), 
            median_duration_h = median(duration_h, na.rm = TRUE), 
            .groups = "drop")

cor_test_mean <- cor.test(hour_summary$n, hour_summary$mean_duration_h)

cor_test_median <- cor.test(hour_summary$n, hour_summary$median_duration_h)

cor_test_mean
cor_test_median

```

The mean-duration correlation is weakly positive (r = 0.15, p = 0.495) while the median-duration correlation is weakly negative (r = −0.23, p = 0.270). Both not significant, and notably pointing in opposite direction, probably due to extreme values. See plots below

```{r auto correlation test, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=9.5}
ggplot(hour_summary, aes(x = n, y = mean_duration_h)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Number of visits (per hour of day)",
       y = "Mean visit duration (hours)",
       title = paste0("r = ", round(cor_test_mean$estimate, 2), ", p = ", signif(cor_test_mean$p.value, 3))) +
  theme_minimal()

ggplot(hour_summary, aes(x = n, y = median_duration_h)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Number of visits (per hour of day)",
       y = "Median visit duration (hours)",
       title = paste0("r = ", round(cor_test_median$estimate, 2), ", p = ", signif(cor_test_median$p.value, 3))) +
  theme_minimal()
```

**We conclude there's no evidence in this data-set that busier hours of the day (with a lot of visits) are associated with systematically shorter (or longer) visits.**
:::

## Inferential analysis

Now we are convinced the circadian cycle plays a genuine role into the level of activity for shorebirds, and so their movement, **we will model the shorebird activity (i.e., number of visit, the duration of the visits) to assess how dependent it is from (1) the level of natural light and (2) the tide height.**

### Import data

::: blockquote-blue
We need to call data-set which informs us about natural ambient light, hourly, spanning across the period our birds are tracked in Newcastle, NSW.

**See here how we extracted measured, modeled, collated and formatted data for natural light level: [extract natural light](https://uoneduau-my.sharepoint.com/:b:/g/personal/c3541851_uon_edu_au/IQB9W1LwxqRMQKZ0pqyAs6qQAfIfrbkUypwxUU18rzPnR5o?e=SsylK3)**
:::

```{r bring data, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=9.5}
light_h <- read.csv(here::here("qmd", "chapter_1", "data", "solar_radiance", "newcastle_natural_light_hourly.csv"))
```

Let's now add tide height, and light level variables to our `r visit_duration`. 

All based on the time at the **start of the visit.**

```{r bring data 2, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, results='asis', fig.height=9.5}

# Avoids daylight-saving problems
visit_time_col <- "visitStart"   #visit_duration
tide_time_col  <- "tideDateTimeAus"     #tide_data
vt <- visit_duration[[visit_time_col]]
vt_utc <- as.POSIXct(format(vt, tz = "UTC"), tz = "UTC")

# Light values for the hour that contains the visit
light_cols <- c("ambiant_lux_log", "day_lux", "cloud_factor", "night_lux", "is_day")

light_start_utc <- as.POSIXct(light_h$hour_end_utc, 
                              format = "%Y-%m-%d %H:%M", 
                              tz = "UTC") - 3600

visit_hour_utc  <- as.POSIXct(trunc(vt_utc, "hours"))

idx <- match(as.numeric(visit_hour_utc), 
             as.numeric(light_start_utc))

visit_duration[light_cols] <- light_h[idx, light_cols]

# Tide: height interpolated to the exact visit time
tide_data <- tide_data[order(tide_data[[tide_time_col]]), ]

visit_duration$tideHeight <- approx(x = as.numeric(tide_data[[tide_time_col]]),
                                    y = tide_data$tideHeight,
                                    xout = as.numeric(vt_utc), ties = mean)$y

```

### Model

Now we can start modelling.


1. Number of visits: light explains it, tide doesn't.

More light means fewer visits. Each +1 SD of light (about 3.6 log units, roughly 4,000 times brighter) reduces visits starting per hour by 25% (×0.75, p < 0.001). Birds start more visits in the dark.
Tide has no effect: ×0.99, p = 0.71.
AIC agrees. The tide-only model is no better than the null model (25,096 vs 25,096). The light-only model is best (24,870), and adding tide doesn't improve it (24,872).

2. Visit duration: both matter, and equally.

Brighter light at visit start makes visits 42% longer per +1 SD (×1.42, about 3.3 log units).
Higher tide at visit start makes visits 41% longer per +1 SD (×1.41, about 0.30 m).
The model with both is clearly best: its AIC is about 100 lower than either single-predictor model, and each predictor is highly significant even with the other included.

Putting it together: a plausible reading, not yet tested, is that nights bring many short visits, while days, especially at high tide, bring fewer but much longer ones. That fits birds shuffling between feeding spots at night and roosting through daytime high tides.

Caveats to state:

Light and tide are taken at the start of each visit, which is crude for visits lasting many hours.
The date random effect is essentially zero in the count model. That's harmless: it means days don't differ once bird identity is accounted for.




Figure 1 shows both standardised effects side by side for each response. It's the quickest way to see which predictor matters more.
Figure 2 shows predicted visits per hour against light (x-axis labelled in lux) and against tide, where the flat line shows no effect.
Figure 3 shows predicted visit duration against light and against tide.

The shaded bands are 95% confidence intervals, and the small ticks along the x-axis show where the data actually lie.





OTHER TBC (AI HELP):
This split works: 0.18 h (11 minutes) sits right in the dip between the two humps. The results now tell a clear and simple story.

(a) Is a visit a long stay (>11 min) or a short pass? Mostly light, and tide a little.

Per +1 SD of light (about 3.3 log units), the odds of a long stay double: ×1.97, z = 17.2.
Per +1 SD of tide (about 0.30 m), the odds rise about 29%: ×1.29, z = 6.8.
Both effects are significant, but light is about 2.7 times stronger on the standardised scale.

(b) Once a bird stays, how long does it stay? Only tide matters.

Light has no effect at all: ×1.00, p = 0.997.
Each +1 SD of tide makes stays about 19% longer: ×1.19, p < 0.001.

Putting it together:

Light decides the type of visit. In the dark, birds make many short passes, which is also why the hourly visit count was higher at night. In daylight, visits are more often real stays.
Tide decides how long those stays last. Higher tide means longer stays, which fits birds roosting on the site over high tide.

Residuals (the plot you sent, presumably d_len): acceptable. The QQ line is almost straight, and the quantile curves stay close to 0.25 / 0.5 / 0.75. The tests are flagged mainly because n = 2,008 makes DHARMa very sensitive. Some deviation is also expected because the model only sees durations above 11 minutes, so the lower tail is cut off. State that in your methods and it's fine to report. Check that the d_long residual plot looks similar before using that model too.

For your methods, the approach can be summarised in one sentence: visit durations were bimodal, so we modelled (a) the probability that a visit was a long stay (>0.18 h, the trough between the two modes; binomial GLMM) and (b) the log-duration of long stays (Gaussian GLMM), with bird and date as random effects.

The earlier Figure 3 used the Gamma duration model, so it's now out of date. If you'd like, I can adapt the plotting script for these two models: probability of a long stay against light and tide, and predicted stay length against tide.




