---
title: "Entropy, Eveness & Composition (bis)"
format: html
toc-location: left
---
  
::: {.callout style="border-left: 6px solid #D99B8F; --bs-callout-border: #D99B8F; padding: 1.5rem; border-radius: 8px; margin-bottom: 1rem;"}


TO BE COMPLETED


:::
  
```{r set-up 1, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, include = FALSE, results = 'hide'}
# Source Station usage - hours (OBP method) script into this environment
source(knitr::purl(here::here("qmd", "chapter_1", "ch1_12.qmd"),
                   output = tempfile(fileext = ".R"),
                   quiet = FALSE))
```

```{r local packages, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
library(mixtools)
library(lubridate)
library(gt)
library(lubridate)
library(ggpattern)
```

## Set-up data

As **we will be re-using some of the data already loaded from previous workflow (see [Station usage - hours (OBP method)](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_12.html)**), we will fetch here some of the objects we will re-use including: *visit_hours_tide, tide_data, data_all, recv, shorebird_class, and available_bird_recv_time*.

::: blockquote-blue
*visit_hours_tide* is then used onward so, the same rule applies in the data-set: **\< 1 hour gap between two runs are considered as a unique use of the area covered by a Motus station - the bird remained in the area within this hour gap**.
:::
  
```{r station visits, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}


```

Now, we pre-fill some **key-values** we need to set-up the workflow.

```{r setup, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide'}


```





