---
title: "Site use - hours (OBP method)"
format: html
---

```{r style legend, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, include = FALSE, results = 'hide'}
# Source the script into that environment
source(knitr::purl(here::here("qmd", "chapter_1", "ch1_3.qmd"), 
                   output = tempfile(fileext = ".R"),    
                   quiet = TRUE))    
```

::: {.callout style="border-left: 6px solid #D99B8F; --bs-callout-border: #D99B8F; padding: 1.5rem; border-radius: 8px; margin-bottom: 1rem;"}
TO BE COMPLETED

Most of the analysis below is widely inspired from *Orange-bellied Parrot Migration Tracking Project, 2025 Report, Department of Natural Resources and Environment Tasmania*
:::

::: blockquote-blue
**Load your data** in your R environment *- see [Load & Format \> Reproducibility](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_1.html#reproducibility)*
:::

```{r packages,  message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
library(motus)
library(dplyr)
library(here)
library(ggridges)
library(tidyr)
library(readr)
library(dplyr)
library(ggplot2)
library(vegan)  
library(philentropy) 
library(gridExtra)
library(gt)
library(ggpubr)
library(car)
library(FSA)
library(purrr)
library(patchwork)
library(mixtools)
library(lubridate)
library(data.table)
```

```{r my data, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results = 'hide', include = FALSE}
# Birds
data_all <- readRDS(
  tail(sort(list.files(
    here::here("qmd", "chapter_1", "data", "motus"),
    pattern = "-data\\.rds$", full.names = TRUE
  )), 1)) 

# Receivers info
recv <- readRDS(
  tail(sort(list.files(
    here::here("qmd", "chapter_1", "data", "motus"),
    pattern = "-recv-info\\.rds$", full.names = TRUE
  )), 1)) 

# Receivers activity
sql.motus <- DBI::dbConnect(RSQLite::SQLite(), here::here("qmd", "chapter_1", "data", "project-294.motus"))
recv.act <- tbl(sql.motus, "activity")  %>% 
  collect() %>% 
  as.data.frame() %>%
  rename(deviceID = "motusDeviceID") %>%
  filter(deviceID %in% unique(recv$deviceID)) %>% # keep our deployed antennas only 
  
  # Set the time properly - IMPORTANT
  mutate(date = as_datetime(as.POSIXct(hourBin* 3600, origin = "1970-01-06", tz = "UTC")),
         dateAus = as_datetime(as.POSIXct(hourBin* 3600, origin = "1970-01-06", tz = "UTC"), 
                             tz = "Australia/Sydney")) 

# Tide
tide_data <- readRDS(here("qmd", "chapter_1", "data", "tides", "tideData.rds"))

```

::: blockquote-yellow
**Note**: As Fullerton Entrance is particular (single directional antenna), we will treat this case particularly and separately (see [Fullerton Entrance (Fly-by vs Site Use)](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_13.html)).
:::

```{r my data filter, message = FALSE, warning = FALSE, eval = FALSE, echo = TRUE, results = 'hide', include = FALSE}

data_all <- data_all %>%
  filter(recvDeployName != "Fullerton Entrance")

recv <- recv %>%
  filter(recvDeployName != "Fullerton Entrance")

```


## Site use adjusted for detection gap threshold

### Detection threshold

To define independent observations, we needed a time gap that separates a bird leaving a site from just briefly dropping out of antenna range (e.g., by turning away or moving behind vegetation). Short gaps likely mean the bird is still present; long gaps likely mean it has left. We analysed the distribution of time gaps between successive detection runs of the same bird at the same station, and across all individuals, and used this to choose a cut-off separating "same visit" from "new visit". A three-component mixture model (MIXTOOLS 2.0.0.1) was fitted to the gap durations, distinguishing short gaps (same visit) from medium/long gaps (new visit). Gaps were truncated to 5 min – 30 hours to remove noise from very short or very long outliers, and each gap was classified by whichever distribution best matched it.

**1. Build run-level summary** (one row per run) from hits. A "run" is already identified by runID in data_all [(see Motus documentation - Chapter 3)](https://motuswts.github.io/motus/articles/03-accessing-data.html). We need the start/end time of each run, per individual and per station.

```{r run , message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = TRUE}
runs <- data_all %>%
  group_by(speciesEN, Band.ID, recvDeployName, runID) %>%
  summarise(
    runStart = min(timeAus, na.rm = TRUE),
    runEnd   = max(timeAus, na.rm = TRUE),
    n_detect = n(), 
    .groups = "drop")
```

**2. Compute gaps** between successive runs of the same individual at the same antenna/site.

```{r gap inclusion, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = TRUE}
gaps <- runs %>%
  arrange(Band.ID, recvDeployName, runStart) %>%
  group_by(Band.ID, recvDeployName) %>%
  mutate(
    prevRunEnd = lag(runEnd),
    gapHours   = round(as.numeric(difftime(runStart, prevRunEnd, units = "hours")), 2)) %>%
  ungroup() %>%
  filter(!is.na(gapHours)) %>%
  filter(gapHours > 0) # we are not dealing with overlaps yet (simultaneous detection of the same tag from two or more antenna)
```

**3. Truncate gaps** (hours) for mixture model fitting. ::: blockquote-yellow **NOTE:** Below are values you should choose based on your own data and on the ecological plausibility of your system. :::

```{r gap selection, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results = 'hide', include = TRUE}

gap_trunc <- gaps %>%
  filter(gapHours > 0.3,     # ~5 minutes                   
         gapHours < 96)      # 96 hours
   # pull(gapHours)

ggplot(gap_trunc, aes(x = gapHours, y = after_stat(density))) +
  geom_histogram(binwidth = 2, # 2h
                 fill = "grey30", 
                 color = "white") +
  scale_x_continuous(breaks = seq(0, 96, 5)) + # 5h
  labs(x = "Gap Duration (hours)", y = "density") +
  theme_minimal(base_size = 13) +
   annotate("rect", 
           xmin = 0, xmax = 30, 
           ymin = 0, ymax = Inf,   # spans full height of the plot
           fill = NA, color = "red", linewidth = 1) +
  theme(panel.grid.minor = element_blank()) 

ggplot(gap_trunc %>% filter(gapHours < 30), aes(x = gapHours, y = after_stat(density))) +
  geom_histogram(binwidth = 1, # 1h
                 fill = "grey30", 
                 color = "white") +
  scale_x_continuous(breaks = seq(0, 30, 1)) + # 1h
  labs(x = "Gap Duration (hours)", y = "density") +
  theme_minimal(base_size = 13) +
   annotate("rect", 
           xmin = 0, xmax = 15, 
           ymin = 0, ymax = Inf,   # spans full height of the plot
           fill = NA, color = "red", linewidth = 1) +
  theme(panel.grid.minor = element_blank())

ggplot(gap_trunc %>% filter(gapHours < 15), aes(x = gapHours, y = after_stat(density))) +
  geom_histogram(binwidth = 0.12, # ~5min
                 fill = "grey30", 
                 color = "white") +
  scale_x_continuous(breaks = seq(0, 15, 1)) + # 1h
  labs(x = "Gap Duration (hours)", y = "density") +
  theme_minimal(base_size = 13) +
   annotate("rect", 
           xmin = 0, xmax = 2, 
           ymin = 0, ymax = Inf,   # spans full height of the plot
           fill = NA, color = "red", linewidth = 1) +
  theme(panel.grid.minor = element_blank())    # /!\ BASED ON VISUAL GRAPH, I SELECT SUBJECTIVELY 1h AS TRESHOLD - conservative as 2h could be /!\

```

**4. Fit a 3-component normal mixture model**. Now we have visually explored how data is distributed to select a gap threshold value, we might use statistical model to assess this value more systematical and based on the data only, besides visual selection accounting for ecologically plausibility.

```{r gap selection 2, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE,, include = TRUE}

gap_trunc <- gaps %>%
  filter(gapHours > 0.3,     # ~5 minutes                   
         gapHours < 30) %>%  # 30 hours
  pull(gapHours)

set.seed(123)  # for reproducibility of EM starting values

mix_model <- normalmixEM(
  gap_trunc,
  k = 3,
  maxit = 5000,
  epsilon = 1e-08)

summary(mix_model)

```

**6. Find the threshold value,** the cut-point: the gap value where classification switches from "Non-independent" to an "Independent" class (i.e. where posterior prob of the non-independent component drops below 0.5 as gap duration increases). We will first classify each gap to the component with highest posterior probability, and order components by their mean, so component 1 = shortest-gap (non-independent), 2 = medium, 3 = long (independent).

```{r gap selection 3, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results = 'hide', include = TRUE}
posteriors <- mix_model$posterior
classification <- apply(posteriors, 1, which.max)
comp_order <- order(mix_model$mu)
comp_labels <- c("Non-independent", "Independent (medium)", "Independent (long)")
names(comp_labels) <- as.character(comp_order)

class_df <- data.frame(
  gapHours = gap_trunc,
  component = comp_labels[as.character(classification)])

nonindep_comp <- comp_order[1]

cutpoint_df <- data.frame(
  gapHours = gap_trunc,
  p_nonindep = posteriors[, nonindep_comp]) %>%
  arrange(gapHours)

# first point where posterior probability of non-independence < 0.5
cutpoint <- cutpoint_df$gapHours[which(cutpoint_df$p_nonindep < 0.5)[1]]
cat("Threshold value or Cut-point (hours):", round(cutpoint, 2), "\n")

# Plot
x_grid <- seq(0, 30, length.out = 500)

dens_df <- do.call(rbind, lapply(seq_along(mix_model$mu), function(i) {
  data.frame(
    x = x_grid,
    y = mix_model$lambda[i] * dnorm(x_grid, mix_model$mu[i], mix_model$sigma[i]),
    component = comp_labels[as.character(i)])
}))

comp_colors <- c(
  "Non-independent"      = "darkred",
  "Independent (medium)" = "lightblue",
  "Independent (long)"   = "darkgreen")

ggplot() +
  geom_histogram(
    data = data.frame(gapHours = gap_trunc),
    aes(x = gapHours, y = after_stat(density)),
    binwidth = 1, fill = "grey30", color = "white", boundary = 0) +
  geom_line(
    data = dens_df,
    aes(x = x, y = y, color = component),
    linewidth = 1) +
  geom_vline(xintercept = cutpoint, color = "red", linewidth = 0.6) +
  scale_color_manual(values = comp_colors) +
  scale_x_continuous(breaks = seq(0, 30, 2)) +
  labs(
    x = "Gap Duration (hours)",
    y = "density",
    color = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank())

```

### Define used time (hours)

Now, we decide to merge into one and unique continuous detection, all the tags (birds) detected per a same antenna, including gaps under a duration of 1 (one) hour. Although the model is highlighting a value of `r  round(cutpoint, 2)`, we decided to remain conservative.

**Once detection data summarised as described above, we can see, overall, how long birds stay around one Motus station** - see plot below.

```{r site use, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results = 'hide', include = TRUE}

visit_duration <- runs %>%
  arrange(Band.ID, recvDeployName, runStart) %>%
  group_by(Band.ID, recvDeployName) %>%
  mutate(
    gapHours = round(as.numeric(runStart - lag(runEnd))  / 3600, 2),
    new_visit = is.na(gapHours) | gapHours >= 1,                 # <-- HERE IS THE THRESHOLD VALUE TO SET A "NEW VISIT" (TRUE when gapHours >= 1) 
    visitID = cumsum(new_visit)) %>%
  group_by(Band.ID, speciesEN, recvDeployName, visitID) %>%
  summarise(
    visitStart = min(runStart),
    visitEnd   = max(runEnd),
    n_detect   = sum(n_detect),                                  # total hits/detection across merged runs
    n_runs     = n(),                                            # how many runs got merged into this visit
    .groups = "drop") %>%
  
  mutate(duration_h = round(as.numeric(visitEnd - visitStart) / 3600, 2))    # visit duration in hours


ggplot(visit_duration %>%
         filter(duration_h < 20),
       aes(x = duration_h, y = after_stat(count))) +
  geom_histogram(binwidth = 0.5, 
                 fill = "grey30", 
                 color = "white") +
  scale_x_continuous(breaks = seq(0, 20, 1)) + # 1h
  labs(x = "Visit Duration (hours)", y = "Count (n)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())    

```

\*Note Newcastle NSW tides are turning \~6h cycles between high/low, and our birds seem to mostly visits any site less than 3 hours. An early descriptive sign of site use very tide dependent, as they would stay about 3 hours in the same area, around the highest/lowest point of the tide line, before to leave site, as the tide might be rising or receiving.\*

```{r site use filter, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, include = TRUE}
ggplot(visit_duration %>%
         filter(duration_h > 20),
       aes(x = duration_h, y = after_stat(count))) +
  geom_histogram(binwidth = 25, 
                 boundary = 20,       
                 fill = "grey30", 
                 color = "white") +
  scale_x_continuous(breaks = sort(c(20, seq(500, 3000, 500)))) +
  coord_cartesian(xlim = c(0, 3000)) +
  labs(x = "Visit Duration (hours)", y = "Count (n)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

To avoid inflating site use duration estimates with the very few **individuals showing more than 25 hours of continuous presence in the same area, we excluded these cases**. Such extended duration likely reflect a dropped tag rather than genuine site fidelity. Verifying each case individually would be time-consuming, so we treat this as a working assumption for now; should this pattern become more frequent in future data, it would warrant closer investigation.

```{r site use filter 2, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = TRUE}
visit_duration <- visit_duration %>%
         filter(duration_h < 25)
```

**It remains to attribute at each visit period a tide category.** We expand each visit into hourly bins spanning `visitStart` to `visitEnd` (`visit_duration` itself is untouched). Then, we compute exact overlap (in hours) between each hourly bin and the actual visit window.

```{r site use tide, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = TRUE}
visit_hours <- visit_duration %>%
  rowwise() %>%
  mutate(hour_dt = list(seq(from = floor_date(visitStart, "hour"),
                             to   = floor_date(visitEnd, "hour"),
                             by   = "hour"))) %>%
  unnest(cols = c(hour_dt)) %>%
  ungroup()

visit_hours <- visit_hours %>%
  mutate(
    bin_start     = hour_dt,
    bin_end       = hour_dt + hours(1),
    overlap_start = pmax(visitStart, bin_start),
    overlap_end   = pmin(visitEnd, bin_end),
    overlap_h     = as.numeric(difftime(overlap_end, overlap_start, units = "hours"))
  ) %>%
  filter(overlap_h > 0)   # drops bins with no actual overlap

# Attach nearest tide category to each hourly bin
tide_dt <- as.data.table(tide_data)[, .(tideDateTimeAus, tideCategory)]
setkey(tide_dt, tideDateTimeAus)

vh_dt <- as.data.table(visit_hours)
setkey(vh_dt, hour_dt)

visit_hours_tide <- tide_dt[vh_dt, on = .(tideDateTimeAus = hour_dt), roll = "nearest"]
setnames(visit_hours_tide, "tideDateTimeAus", "hour_dt")

```

**Let's finalise the table giving data for detection relative to the *trio* bird-station-tide.**

```{r site use final table, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = TRUE}

used_bird_recv_time <- visit_hours_tide %>%
  mutate(tideDiel = if_else(grepl("Diurnal", tideCategory), "Diurnal", "Nocturnal"),
         tideHighLow = if_else(grepl("High", tideCategory), "High", "Low")) %>%
  group_by(Band.ID, speciesEN, recvDeployName, tideCategory, tideDiel, tideHighLow) %>%
  summarise(duration_h = sum(overlap_h), .groups = "drop")

```

### Define available time (hours)

Now used time addressed, we need to define for each individual (birds) and across each Motus stations, what was the amount of time available for the four tide categories.

Available time for each Motus stations is accessed with `activity` table: each row is a dated record of noise or a tag detected *- see [Survey effort \> Activity table](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_5.html#activity-table)*.

We need to extract the temporal coverage of each station, starting from deployment date to the last recorded activity (noise or detection).

```{r Available vs Used RECV AVAILABLE, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE, results = 'asis'}

# Sort the terminated serno (if terminated, ie. one box removed from one antenna site, a date comes along)
# but still needed for accessing survey effort as the station is currently running with another serno 
recv.act.term <- recv.act %>%
  left_join(recv %>% 
              filter(!is.na(timeEndAus)) %>%
              select(deviceID, serno, recvDeployName),
            "deviceID") %>%
  filter(!is.na(recvDeployName)) %>%
  mutate(SernoStation = paste0(recvDeployName, "_", serno))

# Sort the currently running serno
recv.act.runn <- recv.act %>%
  left_join(recv %>% 
              filter(is.na(timeEndAus)) %>%
              select(deviceID, serno, recvDeployName),
            "deviceID") %>%
  filter(!is.na(recvDeployName)) %>%
  mutate(SernoStation = paste0(recvDeployName, "_", serno))

# Merging in one data-set to use Station's name further + pick-up the rounded hours
recv.act <- bind_rows(recv.act.runn, recv.act.term) %>%
  mutate(hour_dt = round_date(dateAus, "hour"))

# Providing helpful variables
recv <- recv %>%
  mutate(SernoStation = paste0(recvDeployName, "_", serno),
         lisStart = timeStartAus,
         lisEnd = if_else(
           is.na(timeEndAus), # means the station is still running since the last data downloading
           with_tz(Sys.time(), "Australia/Sydney"),
           with_tz(as_datetime(timeEndAus, tz = "UTC"), "Australia/Sydney")) )

# Generating hourly sequences per SernoStation from start to end dates of the deviceID at particular sites
recv_hours <- recv %>%
  select(recvDeployName, deviceID, SernoStation, lisStart, lisEnd) %>%
  group_by(SernoStation) %>%
  rowwise() %>%
  mutate(hour_dt = list(seq(from = round_date(lisStart, unit = "hour"),
                            to = round_date(lisEnd, unit = "hour"),
                            by = "hour")) ) %>%
  unnest(cols = c(hour_dt)) %>%
  ungroup()

# Simplify station variables (recvDeployName)
recv <- recv %>% 
  select(!recvDeployName) 
recv$recvDeployName <- sub("_SG-.*", "", recv$SernoStation)

recv_hours <- recv_hours %>% 
  select(!recvDeployName) 
recv_hours$recvDeployName <- sub("_SG-.*", "", recv_hours$SernoStation)


```

<br>

We now how the complete temporal coverage, expanded in hours and for each Motus station so it matches with our used time dataset.

However, due to some temporary failures or maintenance *(see [Survey effort \> Operational periods](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_5.html#operational-periods))*, the actual period a station was operational may be different than the complete temporal coverage. So, we must determine when a station was not operational and substrate these off-hours to the complete temporal coverage.

<br>

```{r Available vs Used RECV OFF PERIODS, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE, results = 'asis'}


# Distinguish station from mixed recv + Giving operational variable (= TRUE when existing values from act table)
recv_hours <- recv_hours %>%
  select(-any_of("operational")) %>%   # drop if it already exists, no-op if it doesn't
  left_join(recv.act %>%
              distinct(recvDeployName, hour_dt) %>%
              mutate(operational = TRUE),
            by = c("recvDeployName", "hour_dt")) %>%
  mutate(operational = if_else(is.na(operational), FALSE, TRUE))

# /!\ Due to unknown error? Have to set this manually
recv_hours <- recv_hours %>%
  mutate(operational = case_when(
    recvDeployName == "Fullerton Entrance" & 
      hour_dt > as.POSIXct("2023-04-02") & 
      hour_dt < as.POSIXct("2023-04-05") ~ FALSE,
    TRUE ~ operational))

# Only the hours the station was actually ON — computed once, no regeneration of gaps
valid_station_hours <- recv_hours %>%
  filter(operational == TRUE) %>%
  distinct(recvDeployName, hour_dt)

# Attach nearest tide category directly (this also folds in the Point 7 fix —
tide_dt <- as.data.table(tide_data)[, .(tideDateTimeAus, tideCategory)]
setkey(tide_dt, tideDateTimeAus)

vsh_dt <- as.data.table(valid_station_hours)
setkey(vsh_dt, hour_dt)

valid_station_hours_tide <- tide_dt[vsh_dt, on = .(tideDateTimeAus = hour_dt), roll = "nearest"]
setnames(valid_station_hours_tide, "tideDateTimeAus", "hour_dt")

```

<br>

Temporal coverage for each station is now reduced to its operational periods. **We now assign each bird its own available time period**, spanning from its tagging date to its last Motus detection, and expanded to hourly resolution too.

<br>

```{r Available vs Used BIRD AVAILABLE TIME, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE, results = 'asis'}


# Get the monitored period of each bird
period_sp <- data_all  %>%
  group_by(Band.ID) %>%
  reframe(DateAUS.Trap = first(DateAUS.Trap), 
          last_dateAus = max(timeAus),
          speciesEN = first(speciesEN)) %>%
  unique()

# Expand one row per hours to each individual across its whole period (this is the available time)
bird_hours <- period_sp %>%
  group_by(Band.ID) %>%
  rowwise() %>%
  mutate(hour_dt = list(seq(from = as.POSIXct(ymd(DateAUS.Trap), tz = "UTC"),
                            to = as.POSIXct(last_dateAus, tz = "UTC"),
                            by = "hour"))) %>%
  unnest(cols = c(hour_dt)) %>%
  ungroup() 

# Duplicate in as many list as many stations
recv_names <- unique(valid_station_hours_tide$recvDeployName)
bird_data_list <- setNames(vector("list", length(recv_names)), recv_names)

for(name in recv_names) {
  valid_hours_recv <- valid_station_hours_tide %>%
    filter(recvDeployName == name) %>%
    pull(hour_dt)
  
  valid_hours_bird_recv <- bird_hours %>%
    filter(hour_dt %in% valid_hours_recv)
  
  valid_hours_bird_recv$name <- name
  bird_data_list[[name]] <- valid_hours_bird_recv
}

# # FINALISE AVAILABLE TIME 
available_bird_recv_time <- bind_rows(bird_data_list) %>%
  mutate(hour_dt = force_tz(hour_dt, "Australia/Sydney")) %>%
  filter(if_all(everything(), ~ !is.na(.))) %>%
  rename(recvDeployName = name) %>%
  left_join(valid_station_hours_tide, by = c("recvDeployName", "hour_dt")) %>%
  filter(!is.na(tideCategory)) %>%    # bird-hours the station wasn't actually ON for get dropped here
  group_by(Band.ID, speciesEN, recvDeployName, tideCategory) %>%
  summarise(duration_h = n(), .groups = "drop") %>%
  mutate(tideDiel = if_else(grepl("Diurnal", tideCategory), "Diurnal", "Nocturnal"),
         tideHighLow = if_else(grepl("High", tideCategory), "High", "Low")) %>%
  select(Band.ID, speciesEN, recvDeployName, tideCategory, tideDiel, tideHighLow, duration_h)
```

::: blockquote-blue
**The *available* time is now defined** for each bird across each Motus station, taking into account the periods where stations where ON only. In the data, the *available* periods are expanded in hours and the tide condition is associated. While most of the couple station listening-bird tagged are potentially detectable between 0 and 1000 hours; note two small peaks around 2700 and 4000 hours of detectability potential as it might be relative to the oldest couple ON.
:::

```{r Available plot, message = FALSE, warning = FALSE, echo = FALSE, include = TRUE,, eval = TRUE }
ggplot() +
  geom_histogram(
    data = available_bird_recv_time,
    aes(x = duration_h, y = after_stat(count)),
    binwidth = 100, 
    fill = "grey30", 
    color = "white", 
    boundary = 0) +
  scale_x_continuous(breaks = seq(0, 4000, 500)) +
  labs(
    x = "Listening Duration for a typical Motus station (hours per birds) [all data]",
    y = "Count (n)",
    color = NULL) +
     annotate("rect", 
           xmin = 0, xmax = 1000, 
           ymin = 0, ymax = Inf,   # spans full height of the plot
           fill = NA, color = "red", linewidth = 1) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank())

ggplot() +
  geom_histogram(
    data = available_bird_recv_time %>%
      filter(duration_h < 1000),
    aes(x = duration_h, y = after_stat(count)),
    binwidth = 100, 
    fill = "grey30", 
    color = "white", 
    boundary = 0) +
  scale_x_continuous(breaks = seq(0, 1000, 100)) +
  labs(
    x = "Listening Duration for a typical Motus station (hours per birds) [0-1000 hours]",
    y = "Count (n)",
    color = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank())

ggplot() +
  geom_histogram(
    data = available_bird_recv_time %>%
      filter(duration_h > 1000),
    aes(x = duration_h, y = after_stat(count)),
    binwidth = 100, 
    fill = "grey30", 
    color = "white", 
    boundary = 0) +
  scale_x_continuous(breaks = seq(0, 4000, 1000)) +
  labs(
    x = "Listening Duration for a typical Motus station (hours per birds) [1000-Inf hours]",
    y = "Count (n)",
    color = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank())

```

### Define rate of Use (%)

Now that the *used* and *available* time have been determined for each individual and for each Motus station, we can relate these two elements together with the **Usage rate** (or rate of use).

::: {style="overflow-x: auto; width: 100%;"}
$$
\text{Usage rate} = \frac{\text{Time used}}{\text{Time available}} \times 100 \quad \textit{across stations and species} 
$$
:::

```{r final table, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE}
use_rate_plot <- used_bird_recv_time %>%
  select(-tideDiel, -tideHighLow) %>%   # drop here, keep the copy from available_bird_recv_time
  rename(used_h = duration_h) %>%
  
  left_join(available_bird_recv_time %>%
      rename(available_h = duration_h),   
    by = c("Band.ID", "speciesEN", "recvDeployName", "tideCategory")) %>%
  
  mutate(rate_use = (used_h / available_h) * 100) %>%
  select(Band.ID, speciesEN, recvDeployName, tideCategory, 
         tideDiel, tideHighLow, used_h, available_h, rate_use)

```

```{r potential error, message = FALSE, warning = FALSE, echo = FALSE, eval = TRUE}

above <- use_rate_plot %>%
  filter(rate_use > 100) %>%
  nrow()

```

::: blockquote-yellow
**Note**: A very few cases (n = `r above`) where the *time used* was over the *time* *available* has been excluded from the data. Reasons might be the tag fell off within range of the antenna. Due to the negligible number of such cases, we have not further investigated.
:::

```{r potential error 2, message = FALSE, warning = FALSE, echo = FALSE, eval = TRUE}

use_rate_plot %>%
  filter(rate_use > 100) %>%
  mutate(used_h = round(used_h, 2),
         available_h = round(available_h, 2),
         rate_use = round(rate_use, 2)) %>%
  DT::datatable()

```

+----------+
| \        |
| Nocturna |
+----------+

```{r potential error filtered, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE}

use_rate_plot <- use_rate_plot %>%
  filter(rate_use < 100)

```

::: blockquote-yellow
**Note:** We filtered out Masked Lapwing from the analysis, since they have too few data.
:::

```{r Time Used Rate, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE, results = 'asis'}

figure_plot <- use_rate_plot %>%
  mutate(speciesType = factor(shorebird_class[speciesEN], 
                               levels = c("migratory", "resident"))) %>%
  
  filter(!speciesEN %in% c("Masked Lapwing")) 

```


### Results

**The plot below shows you how much each species have been detected over the array, by night and by day.** First displayed overall. Second displayed per condition and showing individual variation.

```{r 4 Used rate tot, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
# Per-bird mean rate_use by species and tide/diel category
per_bird <- use_rate_plot %>%
  group_by(Band.ID, speciesEN, tideDiel) %>%
  summarise(rate_use = mean(rate_use, na.rm = TRUE), .groups = "drop")

# Per-bird "Undetected" category
other_per_bird <- per_bird %>%
  group_by(Band.ID, speciesEN) %>%
  summarise(rate_use = 100 - sum(rate_use, na.rm = TRUE), .groups = "drop") %>%
  mutate(tideDiel = "Undetected")

# Combine per-bird data (this was missing)
per_bird_full <- bind_rows(per_bird, other_per_bird) %>%
  mutate(tideDiel = factor(tideDiel, levels = c("Undetected", "Diurnal", "Nocturnal")))

# --- Stacked bar (species-level means, from per-bird data) ---
plot_data_full <- per_bird_full %>%
  group_by(speciesEN, tideDiel) %>%
  summarise(mean_rate_use = mean(rate_use, na.rm = TRUE), .groups = "drop")

ggplot(plot_data_full, aes(x = speciesEN, y = mean_rate_use, fill = tideDiel)) +
  geom_bar(stat = "identity", position = "stack", width = 0.7) +
  scale_fill_manual(values = c("Nocturnal" = "#2C3E50",
                                "Diurnal" = "#F4A300",
                                "Undetected" = "#B0B0B0"),
                     name = "Tide period",
                     breaks = c("Nocturnal", "Diurnal", "Undetected")) +
  labs(x = "Species", y = "Use rate (%)",
       title = "Use rate by species (day vs night vs undetected)") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")

# --- Dodged bars + individual points (per-bird variation) ---
summary_data <- per_bird_full %>%
  group_by(speciesEN, tideDiel) %>%
  summarise(mean_rate = mean(rate_use, na.rm = TRUE),
            se = sd(rate_use, na.rm = TRUE) / sqrt(sum(!is.na(rate_use))),
            .groups = "drop")

ggplot() +
  geom_bar(data = summary_data,
           aes(x = speciesEN, y = mean_rate, fill = tideDiel),
           stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(data = summary_data,
                aes(x = speciesEN, ymin = pmax(mean_rate - se, 0), ymax = mean_rate + se,
                    group = tideDiel),
                position = position_dodge(width = 0.8), width = 0.2) +
  geom_point(data = per_bird_full,
             aes(x = speciesEN, y = rate_use, group = tideDiel),
             position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.15),
             size = 1.2, alpha = 0.4, color = "black") +
  scale_fill_manual(values = c("Nocturnal" = "#2C3E50",
                                "Diurnal" = "#F4A300",
                                "Undetected" = "#B0B0B0"),
                     name = "Tide period") +
  labs(x = "Species", y = "Use rate (%)",
       title = "Use rate by species (day vs night vs undetected)",
       subtitle = "Bars = mean ± SE, points = individual birds") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")
```

```{r tool, message = FALSE, warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
# Track size sample
counts <- data_all %>%
      group_by(speciesEN) %>%
      summarise(n = n_distinct(Band.ID)) %>%
      mutate(label = paste0(speciesEN, " (n = ", n, ")"))
label_vec <- setNames(counts$label, counts$speciesEN)

# Tide and species groupings
tide_levels <- c("Low", "High")
species_types <- unique(shorebird_class)  

# Simplified function - no min_n_nonzero check
make_plot <- function(tide_level, species_type) {
  data_sub <- figure_plot %>%
    filter(tideHighLow == tide_level) %>%
    mutate(species_class = shorebird_class[speciesEN]) %>%
    filter(species_class == species_type)
  
  p <- ggplot(data_sub,
              aes(x    = factor(recvDeployName, levels = sort(unique(recvDeployName))),
                  y    = rate_use,
                  fill = tideDiel))
  
  # Always add boxplot for all available data
  p <- p + 
    geom_boxplot(outlier.shape = NA,
                 varwidth = FALSE,
                 position = position_dodge(width = 0.8, preserve = "single")) +
    
    # Always show individual points
    geom_point(aes(shape = tideDiel), 
               position = position_dodge(width = 0.8),
               alpha = 1, size = 1.5,
               show.legend = FALSE) +
    scale_shape_manual(values = c("Diurnal" = 21, "Nocturnal" = 16)) +
    
    facet_wrap(~ speciesEN,
               labeller = labeller(speciesEN = label_vec)) +
    labs(x     = "Receiver Deployment",
         y     = "Rate of Use (%)",
         title = paste(ifelse(species_type == "migratory",
                              "Migratory species", "Resident species"),
                       "during", tide_level, "tide")) +
    coord_cartesian(ylim = c(0, 100)) +
    theme_minimal() +
    scale_fill_manual(values = c("Diurnal" = "white", "Nocturnal" = "darkgrey")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  return(p)
}

# Generate all plots
plots_used_rate <- purrr::cross2(tide_levels, species_types) %>%
  purrr::map(~ make_plot(.x[[1]], .x[[2]]))








make_plot_v <- function(tide_level, species_type) {
  data_sub <- figure_plot %>%
    filter(tideHighLow == tide_level) %>%
    mutate(species_class = shorebird_class[speciesEN]) %>%
    filter(species_class == species_type)
  
  p <- ggplot(data_sub,
              aes(x    = factor(recvDeployName, levels = sort(unique(recvDeployName))),
                  y    = rate_use,
                  fill = tideDiel))
  
  # Violin plot instead
  p <- p + 
    geom_violin(outlier.shape = NA,
                trim = TRUE,
                varwidth = FALSE,
                scale = "width",
                width = 0.8,
                position = position_dodge(width = 1.2, preserve = "single"),
                alpha = 0.7) +
    
    # Always show individual points
    geom_point(aes(shape = tideDiel), 
               position = position_dodge(width = 0.8),
               alpha = 1, size = 1.5,
               show.legend = FALSE) +
    scale_shape_manual(values = c("Diurnal" = 21, "Nocturnal" = 16)) +
    
    facet_wrap(~ speciesEN,
               labeller = labeller(speciesEN = label_vec)) +
    labs(x     = "Receiver Deployment",
         y     = "Rate of Use (%)",
         title = paste(ifelse(species_type == "migratory",
                              "Migratory species", "Resident species"),
                       "during", tide_level, "tide")) +
    coord_cartesian(ylim = c(0, 100)) +
    theme_minimal() +
    scale_fill_manual(values = c("Diurnal" = "white", "Nocturnal" = "darkgrey")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  return(p)
}

# Generate all plots
plots_used_rate_v <- purrr::cross2(tide_levels, species_types) %>%
  purrr::map(~ make_plot_v(.x[[1]], .x[[2]]))

```

```{r stats for plots, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}

# General function to provide one table with stats values
make_summary_table <- function(tide_level, species_type) {
  data_sub <- figure_plot %>%
    filter(tideHighLow == tide_level) %>%
    mutate(species_class = shorebird_class[speciesEN]) %>%
    filter(species_class == species_type)
  
  summary_table <- data_sub %>%
    group_by(speciesEN, recvDeployName, tideDiel) %>%
    summarise(
      n_individual = n_distinct(Band.ID),
      min = round(min(rate_use, na.rm = TRUE), 1),
      q1 = round(quantile(rate_use, 0.25, na.rm = TRUE),1),
      median = round(median(rate_use, na.rm = TRUE),1),
      q3 = round(quantile(rate_use, 0.75, na.rm = TRUE),1),
      max = round(max(rate_use, na.rm = TRUE),1),
      mean = round(mean(rate_use, na.rm = TRUE),1),
      sd = round(sd(rate_use, na.rm = TRUE),1),
      .groups = 'drop'
    ) %>%
    arrange(speciesEN, recvDeployName, tideDiel) %>%
    rename(Species = "speciesEN", Station = "recvDeployName", condition = "tideDiel")
  
  return(summary_table)
}

# One table per conditions
high_migratory <- make_summary_table("High", "migratory")
high_resident <- make_summary_table("High", "resident")
low_migratory <- make_summary_table("Low", "migratory")
low_resident <- make_summary_table("Low", "resident")
```

### Resident species

::::: panel-tabset
##### High Tide

```{r 4 Used rate 1, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
plots_used_rate[[4]]
plots_used_rate_v[[4]]
```

::: {.callout-tip collapse="true" title="High Tide Resident birds"}
```{r 4 Used rate table, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
high_resident  %>%
  gt() %>%
  tab_header(title = "Statistics - High Tide Resident Species") %>%
  opt_align_table_header(align = "left") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(Species))) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "darkgrey",
      weight = px(2)),
    locations = cells_body(
      rows = which(!duplicated(high_resident$Species))[-1]))
```
:::

##### Low Tide

```{r 4 Used rate 2, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}

plots_used_rate[[3]]
plots_used_rate_v[[3]]

```

::: {.callout-tip collapse="true" title="Low Tide Resident birds"}
```{r 4 Used rate table 2, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
low_resident  %>%
  gt() %>%
  tab_header(title = "Statistics - Low Tide Resident Species") %>%
  opt_align_table_header(align = "left") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(Species))) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "darkgrey",
      weight = px(2)),
    locations = cells_body(
      rows = which(!duplicated(low_resident$Species))[-1]))
```
:::
:::::

### Migrant species

::::: panel-tabset
##### High Tide

```{r 4 Used rate 3, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
plots_used_rate[[2]]
plots_used_rate_v[[2]]

```

::: {.callout-tip collapse="true" title="High Tide Migratory birds"}
```{r 4 Used rate table 3, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
high_migratory  %>%
  gt() %>%
  tab_header(title = "Statistics - High Tide Migratory Species") %>%
  opt_align_table_header(align = "left") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(Species))) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "darkgrey",
      weight = px(2)),
    locations = cells_body(
      rows = which(!duplicated(high_migratory$Species))[-1]))
```
:::

##### Low Tide

```{r 4 Used rate 4, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}

plots_used_rate[[1]]
plots_used_rate_v[[1]]

```

::: {.callout-tip collapse="true" title="Low Tide Migratory birds"}
```{r 4 Used rate table 5, message = FALSE,  warning = FALSE, echo = FALSE, eval = TRUE, results = 'asis'}
low_migratory  %>%
  gt() %>%
  tab_header(title = "Statistics - Low Tide Migratory Species") %>%
  opt_align_table_header(align = "left") %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(Species))) %>%
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "darkgrey",
      weight = px(2)),
    locations = cells_body(
      rows = which(!duplicated(low_migratory$Species))[-1]))
```
:::
:::::
