---
title: "Legend"
---

All the way through the analysis of **Chapter 1** we will use the same English and Scientific names for species, the same abbreviations, the same names for the stations and the same color codes. Find here on this page clarifications.

::: blockquote-yellow
**Note for authors**: Make sure to keep this script up to date as new data are compiled with eventual new species and/or Motus stations.
:::

```{r station, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}

station_rename <- list("Barry_Fullerton_cove"  = "Fullerton Entrance", 
                       "North Swann Pond" = "Swan Pond" ,
                       "Ramsar Road Floodgate" = "Ramsar Road",   
                       "Milham's Pond" = "Milhams Pond") 

```

```{r color, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}

species_colors <- c(
  "Bar-tailed Godwit"      = "#FF1493",  
  "Far Eastern Curlew"     = "#FF8C00",  
  "Masked Lapwing"         = "#32CD32",  
  "Pacific Golden-Plover"  = "#FFD700",  
  "Pied Stilt"             = "#00CED1",  
  "Curlew Sandpiper"       = "#9370DB",  
  "Red-necked Avocet"      = "#FF4500", 
  "Eurasian Whimbrel"      = "#1E90FF"   
)

species_init_colors <- c(
  "BTG" = "#FF1493",  
  "FEC" = "#FF8C00",  
  "ML"  = "#32CD32",  
  "PGP" = "#FFD700",  
  "PS"  = "#00CED1",  
  "CS"  = "#9370DB",  
  "RNA" = "#FF4500",  
  "EW"  = "#1E90FF"   
  )

```

```{r initials, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
init_species <- c(
  "BTG" = "Bar-tailed Godwit",  
  "FEC" = "Far Eastern Curlew", 
  "ML"  = "Masked Lapwing", 
  "PGP" = "Pacific Golden-Plover", 
  "PS"  = "Pied Stilt", 
  "CS"  = "Curlew Sandpiper",  
  "RNA" = "Red-necked Avocet",
  "EW"  = "Eurasian Whimbrel"  
  )
```

```{r english, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
eng_names <- c(
  "Bar-tailed Godwit"     = "Limosa lapponica",
  "Far Eastern Curlew"    = "Numenius madagascariensis",
  "Masked Lapwing"        = "Vanellus miles",
  "Pacific Golden-Plover" = "Pluvialis fulva",
  "Pied Stilt"            = "Himantopus leucocephalus",
  "Curlew Sandpiper"      = "Calidris ferruginea",
  "Red-necked Avocet"     = "Recurvirostra novaehollandiae",
  "Eurasian Whimbrel"     = "Numenius phaeopus"
  )
```

```{r scientific, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
sci_names <- c(
  "Limosa lapponica"              = "Bar-tailed Godwit",
  "Numenius madagascariensis"     = "Far Eastern Curlew",
  "Vanellus miles"                = "Masked Lapwing",
  "Pluvialis fulva"               = "Pacific Golden-Plover",
  "Himantopus leucocephalus"      = "Pied Stilt",
  "Calidris ferruginea"           = "Curlew Sandpiper",
  "Recurvirostra novaehollandiae" = "Red-necked Avocet",
  "Numenius phaeopus"             = "Eurasian Whimbrel"
  )
```

```{r class, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
shorebird_class <- c(
  "Bar-tailed Godwit"     = "migratory",
  "Far Eastern Curlew"    = "migratory",
  "Masked Lapwing"        = "resident",
  "Pacific Golden-Plover" = "migratory",
  "Pied Stilt"            = "resident",
  "Curlew Sandpiper"      = "migratory",
  "Red-necked Avocet"     = "resident",
  "Eurasian Whimbrel"     = "migratory"
  )
```

## Color Table

```{r display_colors, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results='asis'}
library(knitr)
# Create a data frame with species and colors
color_df <- data.frame(
  Species = names(species_colors),
  Color_Code = unname(species_colors),
  stringsAsFactors = FALSE
)
cat('<table style="width: 100%; border-collapse: collapse;">\n')
cat('<thead><tr style="background-color: #f0f0f0;">\n')
cat('<th style="padding: 10px; text-align: left; border: 1px solid #ddd;">Species</th>\n')
cat('<th style="padding: 10px; text-align: left; border: 1px solid #ddd;">Color Code</th>\n')
cat('<th style="padding: 10px; text-align: center; border: 1px solid #ddd;">Color Swatch</th>\n')
cat('</tr></thead>\n<tbody>\n')
for(i in 1:nrow(color_df)) {
  cat('<tr>\n')
  cat(sprintf('<td style="padding: 10px; border: 1px solid #ddd; font-weight: bold;">%s</td>\n', color_df$Species[i]))
  cat(sprintf('<td style="padding: 10px; border: 1px solid #ddd; font-family: monospace;">%s</td>\n', color_df$Color_Code[i]))
  cat(sprintf('<td style="padding: 10px; border: 1px solid #ddd; text-align: center;"><div style="width: 60px; height: 30px; background-color: %s; border: 1px solid #333; margin: 0 auto; border-radius: 4px;"></div></td>\n', color_df$Color_Code[i]))
  cat('</tr>\n')
}
cat('</tbody>\n</table>\n')

```

## Species Information Table

```{r species_table, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results='asis'}
library(knitr)

# Create a comprehensive data frame
species_df <- data.frame(
  Initial = names(init_species),
  English_Name = unname(init_species),
  Scientific_Name = names(sci_names),
  Class = unname(shorebird_class),
  stringsAsFactors = FALSE
)

# Generate HTML table
cat('<table style="width: 100%; border-collapse: collapse;">\n')
cat('<thead><tr style="background-color: #f0f0f0;">\n')
cat('<th style="padding: 10px; text-align: center; border: 1px solid #ddd;">Initial</th>\n')
cat('<th style="padding: 10px; text-align: left; border: 1px solid #ddd;">English Name</th>\n')
cat('<th style="padding: 10px; text-align: left; border: 1px solid #ddd;">Scientific Name</th>\n')
cat('<th style="padding: 10px; text-align: center; border: 1px solid #ddd;">Class</th>\n')
cat('</tr></thead>\n<tbody>\n')

for(i in 1:nrow(species_df)) {
  # Alternate row colors
  bg_color <- if(i %% 2 == 0) "#f9f9f9" else "white"
  
  # Color code for class
  class_color <- if(species_df$Class[i] == "migratory") "#E7298A" else "#66A61E"
  
  cat(sprintf('<tr style="background-color: %s;">\n', bg_color))
  cat(sprintf('<td style="padding: 10px; border: 1px solid #ddd; text-align: center; font-weight: bold; font-family: monospace;">%s</td>\n', species_df$Initial[i]))
  cat(sprintf('<td style="padding: 10px; border: 1px solid #ddd;">%s</td>\n', species_df$English_Name[i]))
  cat(sprintf('<td style="padding: 10px; border: 1px solid #ddd; font-style: italic;">%s</td>\n', species_df$Scientific_Name[i]))
  cat(sprintf('<td style="padding: 10px; border: 1px solid #ddd; text-align: center;"><span style="background-color: %s; color: white; padding: 4px 12px; border-radius: 12px; font-size: 0.85rem; font-weight: 600;">%s</span></td>\n', 
              class_color, species_df$Class[i]))
  cat('</tr>\n')
}

cat('</tbody>\n</table>\n')
```
