---
title: "Signal strength"
format: html
---

```{r style legend, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, include = FALSE, results = 'hide'}
# Source the script into that environment
source(knitr::purl(here::here("qmd", "chapter_1", "ch1_3.qmd"), 
                   output = tempfile(fileext = ".R"), 
                   quiet = TRUE))
```

::: {.callout style="border-left: 6px solid #D99B8F; --bs-callout-border: #D99B8F; padding: 1.5rem; border-radius: 8px; margin-bottom: 1rem;"}
*This is preliminary analysis.*

**Signal strength** is a metric we get from a tag detection which is likely related to the distance of the tag from a Motus station at the time its detection - [*see documentation*](https://motuswts.github.io/motus/articles/signal-strength.html?q=signal%20strenght#basic-plot).

Using this metric, we could characterise behaviors such as foraging or roosting activity.

-   **Foraging:** If we assume that large signal strength variations reflect substantial movement within the Motus station’s coverage area, these variations could be associated with the research of food.
-   **Roosting:** In contrast, small variations would suggest limited movement, potentially corresponding to the seek for rest.
:::

::: blockquote-blue
**Load your data** in your R environment *- see [Load & Format \> Reproducibility](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_1.html#reproducibility)*
:::

```{r packages,  message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
library(dplyr)
library(here)
library(forcats) 
library(lubridate)
library(purrr) 
library(lubridate)
library(ggplot2)
library(stringr)
```

```{r my data, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, results = 'hide', include = FALSE}
# Birds
data_all <- readRDS(
  tail(sort(list.files(
    here::here("qmd", "chapter_1", "data", "motus"),
    pattern = "-data\\.rds$", full.names = TRUE
  )), 1)) 

# Receivers info
recv <- readRDS(
  tail(sort(list.files(
    here::here("qmd", "chapter_1", "data", "motus"),
    pattern = "-recv-info\\.rds$", full.names = TRUE
  )), 1)) 

# Selecting upon our needs
data_all <- data_all %>%
  select(Band.ID, recvDeployName, recv, speciesEN, sig, sigsd, sigPositive, tideCategory, timeAus)
```

```{r compute variables, echo = TRUE, eval = TRUE}
# Mean signal strength per individual, site and time categories (HD, HN, LD, LN)
data_all <- data_all %>%
  group_by(Band.ID, tideCategory, recvDeployName) %>%
  mutate(sig_indiv_tideCat = mean(sigPositive)) %>%
  ungroup()  %>%
  mutate(speciesINIT = str_split(speciesEN, " ") %>%            
           map_chr(~ str_c(str_sub(.x, 1, 1), collapse = "")) %>%    
           toupper(),
         tideCat = str_split(tideCategory, "_") %>%          
           map_chr(~ str_c(str_sub(.x, 1, 1), collapse = "")) %>%    
           toupper())
```

Here, we take the relative **signal strength value** `sigPositive` (a [*derivative*](https://themaskou.github.io/movement_ecology_shorebirds/qmd/chapter_1/ch1_1.html#add-key-variables) of the raw values `sig`) to calculate the mean for each individual and per tide category.

Therefore, we group the means per species and per site to observe how the signal strength varies. To that aim, we need to create a new variable: the average signal strength per individuals `sig_indiv_tideCat`.

Then, we can create a list which plots the distribution of the average *signal strength* *variation* across species.

```{r plot list, echo = TRUE, eval = TRUE, results = 'asis'}
# One plot per species for better understanding
plots_list <- data_all %>%
  
  split(.$speciesEN) %>%
  
  purrr::imap(~ {
    
    species_full <- .y  # The split name (full species name)
    
    ggplot(.x, aes(x = tideCat, y = sig_indiv_tideCat, fill = speciesEN)) +
      geom_boxplot() +
      facet_wrap(~ recvDeployName) +
      
      scale_fill_manual(values = species_colors, guide = "none") +
      theme_bw() +
      labs(title = paste0("Signal Strength variance for ", species_full, " over Motus stations"),
           x = "Tide Category",
           y = "Signal Strength variance (RSSI)",
           fill = "Species") 
    })

# Rename list elements to initials using the inverse mapping
names(plots_list) <- names(species_init_colors)[match(names(plots_list), names(species_colors))]

```

::: panel-tabset

```{r final tabs, echo = FALSE, eval = TRUE, results = 'asis', fig.width = 10, fig.height = 6}
# Create tabs dynamically
for (species_init in names(plots_list)) {
  species_full <- names(species_colors)[match(species_init, names(species_init_colors))]
  cat("###", species_full, "\n\n")
  print(plots_list[[species_init]])
  cat("\n\n")
}
```

:::

**Legend**: The average value of the *signal strength* has been calculated for each individual, so you can see its distribution (boxes on vertical axis) depending the tide condition (horizontal axis) and for each species (tabs).
