---
title: "Estuary Usage"
format: html
---

```{r style legend, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE, include = FALSE, results = 'hide'}
# Source the script into that environment
source(knitr::purl(here::here("qmd", "chapter_1", "ch1_3.qmd"), 
                   output = tempfile(fileext = ".R"), 
                   quiet = TRUE))
```

::: {.callout style="border-left: 6px solid #D99B8F; --bs-callout-border: #D99B8F; padding: 1.5rem; border-radius: 8px; margin-bottom: 1rem;"}
Birds remain within the same estuaries. Migratory ones during all summer and residents all year around. That's a statement at species group level. **But do the individuals remain within the same estuary or do they fly from one to another?**

To address this question we'll quantify how many days a bird has been detected at least once on at least one Motus station in one estuary, and we'll put this number of days against the total number of day that same bird was detectable. The result will be expressed in percentage.
:::

## Variables

::: blockquote-blue
**Load your data** in your R environment *- see [Load & Format \> Reproducibility](qmd/chapter_1/ch1_1.qmd)*
:::

```{r packages,  message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE}
library(dplyr)
library(here)
library(ggplot2)
library(lubridate)
library(tidyr)
```

```{r my data, message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE, results = 'hide', include = FALSE}
# Birds
data_all <- readRDS(
  tail(sort(list.files(
    here::here("qmd", "chapter_1", "data", "motus"),
    pattern = "-data\\.rds$", full.names = TRUE
  )), 1)) 
```

Let's count days and detecting stations for each bird.

```{r process,  message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE}

# Get deployment duration for each tags (in days)
tag_dep_duration <- data_all %>%
  group_by(Band.ID) %>%
  summarise(start_date = min(DateAUS.Trap),
            end_date = max(dateAus),
            period_tag_dep_d = 1 + time_length(interval(start = start_date, end = end_date), unit = "day"), # +1 to include the trapping day
            .groups = "drop") %>%
  select(Band.ID, period_tag_dep_d)

# Get number days each tag has been detected at least once in any station
tag_dep_nb_d <- data_all %>%
  mutate(dateAus = as.Date(dateAus)) %>%
  group_by(Band.ID) %>%
  summarise(days_detect = n_distinct(dateAus), .groups = "drop")

# Get number of days one tag has been recorded in >1 station (for each tags)
multiple_site_detect <- data_all %>%
  group_by(Band.ID, dateAus) %>%
  # Counts per day the number of station one tag has been recorded
  summarise(distinct_sites = n_distinct(recvDeployName),
            .groups = "drop_last") %>%
  # Tells if one day one tag has been recorded several sites
  mutate(multiple_detect = distinct_sites > 1) %>%
  group_by(Band.ID) %>%
  summarise(multiple_site_detect = sum(multiple_detect),
            .groups = "drop")

```

We generate `tag_dep_duration` that counts the total of days `period_tag_dep_d` each bird `Band.ID` had a tag on and activated. We make `tag_dep_nb_d` that quantifies the number of days `days_detect` each bird have been detected at least once in the Motus array.

So, we can then compare the first value with the second, and obtain our metric of interest in percentage.

Additionally, we quantified in `multiple_site_detect` the number of days each bird has been detected in multiple stations.

## Results

### INDIVIDUALS

```{r indiv table, echo = TRUE, eval = TRUE}
# Get number of day each tag has been recorded /species and /station
tag_detection_indiv <- data_all %>%
  group_by(Band.ID, recvDeployName, speciesEN) %>%
  # to count sp*ID per recv
  reframe(days_recorded = n_distinct(dateAus)) %>%
  # flip table to get one column per station
  pivot_wider(names_from = recvDeployName, 
              values_from = days_recorded,
              values_fill = 0) %>% # = 0 if NA
  left_join(multiple_site_detect, by = "Band.ID") %>%
  left_join(tag_dep_nb_d, by = "Band.ID") %>%
  left_join(tag_dep_duration, by = "Band.ID") %>%
  mutate(perc = round((days_detect / period_tag_dep_d) * 100, 1)) # nb of day one tag is recorded once / nb of day the tag is deployed

```

```{r indiv table pub,  message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
library(gt)
tag_detection_indiv  %>%
  gt() %>%
  opt_align_table_header(align = "left")  %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels() ) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(speciesEN))) 

```

<br>

The above table presents daily detection for each bird and for each Motus station in the local array:

-   `multiple_site_detect`: number of days a bird has been detected in more than one station
-   `days_detect`: number of days a bird has been detected in at least one station
-   `period_tag_dep_d`: total of days a bird was detectable with a tag on and activated
-   `perc`: **(`days_detect` / `period_tag_dep_d`) x 100**

<br>

Visualise below **for each birds** the number of days the individual has been detected at least once and in at least one Motus station in the array, against the total of days the bird had a tag on and activated (results in %).

<br>

```{r 3 plot indiv, echo = FALSE, eval = TRUE}
ggplot(tag_detection_indiv %>%
         group_by(speciesEN) %>%
         arrange(desc(perc), .by_group = TRUE) %>%
         mutate(Band.ID = factor(Band.ID, levels = unique(Band.ID))), 
       aes(x = Band.ID, y = perc, fill = speciesEN)) +
  geom_col(width = 0.8, alpha = 0.7, color = "black") +
  scale_fill_manual(values = species_colors) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(x = "Band ID",
       y = "Detection (%)",
       title = "Bird detection across MOTUS array (% of days)")
```

### SPECIES

Below are the same results as above, but **for each species**. So the data has been grouped by species.

```{r species table,  message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}

# Get deployment duration for each tags (in days)
tag_dep_duration <- data_all %>%
  group_by(speciesEN) %>%
  summarise(start_date = min(DateAUS.Trap),
            end_date = max(dateAus),
            period_tag_dep_d = 1 + time_length(interval(start = start_date, end = end_date), unit = "day"), # +1 to include the trapping day
            .groups = "drop") %>%
  select(speciesEN, period_tag_dep_d)

# Get number days each tag has been detected at least once in any station
tag_dep_nb_d <- data_all %>%
  mutate(dateAus = as.Date(dateAus)) %>%
  group_by(speciesEN) %>%
  summarise(days_detect = n_distinct(dateAus), .groups = "drop")

# Get number of days one tag has been recorded in >1 station (for each tags)
multiple_site_detect <- data_all %>%
  group_by(speciesEN, dateAus) %>%
  # Counts per day the number of station one tag has been recorded
  summarise(distinct_sites = n_distinct(recvDeployName),
            .groups = "drop_last") %>%
  # Tells if one day one tag has been recorded several sites
  mutate(multiple_detect = distinct_sites > 1) %>%
  group_by(speciesEN) %>%
  summarise(multiple_site_detect = sum(multiple_detect),
            .groups = "drop")

tag_detection <- data_all %>%
  group_by(recvDeployName, speciesEN) %>%
  # to count sp*ID per recv
  reframe(days_recorded = n_distinct(dateAus)) %>%
  # flip table to get one column per station
  pivot_wider(names_from = recvDeployName, 
              values_from = days_recorded,
              values_fill = 0) %>% # = 0 if NA
  left_join(multiple_site_detect, by = "speciesEN") %>%
  left_join(tag_dep_nb_d, by = "speciesEN") %>%
  left_join(tag_dep_duration, by = "speciesEN") %>%
  mutate(perc = round((days_detect / period_tag_dep_d) * 100, 1))

library(gt)
tag_detection  %>%
  gt() %>%
  opt_align_table_header(align = "left")  %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels() ) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = c(speciesEN))) 

```

<br>

```{r 3 species histo, echo = FALSE, eval = TRUE}
ggplot(tag_detection %>% 
         mutate(speciesEN = forcats::fct_reorder(speciesEN, perc, .desc = TRUE)),      
       aes(x = speciesEN, 
           y = perc, 
           fill = speciesEN)) +  # fill mapped to speciesEN
  geom_col(width = 0.8, alpha = 0.7, color = "black") +
  scale_fill_manual(values = species_colors) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none") +
  labs(x = "Species",
       y = "Detection (%)",
       title = "Species detectability across the MOTUS array (Histogram)")

```

```{r 3 species plot boxplot, echo = TRUE, eval = TRUE}
ggplot(tag_detection_indiv %>%
         add_count(speciesEN) %>%
         mutate(species_label = paste0(speciesEN, " (n = ", n, ")")),
       aes(x = reorder(species_label, -perc, FUN = median),
           y = perc, fill = speciesEN)) +   
  geom_boxplot(width = 0.8, alpha = 0.7, color = "black") +
  scale_fill_manual(values = species_colors) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none") +
  labs(x = "Species (n = number of individuals)",
       y = "Detection (%)",
       title = "Species detectability across the MOTUS array (Boxplot)")

```
